---
title: "TDP-43 cross-dataset integrative overview"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  polymenidou_bulk_rds_dir: "../../results/paper/tdp43/polymenidou_tdp_kd_oe_cultures_bulk/ruv_rds"
  lee_bulk_rds_dir: "../../results/paper/tdp43/lee_nuclear_tdp_retention_bulk/ruv_rds"
  polymenidou_sc_rds_dir: "../../results/paper/tdp43/polymenidou_tdp_oe_cultures_scrnaseq/ruv_rds"
  report_outdir: "../../results/paper/tdp43/integrative_rds"
  fdr_cutoff: 0.05
  logfc_cutoff: 0
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 6, fig.height = 5)
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(purrr)
  library(tibble)
  library(scales)
})

source("../workflow/scripts/biology_checks.R")

dir.create(params$report_outdir, recursive = TRUE, showWarnings = FALSE)

save_rds <- function(obj, name) {
  saveRDS(obj, file.path(params$report_outdir, paste0(name, ".rds")))
}

save_csv <- function(df, name) {
  readr::write_csv(df, file.path(params$report_outdir, paste0(name, ".csv")))
}
```

# Inputs

Three studies, each contributing one or more TDP-43 contrasts. The shared axis
across bulk and sc is repeat subfamily (`gene_id` in bulk, STARsolo feature
rows in sc), split by genic vs intergenic origin.

```{r load}
read_rds_safe <- function(path) {
  if (!file.exists(path)) {
    warning(sprintf("missing: %s", path))
    return(NULL)
  }
  readRDS(path)
}

poly_bulk <- read_rds_safe(file.path(params$polymenidou_bulk_rds_dir, "ruv_repeat_de.rds"))
lee_bulk  <- read_rds_safe(file.path(params$lee_bulk_rds_dir,        "ruv_repeat_de.rds"))
poly_sc   <- read_rds_safe(file.path(params$polymenidou_sc_rds_dir,  "ruv_repeat_de.rds"))

lee_bulk_gene_naive  <- read_rds_safe(file.path(params$lee_bulk_rds_dir,
                                                "ruv_naive_gene_de.rds"))
poly_bulk_gene_naive <- read_rds_safe(file.path(params$polymenidou_bulk_rds_dir,
                                                "ruv_naive_gene_de_per_contrast.rds"))
poly_sc_gene_naive   <- read_rds_safe(file.path(params$polymenidou_sc_rds_dir,
                                                "ruv_naive_gene_de.rds"))

biology_annotation <- read_rds_safe(file.path(params$lee_bulk_rds_dir,
                                              "biology_annotation.rds"))
if (is.null(biology_annotation)) {
  biology_annotation <- read_rds_safe(file.path(params$polymenidou_bulk_rds_dir,
                                                "biology_annotation.rds"))
}
if (is.null(biology_annotation)) {
  biology_annotation <- read_rds_safe(file.path(params$polymenidou_sc_rds_dir,
                                                "biology_annotation.rds"))
}
```

```{r inventory}
summarise_rds <- function(obj, label) {
  if (is.null(obj)) return(tibble(study = label, key = NA_character_, n_rows = NA_integer_))
  if (is.list(obj) && !is.data.frame(obj) && length(obj) > 0) {
    tibble(study = label, key = names(obj),
           n_rows = vapply(obj, function(x) nrow(as.data.frame(x)), integer(1)))
  } else if (is.data.frame(obj)) {
    tibble(study = label, key = "single", n_rows = nrow(obj))
  } else {
    tibble(study = label, key = "nested", n_rows = NA_integer_)
  }
}

bind_rows(
  summarise_rds(poly_bulk, "polymenidou_bulk_repeats"),
  summarise_rds(lee_bulk,  "lee_bulk_repeats"),
  summarise_rds(poly_sc,   "polymenidou_sc_repeats")
) %>% knitr::kable()
```

# Cross-study contrast map

```{r contrast_map}
contrast_info <- tribble(
  ~label,          ~study,             ~contrast_id,              ~direction_note,
  "lee_tdppos",    "lee_bulk",         "tdp43pos_vs_tdp43neg",    "up = nuclear TDP-43 retained (Lee sorted nuclei)",
  "poly_kd",       "polymenidou_bulk", "tdp43_kd_vs_nt_kd_ctrl",  "up = knockdown (Polymenidou bulk, nuclear loss analog)",
  "poly_oe",       "polymenidou_bulk", "tdp43_oe_vs_oe_ctrl",     "up = overexpression (Polymenidou bulk)",
  "sc_cluster12",  "polymenidou_sc",   "cluster12_vs_other",      "up = cluster 12 (Polymenidou sc OE)"
)
knitr::kable(contrast_info)
```

# Harmonize repeat DE tables per category

Gene_id granularity is the shared axis across bulk and sc. Genic vs intergenic
is split upstream at the count matrix level, so each study exposes two
repeat tables per contrast.

```{r pick_repeat_tables}
repeat_key_map <- list(
  lee_bulk = list(
    tdp43pos_vs_tdp43neg = list(genic = "genic_geneid", intergenic = "inter_geneid")
  ),
  polymenidou_bulk = list(
    tdp43_kd_vs_nt_kd_ctrl = list(
      genic = "genic_geneid:tdp43_kd_vs_nt_kd_ctrl",
      intergenic = "inter_geneid:tdp43_kd_vs_nt_kd_ctrl"
    ),
    tdp43_oe_vs_oe_ctrl = list(
      genic = "genic_geneid:tdp43_oe_vs_oe_ctrl",
      intergenic = "inter_geneid:tdp43_oe_vs_oe_ctrl"
    )
  ),
  polymenidou_sc = list(
    cluster12_vs_other = list(
      genic = "unique_genic_repeats:cluster12_vs_other",
      intergenic = "unique_intergenic_repeats:cluster12_vs_other"
    )
  )
)

repeat_rds_map <- list(
  lee_bulk = lee_bulk,
  polymenidou_bulk = poly_bulk,
  polymenidou_sc = poly_sc
)

pick_repeat_tbl <- function(study, contrast_id, category) {
  rds <- repeat_rds_map[[study]]
  if (is.null(rds)) return(NULL)
  key <- repeat_key_map[[study]][[contrast_id]][[category]]
  if (is.null(key) || !key %in% names(rds)) return(NULL)
  df <- as_tibble(rds[[key]]) %>%
    select(feature_id, logFC, PValue, FDR) %>%
    mutate(sig = FDR < params$fdr_cutoff & abs(logFC) > params$logfc_cutoff)
  df
}

de_tables <- list()
for (i in seq_len(nrow(contrast_info))) {
  lbl <- contrast_info$label[i]
  study <- contrast_info$study[i]
  cid <- contrast_info$contrast_id[i]
  for (cat in c("genic", "intergenic")) {
    tbl <- pick_repeat_tbl(study, cid, cat)
    if (!is.null(tbl)) {
      de_tables[[paste(lbl, cat, sep = "__")]] <- tbl
    }
  }
}

de_inventory <- tibble(
  table = names(de_tables),
  n = vapply(de_tables, nrow, integer(1)),
  n_sig = vapply(de_tables, function(x) sum(x$sig, na.rm = TRUE), integer(1))
)
knitr::kable(de_inventory)
save_rds(de_tables, "integrative_repeat_de_tables")
```

# Biology annotation join

Each repeat DE table is annotated with the per-subfamily biology TSV produced
by `build_repeat_biology_annotation.py`. Subfamily id in bulk (gene_id) and
STARsolo repeat feature id in sc both map to the `gene_id` column of the
biology annotation.

```{r annotate_biology}
if (!is.null(biology_annotation)) {
  bio <- as_tibble(biology_annotation) %>%
    select(gene_id, family_id, class_id, family_age,
           frac_intergenic, frac_downstream_a_run) %>%
    mutate(polya_class = as.character(polya_label(class_id)),
           age_class   = as.character(family_age_label(family_age)),
           a_run_class = as.character(a_run_label(frac_downstream_a_run)))
  de_tables_anno <- lapply(de_tables, function(df) {
    left_join(df, bio, by = c("feature_id" = "gene_id"))
  })
} else {
  bio <- NULL
  de_tables_anno <- de_tables
}
save_rds(de_tables_anno, "integrative_repeat_de_tables_annotated")
```

# Pairwise repeat logFC concordance, by category

For each genic/intergenic axis, inner-join logFC tables between pairs of
contrasts and compute Pearson / Spearman plus quadrant counts at
FDR < `r params$fdr_cutoff`.

```{r pair_repeat_helper}
pair_logfc <- function(tA, tB, nameA, nameB) {
  if (is.null(tA) || is.null(tB)) return(NULL)
  j <- inner_join(
    tA %>% transmute(feature_id, logFC_A = logFC, FDR_A = FDR, sig_A = sig,
                     polya_class = if ("polya_class" %in% names(.)) polya_class else NA,
                     a_run_class = if ("a_run_class" %in% names(.)) a_run_class else NA,
                     age_class   = if ("age_class"   %in% names(.)) age_class   else NA,
                     class_id    = if ("class_id"    %in% names(.)) class_id    else NA),
    tB %>% transmute(feature_id, logFC_B = logFC, FDR_B = FDR, sig_B = sig),
    by = "feature_id"
  )
  if (nrow(j) < 5) return(NULL)
  list(
    pair = c(nameA, nameB),
    joined = j,
    pearson = cor(j$logFC_A, j$logFC_B, method = "pearson"),
    spearman = cor(j$logFC_A, j$logFC_B, method = "spearman"),
    n = nrow(j),
    n_sig_both = sum(j$sig_A & j$sig_B, na.rm = TRUE),
    n_concordant_sig = sum(j$sig_A & j$sig_B &
                             sign(j$logFC_A) == sign(j$logFC_B), na.rm = TRUE),
    n_discordant_sig = sum(j$sig_A & j$sig_B &
                             sign(j$logFC_A) != sign(j$logFC_B), na.rm = TRUE)
  )
}

pairs_of_interest <- list(
  c("lee_tdppos",   "poly_kd"),
  c("lee_tdppos",   "poly_oe"),
  c("lee_tdppos",   "sc_cluster12"),
  c("poly_kd",      "poly_oe"),
  c("poly_oe",      "sc_cluster12"),
  c("poly_kd",      "sc_cluster12")
)

run_pairs_for_category <- function(category) {
  out <- list()
  for (p in pairs_of_interest) {
    tagA <- paste(p[1], category, sep = "__")
    tagB <- paste(p[2], category, sep = "__")
    r <- pair_logfc(de_tables_anno[[tagA]], de_tables_anno[[tagB]], p[1], p[2])
    if (!is.null(r)) {
      r$category <- category
      out[[paste(p[1], "vs", p[2], sep = "__")]] <- r
    }
  }
  out
}

pair_results_genic <- run_pairs_for_category("genic")
pair_results_inter <- run_pairs_for_category("intergenic")
```

```{r pair_repeat_summary}
summarise_pair_results <- function(lst, category) {
  bind_rows(lapply(lst, function(r) {
    tibble(category = category,
           pair_A = r$pair[1], pair_B = r$pair[2],
           n = r$n, pearson = r$pearson, spearman = r$spearman,
           n_sig_both = r$n_sig_both,
           n_concordant_sig = r$n_concordant_sig,
           n_discordant_sig = r$n_discordant_sig)
  }))
}

summary_df <- bind_rows(
  summarise_pair_results(pair_results_genic, "genic"),
  summarise_pair_results(pair_results_inter, "intergenic")
)
knitr::kable(summary_df, digits = 3)
save_csv(summary_df, "integrative_repeat_concordance_summary")
save_rds(list(genic = pair_results_genic,
              intergenic = pair_results_inter),
         "integrative_repeat_pair_results")
```

```{r pair_repeat_scatter, fig.width = 5.5, fig.height = 5}
scatter_pair <- function(r, title_prefix) {
  j <- r$joined
  subtitle <- sprintf("n=%d  pearson=%.2f  spearman=%.2f  sig-both=%d (concord=%d discord=%d)",
                      r$n, r$pearson, r$spearman,
                      r$n_sig_both, r$n_concordant_sig, r$n_discordant_sig)
  ggplot(j, aes(logFC_A, logFC_B)) +
    geom_vline(xintercept = 0, color = "grey70") +
    geom_hline(yintercept = 0, color = "grey70") +
    geom_point(aes(color = interaction(sig_A, sig_B)),
               alpha = 0.7, size = 1.2) +
    scale_color_manual(values = c("FALSE.FALSE" = "grey80",
                                  "TRUE.FALSE"  = "steelblue",
                                  "FALSE.TRUE"  = "goldenrod",
                                  "TRUE.TRUE"   = "firebrick"),
                       name = "sig in") +
    labs(title = paste(title_prefix, paste(r$pair, collapse = " vs ")),
         subtitle = subtitle,
         x = paste0("logFC ", r$pair[1]),
         y = paste0("logFC ", r$pair[2])) +
    theme_bw()
}

for (nm in names(pair_results_genic)) {
  print(scatter_pair(pair_results_genic[[nm]], "[genic]"))
}
for (nm in names(pair_results_inter)) {
  print(scatter_pair(pair_results_inter[[nm]], "[intergenic]"))
}
```

# Concordance faceted by polyA competence

Class-based polyA competence from `polya_label(class_id)`: LINE/SINE/LTR/
Retroposon subfamilies count as polyA-competent; DNA transposons and
satellites do not. Definition is shared with the per-dataset reports
(`workflow/scripts/biology_checks.R`).

```{r polya_scatter, fig.width = 8, fig.height = 4.5}
scatter_pair_faceted <- function(r, facet_col, title_prefix, palette) {
  j <- r$joined
  if (all(is.na(j[[facet_col]]))) return(invisible(NULL))
  subtitle <- sprintf("n=%d  pearson=%.2f  spearman=%.2f",
                      r$n, r$pearson, r$spearman)
  j <- j[!is.na(j[[facet_col]]), , drop = FALSE]
  ggplot(j, aes(logFC_A, logFC_B, color = .data[[facet_col]])) +
    geom_vline(xintercept = 0, color = "grey70") +
    geom_hline(yintercept = 0, color = "grey70") +
    geom_point(alpha = 0.7, size = 1.2) +
    facet_wrap(as.formula(paste("~", facet_col)), nrow = 1) +
    scale_color_manual(values = palette, na.value = "grey70") +
    labs(title = paste(title_prefix, paste(r$pair, collapse = " vs ")),
         subtitle = subtitle,
         x = paste0("logFC ", r$pair[1]),
         y = paste0("logFC ", r$pair[2])) +
    theme_bw() +
    theme(legend.position = "none")
}

polya_palette <- c("polyA-competent" = "firebrick",
                   "not polyA-competent" = "steelblue",
                   "unknown" = "grey70")

for (nm in names(pair_results_genic)) {
  print(scatter_pair_faceted(pair_results_genic[[nm]], "polya_class",
                             "[genic]", polya_palette))
}
for (nm in names(pair_results_inter)) {
  print(scatter_pair_faceted(pair_results_inter[[nm]], "polya_class",
                             "[intergenic]", polya_palette))
}
```

```{r polya_correlation_table}
stratified_corr <- function(lst, category, facet_col) {
  bind_rows(lapply(lst, function(r) {
    j <- r$joined
    if (all(is.na(j[[facet_col]]))) return(NULL)
    j %>%
      filter(!is.na(.data[[facet_col]])) %>%
      group_by(stratum = .data[[facet_col]]) %>%
      summarise(n = dplyr::n(),
                pearson = if (dplyr::n() > 2)
                  cor(logFC_A, logFC_B, method = "pearson") else NA_real_,
                spearman = if (dplyr::n() > 2)
                  cor(logFC_A, logFC_B, method = "spearman") else NA_real_,
                n_sig_both = sum(sig_A & sig_B, na.rm = TRUE),
                .groups = "drop") %>%
      mutate(stratification = facet_col,
             category = category,
             pair = paste(r$pair, collapse = " vs "))
  }))
}

polya_tbl <- bind_rows(
  stratified_corr(pair_results_genic, "genic", "polya_class"),
  stratified_corr(pair_results_inter, "intergenic", "polya_class")
)
knitr::kable(polya_tbl, digits = 3)
save_csv(polya_tbl, "integrative_polya_stratified_correlation")
```

# Concordance faceted by downstream A-run (polyA priming proxy)

Post-integration 3'-downstream A-run fraction from
`a_run_label(frac_downstream_a_run)` buckets subfamilies into low/medium/high.
Complementary to class-based polyA competence; only present for subfamilies
carried through the biology annotation.

```{r a_run_scatter, fig.width = 9, fig.height = 4.5}
a_run_palette <- c(low = "steelblue", medium = "goldenrod",
                   high = "firebrick", unknown = "grey70")

for (nm in names(pair_results_genic)) {
  print(scatter_pair_faceted(pair_results_genic[[nm]], "a_run_class",
                             "[genic]", a_run_palette))
}
for (nm in names(pair_results_inter)) {
  print(scatter_pair_faceted(pair_results_inter[[nm]], "a_run_class",
                             "[intergenic]", a_run_palette))
}
```

```{r a_run_correlation_table}
a_run_tbl <- bind_rows(
  stratified_corr(pair_results_genic, "genic", "a_run_class"),
  stratified_corr(pair_results_inter, "intergenic", "a_run_class")
)
knitr::kable(a_run_tbl, digits = 3)
save_csv(a_run_tbl, "integrative_a_run_stratified_correlation")
```

# Concordance faceted by family age

```{r age_correlation_table}
age_tbl <- bind_rows(
  stratified_corr(pair_results_genic, "genic", "age_class"),
  stratified_corr(pair_results_inter, "intergenic", "age_class")
)
knitr::kable(age_tbl, digits = 3)
save_csv(age_tbl, "integrative_age_stratified_correlation")
```

# Gene-level naive (gene-norm) concordance

Gene-level DE, gene-TMM normalization, no RUV. Lee bulk and sc save
`ruv_naive_gene_de.rds` directly (two-group contrast or the unique-mode
pseudobulk). Polymenidou bulk splits its four-group design into two
contrasts in `ruv_naive_gene_de_per_contrast.rds`.

```{r harmonise_genes}
as_gene_tbl <- function(obj) {
  if (is.null(obj)) return(NULL)
  df <- as.data.frame(obj)
  if (!"feature_id" %in% names(df)) {
    df$feature_id <- rownames(df)
  }
  as_tibble(df) %>%
    select(feature_id, any_of(c("logFC", "logCPM", "F", "PValue", "FDR"))) %>%
    mutate(sig = if ("FDR" %in% names(.)) FDR < params$fdr_cutoff else NA)
}

unwrap_sc_naive <- function(obj) {
  if (is.null(obj)) return(NULL)
  if (is.list(obj) && !is.data.frame(obj) && "unique" %in% names(obj)) {
    obj[["unique"]]
  } else {
    obj
  }
}

gene_tables <- list(
  lee_tdppos    = as_gene_tbl(lee_bulk_gene_naive),
  poly_kd       = as_gene_tbl(poly_bulk_gene_naive[["tdp43_kd_vs_nt_kd_ctrl"]]),
  poly_oe       = as_gene_tbl(poly_bulk_gene_naive[["tdp43_oe_vs_oe_ctrl"]]),
  sc_cluster12  = as_gene_tbl(unwrap_sc_naive(poly_sc_gene_naive))
)
gene_tables <- gene_tables[!sapply(gene_tables, is.null)]

gene_summary <- bind_rows(lapply(names(gene_tables), function(nm) {
  g <- gene_tables[[nm]]
  tibble(table = nm, n = nrow(g),
         n_sig = sum(g$sig, na.rm = TRUE),
         has_logFC = "logFC" %in% names(g))
}))
knitr::kable(gene_summary)
save_rds(gene_tables, "integrative_gene_tables")
```

```{r gene_pairs, fig.width = 5, fig.height = 5}
pair_gene <- function(tA, tB, nameA, nameB) {
  if (is.null(tA) || is.null(tB)) return(NULL)
  if (!"logFC" %in% names(tA) || !"logFC" %in% names(tB)) return(NULL)
  j <- inner_join(
    tA %>% rename(logFC_A = logFC, FDR_A = FDR, sig_A = sig),
    tB %>% rename(logFC_B = logFC, FDR_B = FDR, sig_B = sig),
    by = "feature_id"
  )
  if (nrow(j) < 5) return(NULL)
  list(pair = c(nameA, nameB), joined = j,
       pearson  = cor(j$logFC_A, j$logFC_B, method = "pearson"),
       spearman = cor(j$logFC_A, j$logFC_B, method = "spearman"),
       n = nrow(j),
       n_sig_both = sum(j$sig_A & j$sig_B, na.rm = TRUE),
       n_concordant_sig = sum(j$sig_A & j$sig_B &
                                sign(j$logFC_A) == sign(j$logFC_B), na.rm = TRUE))
}

gene_pair_results <- list()
for (p in pairs_of_interest) {
  r <- pair_gene(gene_tables[[p[1]]], gene_tables[[p[2]]], p[1], p[2])
  if (!is.null(r)) {
    gene_pair_results[[paste(p[1], "vs", p[2], sep = "__")]] <- r
  }
}

gene_summary_df <- bind_rows(lapply(gene_pair_results, function(r) {
  tibble(pair_A = r$pair[1], pair_B = r$pair[2],
         n = r$n, pearson = r$pearson, spearman = r$spearman,
         n_sig_both = r$n_sig_both,
         n_concordant_sig = r$n_concordant_sig)
}))
knitr::kable(gene_summary_df, digits = 3)
save_csv(gene_summary_df, "integrative_gene_concordance_summary")
save_rds(gene_pair_results, "integrative_gene_pair_results")

for (nm in names(gene_pair_results)) {
  r <- gene_pair_results[[nm]]
  j <- r$joined
  subtitle <- sprintf("n=%d  pearson=%.2f  spearman=%.2f  sig-both=%d (concord=%d)",
                      r$n, r$pearson, r$spearman,
                      r$n_sig_both, r$n_concordant_sig)
  print(
    ggplot(j, aes(logFC_A, logFC_B)) +
      geom_vline(xintercept = 0, color = "grey70") +
      geom_hline(yintercept = 0, color = "grey70") +
      geom_point(alpha = 0.3, size = 0.7) +
      labs(title = nm, subtitle = subtitle,
           x = paste0("logFC ", r$pair[1]),
           y = paste0("logFC ", r$pair[2])) +
      theme_bw()
  )
}
```

# Shared significant repeats, by category

For each pair and category, repeats significant in both tables, ordered by
summed absolute logFC. Biology annotation columns are carried through for
manual inspection.

```{r shared_hits}
collect_shared_hits <- function(lst, category) {
  lapply(lst, function(r) {
    r$joined %>%
      filter(sig_A, sig_B) %>%
      arrange(desc(abs(logFC_A) + abs(logFC_B))) %>%
      mutate(category = category,
             pair = paste(r$pair, collapse = " vs ")) %>%
      select(category, pair, feature_id, class_id, polya_class, a_run_class,
             age_class, logFC_A, FDR_A, logFC_B, FDR_B)
  })
}

shared_genic <- collect_shared_hits(pair_results_genic, "genic")
shared_inter <- collect_shared_hits(pair_results_inter, "intergenic")
shared_all <- bind_rows(c(shared_genic, shared_inter))
save_rds(list(genic = shared_genic, intergenic = shared_inter),
         "integrative_shared_hits")
save_csv(shared_all, "integrative_shared_hits")

for (cat_label in c("genic", "intergenic")) {
  lst <- if (cat_label == "genic") shared_genic else shared_inter
  for (nm in names(lst)) {
    cat("## ", cat_label, " - ", nm, "\n\n", sep = "")
    print(knitr::kable(head(lst[[nm]], 40)))
    cat("\n\n")
  }
}
```


# Concordance under gene-library-size normalization (no RUV)

Mirror of the RUV concordance above, run on per-dataset `gene_lib_repeat_de`
tables. Each dataset's repeat DGEList was built with library-size and
norm factors derived from its own gene count matrix, no RUV covariates.
Comparing the two concordance matrices side by side indicates how much of
the cross-study signal depends on RUV factor estimation.

```{r load_gene_lib}
poly_bulk_gl <- read_rds_safe(file.path(params$polymenidou_bulk_rds_dir,
                                        "gene_lib_repeat_de.rds"))
lee_bulk_gl  <- read_rds_safe(file.path(params$lee_bulk_rds_dir,
                                        "gene_lib_repeat_de.rds"))
poly_sc_gl   <- read_rds_safe(file.path(params$polymenidou_sc_rds_dir,
                                        "gene_lib_repeat_de.rds"))
```

```{r gene_lib_de_tables}
repeat_rds_map_gl <- list(
  lee_bulk = lee_bulk_gl,
  polymenidou_bulk = poly_bulk_gl,
  polymenidou_sc = poly_sc_gl
)

pick_repeat_tbl_gl <- function(study, contrast_id, category) {
  rds <- repeat_rds_map_gl[[study]]
  if (is.null(rds)) return(NULL)
  key <- repeat_key_map[[study]][[contrast_id]][[category]]
  if (is.null(key) || !key %in% names(rds)) return(NULL)
  df <- as_tibble(rds[[key]]) %>%
    select(feature_id, logFC, PValue, FDR) %>%
    mutate(sig = FDR < params$fdr_cutoff & abs(logFC) > params$logfc_cutoff)
  df
}

de_tables_gl <- list()
for (i in seq_len(nrow(contrast_info))) {
  lbl <- contrast_info$label[i]
  study <- contrast_info$study[i]
  cid <- contrast_info$contrast_id[i]
  for (cat in c("genic", "intergenic")) {
    tbl <- pick_repeat_tbl_gl(study, cid, cat)
    if (!is.null(tbl)) {
      de_tables_gl[[paste(lbl, cat, sep = "__")]] <- tbl
    }
  }
}

if (!is.null(bio)) {
  de_tables_gl <- lapply(de_tables_gl, function(df) {
    left_join(df, bio, by = c("feature_id" = "gene_id"))
  })
}
save_rds(de_tables_gl, "integrative_gene_lib_repeat_de_tables_annotated")
```

```{r gene_lib_pair_results}
run_pairs_for_category_gl <- function(category) {
  out <- list()
  for (p in pairs_of_interest) {
    tagA <- paste(p[1], category, sep = "__")
    tagB <- paste(p[2], category, sep = "__")
    r <- pair_logfc(de_tables_gl[[tagA]], de_tables_gl[[tagB]], p[1], p[2])
    if (!is.null(r)) {
      r$category <- category
      out[[paste(p[1], "vs", p[2], sep = "__")]] <- r
    }
  }
  out
}

pair_results_gl_genic <- run_pairs_for_category_gl("genic")
pair_results_gl_inter <- run_pairs_for_category_gl("intergenic")

summary_gl <- bind_rows(
  summarise_pair_results(pair_results_gl_genic, "genic"),
  summarise_pair_results(pair_results_gl_inter, "intergenic")
)
knitr::kable(summary_gl, digits = 3,
             caption = "Cross-study concordance with gene-library-size normalization")
save_csv(summary_gl, "integrative_gene_lib_repeat_concordance_summary")
save_rds(list(genic = pair_results_gl_genic,
              intergenic = pair_results_gl_inter),
         "integrative_gene_lib_repeat_pair_results")
```

```{r ruv_vs_gene_lib_side_by_side}
side_by_side <- summary_df %>%
  rename(pearson_ruv = pearson, spearman_ruv = spearman,
         n_sig_both_ruv = n_sig_both) %>%
  select(category, pair_A, pair_B, n, pearson_ruv, spearman_ruv, n_sig_both_ruv) %>%
  inner_join(
    summary_gl %>%
      rename(pearson_gl = pearson, spearman_gl = spearman,
             n_sig_both_gl = n_sig_both) %>%
      select(category, pair_A, pair_B, pearson_gl, spearman_gl, n_sig_both_gl),
    by = c("category", "pair_A", "pair_B")
  )
knitr::kable(side_by_side, digits = 3,
             caption = "Concordance side by side: RUV vs gene-library-size")
save_csv(side_by_side, "integrative_ruv_vs_gene_lib_concordance_side_by_side")
```

# Per-cluster sc counterpart of bulk OE

Cluster 12 is the authors' primary hypothesis for the HA-TDP-43-expressing
neurons, but nothing forces the bulk-OE-matching repeat signal to live there.
This section ranks every sc cluster by the Pearson and Spearman correlation
between its own cluster-vs-rest repeat logFC profile (gene-library-size
normalization, no RUV) and the Polymenidou bulk `tdp43_oe_vs_oe_ctrl` repeat
logFC profile (also gene-library-size). The cluster at the top of the
ranking is the empirical sc counterpart of bulk OE; if that is cluster 12
the paper story holds, and if it is not the choice deserves manual
inspection before further inference.

Bulk OE is the reference because it is a DOX-on vs DOX-off comparison in
the same Polymenidou system, so it should share the strongest biological
drivers with the sc library if any cluster in the sc library is dominated
by HA-TDP-43-expressing cells.

```{r per_cluster_bulk_oe_ranking, fig.width = 7, fig.height = 5}
per_cluster_gl <- read_rds_safe(file.path(params$polymenidou_sc_rds_dir,
                                          "per_cluster_gene_lib_repeat_de.rds"))

parse_cluster_tag <- function(tag) {
  m <- regmatches(tag, regexec("cluster(\\d+)_vs_other$", tag))[[1]]
  if (length(m) < 2) return(NA_integer_)
  as.integer(m[2])
}

parse_category_from_tag <- function(tag) {
  if (grepl("genic_repeats:", tag) && !grepl("intergenic_repeats:", tag)) "genic"
  else if (grepl("intergenic_repeats:", tag))                            "intergenic"
  else                                                                     NA_character_
}

bulk_oe_ref <- list(
  genic      = poly_bulk_gl[[repeat_key_map$polymenidou_bulk$tdp43_oe_vs_oe_ctrl$genic]],
  intergenic = poly_bulk_gl[[repeat_key_map$polymenidou_bulk$tdp43_oe_vs_oe_ctrl$intergenic]]
)

rank_rows <- list()
if (!is.null(per_cluster_gl) && length(per_cluster_gl) > 0) {
  for (tag in names(per_cluster_gl)) {
    cl <- parse_cluster_tag(tag)
    cat_ <- parse_category_from_tag(tag)
    if (is.na(cl) || is.na(cat_)) next
    ref <- bulk_oe_ref[[cat_]]
    if (is.null(ref)) next
    tbl <- as_tibble(per_cluster_gl[[tag]]) %>% select(feature_id, logFC)
    ref_tbl <- as_tibble(ref) %>% select(feature_id, logFC_bulk = logFC)
    j <- inner_join(tbl, ref_tbl, by = "feature_id")
    if (nrow(j) < 20) next
    rank_rows[[tag]] <- tibble(
      cluster = cl,
      category = cat_,
      tag = tag,
      n = nrow(j),
      pearson = cor(j$logFC, j$logFC_bulk, method = "pearson"),
      spearman = cor(j$logFC, j$logFC_bulk, method = "spearman")
    )
  }
}

rank_df <- bind_rows(rank_rows)
if (nrow(rank_df) > 0) {
  rank_df <- rank_df %>% arrange(category, desc(pearson))
  knitr::kable(rank_df, digits = 3,
               caption = "Per-cluster correlation against Polymenidou bulk OE")
  save_csv(rank_df, "integrative_per_cluster_vs_bulk_oe_ranking")
  save_rds(rank_df, "integrative_per_cluster_vs_bulk_oe_ranking")

  p_rank <- ggplot(rank_df,
                   aes(x = reorder(factor(cluster), pearson),
                       y = pearson, fill = cluster == 12L)) +
    geom_col() +
    facet_wrap(~ category, scales = "free_x") +
    coord_flip() +
    scale_fill_manual(values = c(`FALSE` = "grey60", `TRUE` = "firebrick"),
                      name = "cluster 12?") +
    labs(x = "sc cluster", y = "Pearson r vs Polymenidou bulk OE logFC",
         title = "Which sc cluster best matches bulk OE?") +
    theme_bw()
  print(p_rank)
} else {
  cat("No per-cluster DE tables available; skipping ranking.\n")
}
```

# Session info

```{r session}
sessionInfo()
```
