---
title: "Polymenidou lab TDP-43 KD/OE cultures (GSE230647, bulk): RUVg-normalized edgeR DE"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  counts_dir: "."
  report_outdir: "."
  indices_base: "."
  n_controls: 5000
  k_selected: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(edgeR)
  library(RUVSeq)
  library(matrixStats)
})

source("../workflow/scripts/ruv_common.R")

counts_dir <- params$counts_dir
report_outdir <- params$report_outdir
n_controls <- as.integer(params$n_controls)
k_selected <- as.integer(params$k_selected)
stopifnot(dir.exists(counts_dir), dir.exists(report_outdir),
          n_controls > 100, k_selected >= 1, k_selected <= 4)

save_rds <- function(obj, name) {
  saveRDS(obj, file.path(report_outdir, paste0(name, ".rds")))
}

csv_dir <- file.path(report_outdir, "csv")
dir.create(csv_dir, showWarnings = FALSE, recursive = TRUE)
save_csv <- function(df, name) {
  path <- file.path(csv_dir, paste0(name, ".csv"))
  write.csv(df, path, row.names = FALSE)
  cat("wrote", path, "(", nrow(df), "rows)\n")
  invisible(path)
}
```

Data-driven RUVg normalization for GSE230647 bulk repeat DE (iPSC-derived neural cultures with HA-TDP-43 DOX-inducible overexpression and shRNA knockdown, Polymenidou lab). Empirical control genes are selected from STAR gene counts by ranking least-significant genes in a naive edgeR F-test across the four-group design. The resulting W factors are plugged into edgeR repeat contrasts at each granularity and featureset. Inputs come from the method pipeline run with `workflow/configs/gse230647_bulk.yaml`.

# Inputs

Sample metadata derived from count matrix column names.

```{r meta}
samples <- c(
  "nt_kd_ctrl_r1","nt_kd_ctrl_r2","nt_kd_ctrl_r3","nt_kd_ctrl_r4",
  "oe_ctrl_r1","oe_ctrl_r2","oe_ctrl_r3","oe_ctrl_r4",
  "tdp43_kd_r1","tdp43_kd_r2","tdp43_kd_r3","tdp43_kd_r4",
  "tdp43_oe_r1","tdp43_oe_r2","tdp43_oe_r3","tdp43_oe_r4"
)
group <- factor(sub("_r[0-9]+$", "", samples),
                levels = c("nt_kd_ctrl","oe_ctrl","tdp43_kd","tdp43_oe"))
stopifnot(length(group) == length(samples), !any(is.na(group)),
          all(table(group) == 4))
```

Gene and repeat count matrices share sample columns. Asserted rather than
inferred from filenames.

```{r load}
genes <- load_counts_tsv(file.path(counts_dir, "star_genes_gene_id_unique.tsv"),
                         samples)

repeat_files <- list(
  all_family   = "star_repeats_family_id_unique.tsv",
  all_geneid   = "star_repeats_gene_id_unique.tsv",
  genic_family = "star_genic_repeats_family_id_unique.tsv",
  genic_geneid = "star_genic_repeats_gene_id_unique.tsv",
  inter_family = "star_intergenic_repeats_family_id_unique.tsv",
  inter_geneid = "star_intergenic_repeats_gene_id_unique.tsv"
)
repeats <- lapply(repeat_files, function(f)
  load_counts_tsv(file.path(counts_dir, f), samples))
stopifnot(all(vapply(repeats, function(m) identical(colnames(m), samples), TRUE)))
```

# Empirical control selection

Require count >= 10 in at least half the samples to avoid low-count noise
dominating the ranking.

```{r filter}
keep <- rowSums(genes >= 10) >= floor(ncol(genes) / 2)
genes_expr <- genes[keep, ]
stopifnot(nrow(genes_expr) > 5000)
```

Naive edgeR DE on the four-group design. An F-test across the three
non-reference coefficients flags genes moving in any contrast. Genes with
the highest p-values are the most stable.

```{r rank}
design0 <- model.matrix(~ group)
coefs_test <- setdiff(colnames(design0), "(Intercept)")
ctrl_res <- rank_empirical_controls(
  counts = genes_expr, design = design0,
  test = list(coef = coefs_test), n_controls = n_controls
)
tt0 <- ctrl_res$naive_de
controls <- ctrl_res$controls
save_rds(tt0, "ruv_naive_gene_de")
writeLines(controls, file.path(report_outdir, "ruv_empirical_controls.txt"))
```

# RUVg factor estimation

One W matrix per k in 1..4. An RLE-style mean-abs-deviation score summarises
residual sample heterogeneity; picking k by elbow.

```{r ruvg}
ruv_fit_res <- fit_ruv_range(
  genes_expr = genes_expr, controls = controls,
  pheno_df = data.frame(group, row.names = samples), ks = 1:4
)
ruv_fits <- ruv_fit_res$fits
diag_rle <- ruv_fit_res$diag_rle
knitr::kable(diag_rle, digits = 3)
write.table(diag_rle, file.path(report_outdir, "ruv_rle_diagnostic.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
```

```{r pick_k}
W_sel <- pick_W_matrix(ruv_fits, k_selected, samples)
W_all <- lapply(ruv_fits, function(s) pData(s))
save_rds(list(W_all = W_all, k_selected = k_selected, controls = controls,
              rle = diag_rle, samples = samples, group = group),
         "ruv_state")
```

# Repeat DE with RUV covariates

Two contrasts: tdp43_kd vs nt_kd_ctrl, tdp43_oe vs oe_ctrl. Each repeat
featureset and granularity fit with `~ W + group`.

```{r repeat_de}
build_contrast_vec <- function(design_colnames, group_levels, contrast_levels) {
  stopifnot(length(contrast_levels) == 2, all(contrast_levels %in% group_levels))
  v <- numeric(length(design_colnames))
  names(v) <- design_colnames
  num_col <- paste0("group", contrast_levels[1])
  den_col <- paste0("group", contrast_levels[2])
  stopifnot(num_col %in% names(v))
  v[num_col] <- 1
  if (contrast_levels[2] != group_levels[1]) {
    stopifnot(den_col %in% names(v))
    v[den_col] <- -1
  }
  v
}

contrasts <- list(
  tdp43_kd_vs_nt_kd_ctrl = c("tdp43_kd","nt_kd_ctrl"),
  tdp43_oe_vs_oe_ctrl    = c("tdp43_oe","oe_ctrl")
)

design <- model.matrix(~ W_sel + group)

all_results <- list()
for (fs in names(repeats)) {
  for (cname in names(contrasts)) {
    tag <- paste(fs, cname, sep = ":")
    cvec <- build_contrast_vec(colnames(design), levels(group), contrasts[[cname]])
    tt <- run_ruv_repeat_de(counts = repeats[[fs]], design = design,
                            test = list(contrast = cvec))
    out <- file.path(report_outdir,
                     sprintf("ruv_k%d_%s_%s.tsv", ncol(W_sel), fs, cname))
    write.table(tt, out, sep = "\t", quote = FALSE, row.names = FALSE)
    all_results[[tag]] <- tt
  }
}
stopifnot(length(all_results) == length(repeats) * length(contrasts))
save_rds(all_results, "ruv_repeat_de")
```

# Summary

```{r summary}
summary_rows <- do.call(rbind, lapply(names(all_results), function(tag) {
  r <- all_results[[tag]]
  data.frame(
    tag = tag,
    n_features = nrow(r),
    n_fdr_05 = sum(r$FDR < 0.05, na.rm = TRUE),
    n_fdr_10 = sum(r$FDR < 0.10, na.rm = TRUE),
    top_feature = r$feature_id[1],
    top_logfc = round(r$logFC[1], 3),
    top_fdr = signif(r$FDR[1], 3)
  )
}))
knitr::kable(summary_rows)
write.table(summary_rows, file.path(report_outdir, "ruv_summary.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
```
