---
title: "GSE126543 bulk RNA-seq repeats, RUVg-normalized edgeR DE"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  counts_dir: "."
  report_outdir: "."
  indices_base: "."
  n_controls: 5000
  k_selected: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(edgeR)
  library(RUVSeq)
})

counts_dir <- params$counts_dir
report_outdir <- params$report_outdir
n_controls <- as.integer(params$n_controls)
k_selected <- as.integer(params$k_selected)
stopifnot(dir.exists(counts_dir), dir.exists(report_outdir),
          n_controls > 100, k_selected >= 1, k_selected <= 4)

save_rds <- function(obj, name) {
  saveRDS(obj, file.path(report_outdir, paste0(name, ".rds")))
}
```

Data-driven RUVg normalization for gse126543 (tdp43pos vs tdp43neg sorted
nuclei). Unsorted samples are dropped because they are not part of the
contrast and would inflate W with non-biological group structure. STAR
unique-count matrices for genes and repeats.

# Inputs

```{r meta}
samples <- c(paste0("tdp43neg_s", 1:7), paste0("tdp43pos_s", 1:7))
group <- factor(sub("_s[0-9]+$", "", samples),
                levels = c("tdp43neg","tdp43pos"))
stopifnot(length(group) == length(samples), !any(is.na(group)),
          table(group)["tdp43neg"] == 7, table(group)["tdp43pos"] == 7)
```

```{r load}
load_counts <- function(path, samples) {
  stopifnot(file.exists(path))
  m <- read.table(path, header = TRUE, sep = "\t", row.names = 1,
                  check.names = FALSE, stringsAsFactors = FALSE)
  m <- as.matrix(m)
  stopifnot(all(samples %in% colnames(m)))
  m <- m[, samples, drop = FALSE]
  storage.mode(m) <- "integer"
  stopifnot(all(is.finite(m)), all(m >= 0))
  m
}

genes <- load_counts(file.path(counts_dir, "star_genes_gene_id_unique.tsv"),
                     samples)

repeat_files <- list(
  all_family   = "star_repeats_family_id_unique.tsv",
  all_geneid   = "star_repeats_gene_id_unique.tsv",
  genic_family = "star_genic_repeats_family_id_unique.tsv",
  genic_geneid = "star_genic_repeats_gene_id_unique.tsv",
  inter_family = "star_intergenic_repeats_family_id_unique.tsv",
  inter_geneid = "star_intergenic_repeats_gene_id_unique.tsv"
)
repeats <- lapply(repeat_files, function(f)
  load_counts(file.path(counts_dir, f), samples))
stopifnot(all(vapply(repeats, function(m) identical(colnames(m), samples), TRUE)))
```

# Empirical control selection

Two-group naive DE ranks genes by p-value for the pos-vs-neg contrast; top
p-values define the empirical stable-gene panel.

```{r filter_rank}
keep <- rowSums(genes >= 10) >= floor(ncol(genes) / 2)
genes_expr <- genes[keep, ]
stopifnot(nrow(genes_expr) > 5000)

design0 <- model.matrix(~ group)
stopifnot(qr(design0)$rank == ncol(design0))
dge0 <- DGEList(counts = genes_expr, group = group)
dge0 <- calcNormFactors(dge0, method = "TMM")
dge0 <- estimateDisp(dge0, design0)
fit0 <- glmQLFit(dge0, design0)
qlf0 <- glmQLFTest(fit0, coef = "grouptdp43pos")
tt0 <- topTags(qlf0, n = Inf, sort.by = "none")$table
stopifnot(nrow(tt0) == nrow(genes_expr))
save_rds(tt0, "ruv_naive_gene_de")
```

```{r controls}
stopifnot(n_controls <= nrow(tt0))
controls <- rownames(tt0)[order(tt0$PValue, decreasing = TRUE)][seq_len(n_controls)]
stopifnot(length(controls) == n_controls, !anyDuplicated(controls))
writeLines(controls, file.path(report_outdir, "ruv_empirical_controls.txt"))
```

# RUVg factor estimation

```{r ruvg}
set1 <- newSeqExpressionSet(counts = genes_expr,
                            phenoData = data.frame(group, row.names = samples))
rle_score <- function(mat) {
  med <- matrixStats::rowMedians(log2(mat + 1))
  rle <- sweep(log2(mat + 1), 1, med, "-")
  mean(apply(rle, 2, function(x) mean(abs(x - median(x)))))
}

ruv_fits <- list()
for (k in 1:4) {
  ruv_fits[[as.character(k)]] <- RUVg(set1, controls, k = k)
}
rle_before <- rle_score(genes_expr)
rle_after <- vapply(ruv_fits, function(s) rle_score(normCounts(s)), numeric(1))
diag_rle <- data.frame(k = 0:4,
                       rle_mean_abs_dev = c(rle_before, rle_after))
knitr::kable(diag_rle, digits = 3)
write.table(diag_rle, file.path(report_outdir, "ruv_rle_diagnostic.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
```

```{r pick_k}
W_all <- lapply(ruv_fits, function(s) pData(s))
W_sel <- as.matrix(W_all[[as.character(k_selected)]][,
          grep("^W_", colnames(W_all[[as.character(k_selected)]])), drop = FALSE])
stopifnot(nrow(W_sel) == length(samples), identical(rownames(W_sel), samples),
          ncol(W_sel) == k_selected)
save_rds(list(W_all = W_all, k_selected = k_selected, controls = controls,
              rle = diag_rle, samples = samples, group = group),
         "ruv_state")
```

# Repeat DE with RUV covariates

```{r repeat_de}
run_contrast <- function(counts, W, group, label, featureset) {
  design <- model.matrix(~ W + group)
  stopifnot(qr(design)$rank == ncol(design))
  d <- DGEList(counts = counts, group = group)
  d <- calcNormFactors(d, method = "TMM")
  d <- estimateDisp(d, design)
  fit <- glmQLFit(d, design)
  qlf <- glmQLFTest(fit, coef = "grouptdp43pos")
  tt <- topTags(qlf, n = Inf, sort.by = "none")$table
  tt$feature_id <- rownames(tt)
  tt <- tt[order(tt$PValue), c("feature_id","logFC","logCPM","F","PValue","FDR")]
  out <- file.path(report_outdir,
                   sprintf("ruv_k%d_%s_%s.tsv", ncol(W), featureset, label))
  write.table(tt, out, sep = "\t", quote = FALSE, row.names = FALSE)
  tt
}

all_results <- list()
for (fs in names(repeats)) {
  all_results[[fs]] <- run_contrast(
    counts = repeats[[fs]], W = W_sel, group = group,
    label = "tdp43pos_vs_tdp43neg", featureset = fs
  )
}
stopifnot(length(all_results) == length(repeats))
save_rds(all_results, "ruv_repeat_de")
```

# Summary

```{r summary}
summary_rows <- do.call(rbind, lapply(names(all_results), function(fs) {
  r <- all_results[[fs]]
  data.frame(
    featureset = fs,
    n_features = nrow(r),
    n_fdr_05 = sum(r$FDR < 0.05, na.rm = TRUE),
    n_fdr_10 = sum(r$FDR < 0.10, na.rm = TRUE),
    top_feature = r$feature_id[1],
    top_logfc = round(r$logFC[1], 3),
    top_fdr = signif(r$FDR[1], 3)
  )
}))
knitr::kable(summary_rows)
write.table(summary_rows, file.path(report_outdir, "ruv_summary.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
```
