---
title: "Lee et al., nuclear TDP-43 retention (GSE126543, bulk): RUVg-normalized edgeR DE"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  counts_dir: "."
  report_outdir: "."
  indices_base: "."
  n_controls: 5000
  k_selected: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(edgeR)
  library(RUVSeq)
  library(matrixStats)
})

source("../workflow/scripts/ruv_common.R")

counts_dir <- params$counts_dir
report_outdir <- params$report_outdir
n_controls <- as.integer(params$n_controls)
k_selected <- as.integer(params$k_selected)
stopifnot(dir.exists(counts_dir), dir.exists(report_outdir),
          n_controls > 100, k_selected >= 1, k_selected <= 4)

save_rds <- function(obj, name) {
  saveRDS(obj, file.path(report_outdir, paste0(name, ".rds")))
}
```

Data-driven RUVg normalization for GSE126543 (Liu, Russ, Cali, Phan, Amlie-Wolf, Lee; "Loss of Nuclear TDP-43 Is Associated with Decondensation of LINE Retrotransposons"). Contrast: tdp43pos vs tdp43neg sorted nuclei. Unsorted samples are dropped because they are not part of the contrast and would inflate W with non-biological group structure. Inputs are STAR unique-count matrices for genes and repeats produced by the method pipeline with `workflow/configs/gse126543_bulk.yaml`.

# Inputs

```{r meta}
samples <- c(paste0("tdp43neg_s", 1:7), paste0("tdp43pos_s", 1:7))
group <- factor(sub("_s[0-9]+$", "", samples),
                levels = c("tdp43neg","tdp43pos"))
stopifnot(length(group) == length(samples), !any(is.na(group)),
          table(group)["tdp43neg"] == 7, table(group)["tdp43pos"] == 7)
```

```{r load}
genes <- load_counts_tsv(file.path(counts_dir, "star_genes_gene_id_unique.tsv"),
                         samples, require_all = FALSE)

repeat_files <- list(
  all_family   = "star_repeats_family_id_unique.tsv",
  all_geneid   = "star_repeats_gene_id_unique.tsv",
  genic_family = "star_genic_repeats_family_id_unique.tsv",
  genic_geneid = "star_genic_repeats_gene_id_unique.tsv",
  inter_family = "star_intergenic_repeats_family_id_unique.tsv",
  inter_geneid = "star_intergenic_repeats_gene_id_unique.tsv"
)
repeats <- lapply(repeat_files, function(f)
  load_counts_tsv(file.path(counts_dir, f), samples, require_all = FALSE))
stopifnot(all(vapply(repeats, function(m) identical(colnames(m), samples), TRUE)))
```

# Empirical control selection

Two-group naive DE ranks genes by p-value for the pos-vs-neg contrast; top
p-values define the empirical stable-gene panel.

```{r filter_rank}
keep <- rowSums(genes >= 10) >= floor(ncol(genes) / 2)
genes_expr <- genes[keep, ]
stopifnot(nrow(genes_expr) > 5000)

design0 <- model.matrix(~ group)
ctrl_res <- rank_empirical_controls(
  counts = genes_expr, design = design0,
  test = list(coef = "grouptdp43pos"), n_controls = n_controls
)
tt0 <- ctrl_res$naive_de
controls <- ctrl_res$controls
save_rds(tt0, "ruv_naive_gene_de")
writeLines(controls, file.path(report_outdir, "ruv_empirical_controls.txt"))
```

# RUVg factor estimation

```{r ruvg}
ruv_fit_res <- fit_ruv_range(
  genes_expr = genes_expr, controls = controls,
  pheno_df = data.frame(group, row.names = samples), ks = 1:4
)
ruv_fits <- ruv_fit_res$fits
diag_rle <- ruv_fit_res$diag_rle
knitr::kable(diag_rle, digits = 3)
write.table(diag_rle, file.path(report_outdir, "ruv_rle_diagnostic.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
```

```{r pick_k}
W_sel <- pick_W_matrix(ruv_fits, k_selected, samples)
W_all <- lapply(ruv_fits, function(s) pData(s))
save_rds(list(W_all = W_all, k_selected = k_selected, controls = controls,
              rle = diag_rle, samples = samples, group = group),
         "ruv_state")
```

# Repeat DE with RUV covariates

```{r repeat_de}
design <- model.matrix(~ W_sel + group)
label <- "tdp43pos_vs_tdp43neg"

all_results <- list()
for (fs in names(repeats)) {
  tt <- run_ruv_repeat_de(counts = repeats[[fs]], design = design,
                          test = list(coef = "grouptdp43pos"))
  out <- file.path(report_outdir,
                   sprintf("ruv_k%d_%s_%s.tsv", ncol(W_sel), fs, label))
  write.table(tt, out, sep = "\t", quote = FALSE, row.names = FALSE)
  all_results[[fs]] <- tt
}
stopifnot(length(all_results) == length(repeats))
save_rds(all_results, "ruv_repeat_de")
```

# Summary

```{r summary}
summary_rows <- do.call(rbind, lapply(names(all_results), function(fs) {
  r <- all_results[[fs]]
  data.frame(
    featureset = fs,
    n_features = nrow(r),
    n_fdr_05 = sum(r$FDR < 0.05, na.rm = TRUE),
    n_fdr_10 = sum(r$FDR < 0.10, na.rm = TRUE),
    top_feature = r$feature_id[1],
    top_logfc = round(r$logFC[1], 3),
    top_fdr = signif(r$FDR[1], 3)
  )
}))
knitr::kable(summary_rows)
write.table(summary_rows, file.path(report_outdir, "ruv_summary.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
```
