---
title: "Polymenidou lab TDP-43 OE cultures (GSE230647, scRNA-seq): RUVg-normalized edgeR DE on pseudo-bulk"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  starsolo_base: "."
  metadata: "."
  sample_metadata: "."
  report_outdir: "."
  indices_base: "."
  feature_sets: "genes genic_repeats intergenic_repeats"
  starsolo_modes: "unique"
  n_controls: 3000
  k_selected: 1
  min_cells_per_col: 10
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(Matrix)
  library(edgeR)
  library(RUVSeq)
  library(matrixStats)
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggrepel)
})

source("../workflow/scripts/ruv_common.R")
source("../workflow/scripts/gene_lib_size_common.R")
source("../workflow/scripts/ggtheme.R")
source("../workflow/scripts/ruv_plots.R")

starsolo_base <- params$starsolo_base
report_outdir <- params$report_outdir
indices_base <- params$indices_base
n_controls <- as.integer(params$n_controls)
k_selected <- as.integer(params$k_selected)
min_cells_per_col <- as.integer(params$min_cells_per_col)
feature_sets <- strsplit(params$feature_sets, " ")[[1]]
starsolo_modes <- strsplit(params$starsolo_modes, " ")[[1]]
repeat_fsets <- feature_sets[feature_sets != "genes"]

stopifnot(dir.exists(starsolo_base), dir.exists(report_outdir),
          n_controls >= 500, k_selected >= 1,
          min_cells_per_col >= 1,
          "genes" %in% feature_sets, length(repeat_fsets) >= 1,
          length(starsolo_modes) >= 1)

save_rds <- function(obj, name) {
  saveRDS(obj, file.path(report_outdir, paste0(name, ".rds")))
}

csv_dir <- file.path(report_outdir, "csv")
dir.create(csv_dir, showWarnings = FALSE, recursive = TRUE)
save_csv <- function(df, name) {
  path <- file.path(csv_dir, paste0(name, ".csv"))
  write.csv(df, path, row.names = FALSE)
  cat("wrote", path, "(", nrow(df), "rows)\n")
  invisible(path)
}
```

## Biology annotation for DE stratification

```{r biology-setup}
suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(readr)
})
source("../workflow/scripts/biology_checks.R")

biology_annotation_path <- file.path(indices_base, "repeat_biology_annotation.tsv")
biology_annotation <- load_biology_annotation(biology_annotation_path)
if (is.null(biology_annotation)) {
  cat("No biology annotation at", biology_annotation_path,
      "- biology-stratified panels will be skipped.\n")
} else {
  cat("Loaded biology annotation:", nrow(biology_annotation), "subfamilies\n")
}
save_rds(biology_annotation, "biology_annotation")

annotate_de_with_biology <- function(tt) {
  rownames(tt) <- tt$feature_id
  tt <- annotate_biology(tt, biology_annotation, key = "gene_id")
  tt$polya <- polya_label(tt$class_id)
  tt$age <- family_age_label(tt$family_age)
  tt$strand <- strand_vs_gene_label(tt$strand_vs_gene)
  tt$a_run <- a_run_label(tt$frac_downstream_a_run)
  tt
}

de_stratify_counts <- function(tt, key, fdr = 0.05) {
  tt <- annotate_de_with_biology(tt)
  tt$direction <- ifelse(is.na(tt$FDR) | tt$FDR >= fdr, "ns",
                         ifelse(tt$logFC > 0, "up", "down"))
  facet_cols <- c(polya = "polya", family_age = "age",
                  strand_vs_gene = "strand", downstream_a_run = "a_run")
  rows <- list()
  for (facet_nm in names(facet_cols)) {
    col <- facet_cols[[facet_nm]]
    rows[[facet_nm]] <- tt %>%
      group_by(label = .data[[col]], direction) %>%
      summarize(n = dplyr::n(), .groups = "drop") %>%
      mutate(facet = facet_nm, key = key)
  }
  bind_rows(rows)
}

plot_de_stratification <- function(strat_df, title) {
  if (is.null(strat_df) || nrow(strat_df) == 0) return(invisible(NULL))
  strat_df$direction <- factor(strat_df$direction, levels = c("up", "down", "ns"))
  ggplot(strat_df, aes(x = label, y = n, fill = direction)) +
    geom_col() +
    facet_grid(facet ~ key, scales = "free", space = "free_x") +
    scale_fill_manual(values = c(up = "firebrick", down = "steelblue", ns = "grey70")) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = title, x = NULL, y = "features")
}
```

Paper counterpart of the sc method report's RUVg block. This script owns its own RUVg analysis end-to-end, parallel to the two bulk paper Rmds. Data: GSE230647 (Polymenidou lab), three DOX-on HA-TDP-43 iPSC-derived neural cultures (2-week and 4-week time points). Cluster 12 in the GEO cell metadata flags the TDP-43-HA-expressing neurons and acts as the neuron-resolved counterpart of the bulk `tdp43_oe vs oe_ctrl` contrast. Inputs are STARsolo raw matrices from `results/gse230647_sc/starsolo/` (produced by the method pipeline with `workflow/configs/gse230647_sc.yaml`) plus the GEO cell metadata file. No counts are re-derived from raw reads.

Pseudo-bulking is per (sample, cluster) pair: each cluster gets its own pseudo-bulk column per sample, so with three samples and many GEO-annotated clusters the design ends up with dozens of columns. The binary test contrast is cluster 12 versus any non-12 cluster, encoded as a single `group` factor (`TRUE` when the pseudo-bulk column is cluster 12, `FALSE` otherwise). The sample of origin is absorbed via `sample_block`. Pairs with fewer than `min_cells_per_col` cells are dropped to keep pseudo-bulk columns informative.

# Sample metadata

```{r sample-meta}
sample_meta <- read_tsv(params$sample_metadata, show_col_types = FALSE)
sample_ids <- sample_meta$sample_id
stopifnot(length(sample_ids) == 3, !anyDuplicated(sample_ids))
knitr::kable(sample_meta)
```

# Cell-type annotations from GEO

```{r load-geo-meta}
geo_meta <- read_tsv(params$metadata, show_col_types = FALSE)
stopifnot("biorxiv_v2_cluster_number" %in% colnames(geo_meta),
          "biorxiv_v2_cluster_name" %in% colnames(geo_meta))

barcode_col <- grep("barcode|cell|cb", colnames(geo_meta),
                    ignore.case = TRUE, value = TRUE)[1]
stopifnot(!is.na(barcode_col))

geo_meta$cluster_num <- as.integer(geo_meta$biorxiv_v2_cluster_number)
geo_meta$cluster <- geo_meta$biorxiv_v2_cluster_name
geo_meta$barcode_16bp <- sub("-[0-9]+\\..*$", "", geo_meta[[barcode_col]])
geo_meta$sample_tag <- sub("^[^.]*\\.", "", geo_meta[[barcode_col]])

rep_suffix <- ifelse(is.na(sample_meta$replicate) | sample_meta$replicate == "",
                     "", tolower(sample_meta$replicate))
tag_to_sid <- setNames(sample_meta$sample_id,
                       paste0("TDP", sample_meta$culture_weeks, "wON", rep_suffix))
geo_meta$sample_id <- unname(tag_to_sid[geo_meta$sample_tag])

geo_meta$group <- ifelse(!is.na(geo_meta$cluster_num) & geo_meta$cluster_num == 12L,
                         "cluster12", "other")

cat("cluster 12 cells:", sum(geo_meta$group == "cluster12", na.rm = TRUE), "\n")
cat("other cells:", sum(geo_meta$group == "other", na.rm = TRUE), "\n")
save_rds(geo_meta, "geo_meta_annotated")
```

```{r cluster-sizes}
cluster_sizes <- geo_meta %>%
  filter(!is.na(sample_id), !is.na(cluster_num)) %>%
  count(sample_id, cluster_num, cluster, name = "n_cells") %>%
  arrange(sample_id, cluster_num)
save_csv(cluster_sizes, "cluster_sizes_per_sample")

group_sizes <- geo_meta %>%
  filter(!is.na(sample_id), !is.na(cluster_num)) %>%
  count(sample_id, group, name = "n_cells") %>%
  pivot_wider(names_from = group, values_from = n_cells, values_fill = 0)
save_csv(group_sizes, "group_sizes_per_sample")
```

# Pseudo-bulk from STARsolo raw matrices

```{r pseudobulk-col-meta}
col_meta <- geo_meta %>%
  filter(!is.na(sample_id), !is.na(cluster_num)) %>%
  count(sample_id, cluster_num, cluster, name = "n_cells") %>%
  filter(n_cells >= min_cells_per_col) %>%
  arrange(sample_id, cluster_num) %>%
  mutate(col_tag = sprintf("%s__c%02d", sample_id, cluster_num),
         cluster_name = cluster,
         is_cluster12 = cluster_num == 12L)

stopifnot(!anyDuplicated(col_meta$col_tag),
          any(col_meta$is_cluster12),
          any(!col_meta$is_cluster12))

cat("pseudo-bulk columns:", nrow(col_meta),
    "| cluster 12 cols:", sum(col_meta$is_cluster12),
    "| other cols:", sum(!col_meta$is_cluster12), "\n")
save_csv(col_meta, "pseudobulk_col_meta")
```

```{r load-starsolo}
pseudobulk_by_cluster <- function(raw_dir, sample_id, geo_sample, cols_for_sample) {
  stopifnot(dir.exists(raw_dir))
  barcodes <- readLines(file.path(raw_dir, "barcodes.tsv"))
  features <- read_tsv(file.path(raw_dir, "features.tsv"),
                       col_names = c("id", "name", "type"),
                       show_col_types = FALSE)
  mat <- readMM(file.path(raw_dir, "matrix.mtx"))
  rownames(mat) <- features$id
  colnames(mat) <- barcodes

  matched_idx <- match(barcodes, geo_sample$barcode_16bp)
  cluster_of_cell <- geo_sample$cluster_num[matched_idx]

  out <- matrix(0L, nrow = nrow(mat), ncol = nrow(cols_for_sample),
                dimnames = list(rownames(mat), cols_for_sample$col_tag))
  for (i in seq_len(nrow(cols_for_sample))) {
    cl <- cols_for_sample$cluster_num[i]
    sel <- !is.na(cluster_of_cell) & cluster_of_cell == cl
    out[, i] <- as.integer(Matrix::rowSums(mat[, sel, drop = FALSE]))
  }
  out
}

pb_list <- list()
for (mode in starsolo_modes) {
  for (fs in feature_sets) {
    key <- paste(mode, fs, sep = "_")
    per_sample_mats <- list()
    for (sid in sample_ids) {
      cols_for_sample <- col_meta[col_meta$sample_id == sid, , drop = FALSE]
      if (nrow(cols_for_sample) == 0) next
      raw_dir <- file.path(starsolo_base, sid,
                           paste0(mode, "_", fs),
                           "Solo.out", "Gene", "raw")
      geo_sample <- geo_meta[!is.na(geo_meta$sample_id) &
                             geo_meta$sample_id == sid, ]
      stopifnot(nrow(geo_sample) > 0)
      mat_sid <- pseudobulk_by_cluster(raw_dir, sid, geo_sample, cols_for_sample)
      per_sample_mats[[sid]] <- mat_sid
      cat(key, sid, "| cols:", ncol(mat_sid),
          "| total_counts:", sum(mat_sid), "\n")
    }
    features_union <- Reduce(union, lapply(per_sample_mats, rownames))
    combined <- matrix(0L, nrow = length(features_union), ncol = nrow(col_meta),
                       dimnames = list(features_union, col_meta$col_tag))
    for (sid in names(per_sample_mats)) {
      m <- per_sample_mats[[sid]]
      combined[rownames(m), colnames(m)] <- m
    }
    pb_list[[key]] <- combined
  }
}
save_rds(pb_list, "pseudobulk_counts")
```

```{r pseudobulk-libsizes}
lib_rows <- list()
for (mode in starsolo_modes) {
  gene_key <- paste(mode, "genes", sep = "_")
  stopifnot(gene_key %in% names(pb_list))
  gene_pb <- pb_list[[gene_key]]
  stopifnot(identical(colnames(gene_pb), col_meta$col_tag))
  lib_rows[[mode]] <- data.frame(
    mode = mode,
    col_tag = col_meta$col_tag,
    sample_id = col_meta$sample_id,
    cluster_num = col_meta$cluster_num,
    cluster_name = col_meta$cluster_name,
    n_cells = col_meta$n_cells,
    is_cluster12 = col_meta$is_cluster12,
    gene_lib_size = unname(colSums(gene_pb)))
}
save_csv(bind_rows(lib_rows), "pseudobulk_gene_lib_sizes")
```

# Empirical control selection

Per-mode naive edgeR QL F-test on the pseudo-bulk gene matrix. Design is `~ sample_block + group`, where `sample_block = factor(sample_id)` and `group = factor(is_cluster12)` (`TRUE` for cluster 12 columns, `FALSE` otherwise). The tested coefficient is `groupTRUE`. The top `n_controls` genes with the largest p-values become empirical controls.

```{r rank-controls}
group <- factor(col_meta$is_cluster12, levels = c(FALSE, TRUE))
sample_block <- factor(col_meta$sample_id)
pb_samples <- col_meta$col_tag

ruv_state_by_mode <- list()
naive_de_by_mode <- list()
for (mode in starsolo_modes) {
  gene_key <- paste(mode, "genes", sep = "_")
  genes_mat <- pb_list[[gene_key]]
  stopifnot(identical(colnames(genes_mat), pb_samples))
  storage.mode(genes_mat) <- "integer"
  keep <- rowSums(genes_mat >= 10) >= floor(ncol(genes_mat) / 2)
  genes_expr <- genes_mat[keep, , drop = FALSE]
  cat(mode, ": filtered gene matrix", nrow(genes_expr), "x", ncol(genes_expr), "\n")

  n_ctrl <- min(n_controls, nrow(genes_expr) - 10)
  stopifnot(n_ctrl >= 500)

  design0 <- model.matrix(~ sample_block + group)
  stopifnot(qr(design0)$rank == ncol(design0),
            "groupTRUE" %in% colnames(design0))
  ctrl_res <- rank_empirical_controls(
    counts = genes_expr, design = design0,
    test = list(coef = "groupTRUE"), n_controls = n_ctrl
  )
  naive_de_by_mode[[mode]] <- ctrl_res$naive_de
  writeLines(ctrl_res$controls,
             file.path(report_outdir, paste0("ruv_empirical_controls_", mode, ".txt")))

  ruv_state_by_mode[[mode]] <- list(
    pb_samples = pb_samples, group = group, sample_block = sample_block,
    genes_expr = genes_expr, controls = ctrl_res$controls, n_ctrl = n_ctrl
  )
}
save_rds(naive_de_by_mode, "ruv_naive_gene_de")
```

# RUVg factor estimation

k runs from 1 to 4. The operating k is fixed via `k_selected` (default 1). Residual df is comfortable because the column count now equals the number of (sample, cluster) pairs passing `min_cells_per_col` (tens of columns), not 6.

```{r ruvg}
for (mode in starsolo_modes) {
  st <- ruv_state_by_mode[[mode]]
  ruv_fit_res <- fit_ruv_range(
    genes_expr = st$genes_expr, controls = st$controls,
    pheno_df = data.frame(group = st$group, row.names = st$pb_samples),
    ks = 1:4
  )
  st$diag_rle <- ruv_fit_res$diag_rle
  st$fits <- ruv_fit_res$fits
  write.table(ruv_fit_res$diag_rle,
              file.path(report_outdir, paste0("ruv_rle_diagnostic_", mode, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  knitr::kable(ruv_fit_res$diag_rle, digits = 3,
               caption = paste("RLE diagnostic, mode =", mode)) |> print()

  W <- pick_W_matrix(ruv_fit_res$fits, k_selected, st$pb_samples)
  sample_block <- st$sample_block
  group <- st$group
  design <- model.matrix(~ sample_block + W + group)
  stopifnot(qr(design)$rank == ncol(design))
  resid_df <- ncol(st$genes_expr) - qr(design)$rank
  cat(mode, "| design rank", qr(design)$rank,
      "| residual df", resid_df, "\n")
  st$W <- W
  st$design <- design
  st$resid_df <- resid_df
  ruv_state_by_mode[[mode]] <- st
}
save_rds(ruv_state_by_mode, "ruv_state")
```

```{r w-factors-csv}
w_rows <- list()
rle_rows <- list()
for (mode in names(ruv_state_by_mode)) {
  st <- ruv_state_by_mode[[mode]]
  wdf <- as.data.frame(st$W)
  wdf$col_tag <- rownames(st$W)
  wdf$mode <- mode
  wdf$sample_id <- col_meta$sample_id[match(wdf$col_tag, col_meta$col_tag)]
  wdf$cluster_num <- col_meta$cluster_num[match(wdf$col_tag, col_meta$col_tag)]
  wdf$cluster_name <- col_meta$cluster_name[match(wdf$col_tag, col_meta$col_tag)]
  wdf$n_cells <- col_meta$n_cells[match(wdf$col_tag, col_meta$col_tag)]
  wdf$is_cluster12 <- col_meta$is_cluster12[match(wdf$col_tag, col_meta$col_tag)]
  w_rows[[mode]] <- wdf
  rle_df <- st$diag_rle
  rle_df$mode <- mode
  rle_rows[[mode]] <- rle_df
}
save_csv(bind_rows(w_rows), "ruv_W_factors")
save_csv(bind_rows(rle_rows), "ruv_rle_diagnostic")
```

# Repeat DE with RUV covariates

Per (mode, featureset), contrast `groupTRUE` on the pseudo-bulk repeat counts under `~ sample_block + W + group`.

```{r repeat-de}
all_results <- list()
for (mode in starsolo_modes) {
  st <- ruv_state_by_mode[[mode]]
  for (fs in repeat_fsets) {
    key <- paste(mode, fs, sep = "_")
    counts <- pb_list[[key]]
    stopifnot(identical(colnames(counts), st$pb_samples))
    storage.mode(counts) <- "integer"
    stopifnot("groupTRUE" %in% colnames(st$design))
    tt <- run_ruv_repeat_de(counts = counts, design = st$design,
                            test = list(coef = "groupTRUE"))
    tag <- paste(key, "cluster12_vs_other", sep = ":")
    out_tsv <- file.path(report_outdir,
                         sprintf("ruv_k%d_%s_%s_cluster12_vs_other.tsv",
                                 k_selected, mode, fs))
    write.table(tt, out_tsv, sep = "\t", quote = FALSE, row.names = FALSE)
    all_results[[tag]] <- tt
  }
}
stopifnot(length(all_results) == length(starsolo_modes) * length(repeat_fsets))
save_rds(all_results, "ruv_repeat_de")
```

## Repeat DE with gene-library-size normalization (no RUV)

Parallel baseline to the RUV analysis above, for the same cluster 12 vs
other contrast. Library-size factors come from TMM on the pseudo-bulk gene
matrix (`ruv_state_by_mode[[mode]]$genes_expr`) and are transferred to each
repeat DGEList. No RUV covariates are added. TMM on the repeat matrix is
avoided because TDP-43 perturbation moves many repeat subfamilies in
concert. Concordance between `gene_lib_repeat_de` and `ruv_repeat_de` is
reported in the integrative cross-study Rmd.

```{r gene_lib_repeat_de_cluster12}
all_results_gene_lib <- list()
for (mode in starsolo_modes) {
  st <- ruv_state_by_mode[[mode]]
  design_no_ruv <- model.matrix(~ sample_block + group)
  stopifnot(qr(design_no_ruv)$rank == ncol(design_no_ruv),
            "groupTRUE" %in% colnames(design_no_ruv))
  for (fs in repeat_fsets) {
    key <- paste(mode, fs, sep = "_")
    counts <- pb_list[[key]]
    stopifnot(identical(colnames(counts), st$pb_samples))
    storage.mode(counts) <- "integer"
    tt <- run_gene_lib_repeat_de(repeat_counts = counts,
                                 gene_counts = st$genes_expr,
                                 design = design_no_ruv,
                                 test = list(coef = "groupTRUE"))
    tag <- paste(key, "cluster12_vs_other", sep = ":")
    out_tsv <- file.path(report_outdir,
                         sprintf("gene_lib_%s_%s_cluster12_vs_other.tsv",
                                 mode, fs))
    write.table(tt, out_tsv, sep = "\t", quote = FALSE, row.names = FALSE)
    all_results_gene_lib[[tag]] <- tt
  }
}
stopifnot(length(all_results_gene_lib) == length(starsolo_modes) * length(repeat_fsets))
save_rds(all_results_gene_lib, "gene_lib_repeat_de")
```

## Per-cluster vs rest repeat DE (gene-lib only)

Cluster 12 is the authors' primary hypothesis for the HA-TDP-43-expressing
population. To test that choice empirically, we run the same repeat DE for
every cluster individually ("cluster c vs all others") and later correlate
each cluster's repeat logFC profile against Polymenidou bulk OE in the
integrative Rmd. The cluster that best matches bulk OE is the empirical sc
counterpart; if that is not cluster 12, the choice deserves scrutiny.

Only the gene-library-size path is run here. RUV-per-cluster would require
re-ranking empirical control genes for each cluster's contrast (since RUV-g
controls are selected as the genes least associated with the specific
contrast under test), which is a different analysis in its own right. We
keep the RUV path for the primary cluster-12 contrast only.

A cluster is included if at least two samples have more than
`min_cells_per_col` cells in it, so each pseudo-bulk contrast has
non-trivial replication.

```{r per_cluster_gene_lib_repeat_de}
cluster_support <- col_meta %>%
  group_by(cluster_num) %>%
  summarize(n_samples = n_distinct(sample_id),
            total_cells = sum(n_cells),
            .groups = "drop") %>%
  filter(n_samples >= 2)

clusters_to_sweep <- sort(cluster_support$cluster_num)
cat("clusters to sweep:", length(clusters_to_sweep),
    "|", paste(clusters_to_sweep, collapse = ","), "\n")

per_cluster_results <- list()
for (mode in starsolo_modes) {
  st <- ruv_state_by_mode[[mode]]
  for (cl in clusters_to_sweep) {
    group_c <- factor(col_meta$cluster_num == cl, levels = c(FALSE, TRUE))
    if (sum(group_c == "TRUE") < 2 || sum(group_c == "FALSE") < 2) next
    design_c <- model.matrix(~ sample_block + group_c)
    if (qr(design_c)$rank < ncol(design_c)) next
    if (!"group_cTRUE" %in% colnames(design_c)) next
    for (fs in repeat_fsets) {
      key <- paste(mode, fs, sep = "_")
      counts <- pb_list[[key]]
      stopifnot(identical(colnames(counts), st$pb_samples))
      storage.mode(counts) <- "integer"
      tt <- tryCatch(
        run_gene_lib_repeat_de(repeat_counts = counts,
                               gene_counts = st$genes_expr,
                               design = design_c,
                               test = list(coef = "group_cTRUE")),
        error = function(e) { cat("skip cluster", cl, mode, fs, ":",
                                  conditionMessage(e), "\n"); NULL })
      if (is.null(tt)) next
      tag <- paste(key, sprintf("cluster%02d_vs_other", cl), sep = ":")
      per_cluster_results[[tag]] <- tt
    }
  }
}
cat("per-cluster DE tables:", length(per_cluster_results), "\n")
save_rds(per_cluster_results, "per_cluster_gene_lib_repeat_de")
```

## DE stratified by biology (polyA / age / strand / A-run)

```{r biology-de-strat, fig.width=14, fig.height=10}
if (!is.null(biology_annotation) && length(all_results) > 0) {
  strat_all <- bind_rows(lapply(names(all_results), function(key) {
    de_stratify_counts(all_results[[key]], key)
  }))
  save_rds(strat_all, "ruv_de_biology_stratification")
  for (facet_nm in unique(strat_all$facet)) {
    print(plot_de_stratification(filter(strat_all, facet == facet_nm),
                                 paste("cluster 12 vs other, stratified by",
                                       facet_nm)))
  }
}
```

## Volcano plots per (mode, featureset)

```{r ruv-volcano-plots, fig.width=7, fig.height=5}
for (tag in names(all_results)) {
  print(ruv_volcano(all_results[[tag]], tag, k_selected))
}
```

## Top features per (mode, featureset)

```{r ruv-top-tables, results='asis'}
for (tag in names(all_results)) {
  cat("\n###", tag, "\n\n")
  ruv_top_table(all_results[[tag]])
}
```

```{r de-long-csv}
long_rows <- list()
for (tag in names(all_results)) {
  r <- all_results[[tag]]
  parts <- strsplit(tag, ":", fixed = TRUE)[[1]]
  r$tag <- tag
  r$mode <- sub("_.*$", "", parts[1])
  r$featureset <- sub("^[^_]+_", "", parts[1])
  r$contrast <- parts[2]
  r$k <- k_selected
  long_rows[[tag]] <- r
}
save_csv(bind_rows(long_rows), "ruv_de_long")
```

# Summary

```{r summary}
summary_rows <- do.call(rbind, lapply(names(all_results), function(tag) {
  r <- all_results[[tag]]
  data.frame(
    tag = tag,
    k = k_selected,
    n_features = nrow(r),
    n_fdr_05 = sum(r$FDR < 0.05, na.rm = TRUE),
    n_fdr_10 = sum(r$FDR < 0.10, na.rm = TRUE),
    top_feature = r$feature_id[1],
    top_logfc = round(r$logFC[1], 3),
    top_fdr = signif(r$FDR[1], 3)
  )
}))
knitr::kable(summary_rows)
write.table(summary_rows, file.path(report_outdir, "ruv_summary.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
save_csv(summary_rows, "ruv_summary")
```
