---
title: "Polymenidou lab TDP-43 OE cultures (GSE230647, scRNA-seq): RUVg-normalized edgeR DE on pseudo-bulk"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  starsolo_base: "."
  metadata: "."
  sample_metadata: "."
  report_outdir: "."
  feature_sets: "genes genic_repeats intergenic_repeats"
  starsolo_modes: "unique"
  n_controls: 3000
  k_selected: 1
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(Matrix)
  library(edgeR)
  library(RUVSeq)
  library(matrixStats)
  library(readr)
  library(dplyr)
  library(tidyr)
})

source("../workflow/scripts/ruv_common.R")

starsolo_base <- params$starsolo_base
report_outdir <- params$report_outdir
n_controls <- as.integer(params$n_controls)
k_selected <- as.integer(params$k_selected)
feature_sets <- strsplit(params$feature_sets, " ")[[1]]
starsolo_modes <- strsplit(params$starsolo_modes, " ")[[1]]
repeat_fsets <- feature_sets[feature_sets != "genes"]

stopifnot(dir.exists(starsolo_base), dir.exists(report_outdir),
          n_controls >= 500, k_selected >= 1, k_selected <= 2,
          "genes" %in% feature_sets, length(repeat_fsets) >= 1,
          length(starsolo_modes) >= 1)

save_rds <- function(obj, name) {
  saveRDS(obj, file.path(report_outdir, paste0(name, ".rds")))
}

csv_dir <- file.path(report_outdir, "csv")
dir.create(csv_dir, showWarnings = FALSE, recursive = TRUE)
save_csv <- function(df, name) {
  path <- file.path(csv_dir, paste0(name, ".csv"))
  write.csv(df, path, row.names = FALSE)
  cat("wrote", path, "(", nrow(df), "rows)\n")
  invisible(path)
}
```

Paper counterpart of the sc method report's RUVg block. This script owns its own RUVg analysis end-to-end, parallel to the two bulk paper Rmds. Data: GSE230647 (Polymenidou lab), three DOX-on HA-TDP-43 iPSC-derived neural cultures (2-week and 4-week time points). Cluster 12 in the GEO cell metadata flags the TDP-43-HA-expressing neurons and acts as the neuron-resolved counterpart of the bulk `tdp43_oe vs oe_ctrl` contrast. Inputs are STARsolo raw matrices from `results/gse230647_sc/starsolo/` (produced by the method pipeline with `workflow/configs/gse230647_sc.yaml`) plus the GEO cell metadata file. No counts are re-derived from raw reads.

# Sample metadata

```{r sample-meta}
sample_meta <- read_tsv(params$sample_metadata, show_col_types = FALSE)
sample_ids <- sample_meta$sample_id
stopifnot(length(sample_ids) == 3, !anyDuplicated(sample_ids))
knitr::kable(sample_meta)
```

# Cell-type annotations from GEO

```{r load-geo-meta}
geo_meta <- read_tsv(params$metadata, show_col_types = FALSE)
stopifnot("biorxiv_v2_cluster_number" %in% colnames(geo_meta),
          "biorxiv_v2_cluster_name" %in% colnames(geo_meta))

barcode_col <- grep("barcode|cell|cb", colnames(geo_meta),
                    ignore.case = TRUE, value = TRUE)[1]
stopifnot(!is.na(barcode_col))

geo_meta$cluster_num <- as.integer(geo_meta$biorxiv_v2_cluster_number)
geo_meta$cluster <- geo_meta$biorxiv_v2_cluster_name
geo_meta$barcode_16bp <- sub("-[0-9]+\\..*$", "", geo_meta[[barcode_col]])
geo_meta$sample_tag <- sub("^[^.]*\\.", "", geo_meta[[barcode_col]])

rep_suffix <- ifelse(is.na(sample_meta$replicate) | sample_meta$replicate == "",
                     "", tolower(sample_meta$replicate))
tag_to_sid <- setNames(sample_meta$sample_id,
                       paste0("TDP", sample_meta$culture_weeks, "wON", rep_suffix))
geo_meta$sample_id <- unname(tag_to_sid[geo_meta$sample_tag])

geo_meta$group <- ifelse(!is.na(geo_meta$cluster_num) & geo_meta$cluster_num == 12L,
                         "cluster12", "other")

cat("cluster 12 cells:", sum(geo_meta$group == "cluster12", na.rm = TRUE), "\n")
cat("other cells:", sum(geo_meta$group == "other", na.rm = TRUE), "\n")
save_rds(geo_meta, "geo_meta_annotated")
```

```{r cluster-sizes}
cluster_sizes <- geo_meta %>%
  filter(!is.na(sample_id), !is.na(cluster_num)) %>%
  count(sample_id, cluster_num, cluster, name = "n_cells") %>%
  arrange(sample_id, cluster_num)
save_csv(cluster_sizes, "cluster_sizes_per_sample")

group_sizes <- geo_meta %>%
  filter(!is.na(sample_id), !is.na(cluster_num)) %>%
  count(sample_id, group, name = "n_cells") %>%
  pivot_wider(names_from = group, values_from = n_cells, values_fill = 0)
save_csv(group_sizes, "group_sizes_per_sample")
```

# Pseudo-bulk from STARsolo raw matrices

```{r load-starsolo}
load_and_split <- function(raw_dir, geo_sample) {
  stopifnot(dir.exists(raw_dir))
  barcodes <- readLines(file.path(raw_dir, "barcodes.tsv"))
  features <- read_tsv(file.path(raw_dir, "features.tsv"),
                       col_names = c("id", "name", "type"),
                       show_col_types = FALSE)
  mat <- readMM(file.path(raw_dir, "matrix.mtx"))
  rownames(mat) <- features$id
  colnames(mat) <- barcodes

  matched_idx <- match(barcodes, geo_sample$barcode_16bp)
  matched <- !is.na(matched_idx)
  group_matched <- geo_sample$group[matched_idx]
  idx12 <- matched & !is.na(group_matched) & group_matched == "cluster12"
  idx_other <- matched & !is.na(group_matched) & group_matched == "other"

  list(
    cluster12 = Matrix::rowSums(mat[, idx12, drop = FALSE]),
    other     = Matrix::rowSums(mat[, idx_other, drop = FALSE]),
    n12       = sum(idx12),
    n_other   = sum(idx_other)
  )
}

pb_list <- list()
for (mode in starsolo_modes) {
  for (fs in feature_sets) {
    key <- paste(mode, fs, sep = "_")
    counts12 <- list()
    counts_other <- list()
    for (sid in sample_ids) {
      raw_dir <- file.path(starsolo_base, sid,
                           paste0(mode, "_", fs),
                           "Solo.out", "Gene", "raw")
      geo_sample <- geo_meta[!is.na(geo_meta$sample_id) &
                             geo_meta$sample_id == sid, ]
      stopifnot(nrow(geo_sample) > 0)
      res <- load_and_split(raw_dir, geo_sample)
      counts12[[sid]] <- res$cluster12
      counts_other[[sid]] <- res$other
      cat(key, sid, "| cluster12:", res$n12, "| other:", res$n_other, "\n")
    }
    features_union <- Reduce(union, lapply(counts12, names))
    mat12 <- sapply(counts12, function(x) x[features_union])
    mat_other <- sapply(counts_other, function(x) x[features_union])
    rownames(mat12) <- features_union
    rownames(mat_other) <- features_union
    mat12[is.na(mat12)] <- 0
    mat_other[is.na(mat_other)] <- 0
    pb_list[[key]] <- list(cluster12 = mat12, other = mat_other)
  }
}
save_rds(pb_list, "pseudobulk_counts")
```

```{r pseudobulk-libsizes}
lib_rows <- list()
for (mode in starsolo_modes) {
  gene_key <- paste(mode, "genes", sep = "_")
  stopifnot(gene_key %in% names(pb_list))
  gene_pb <- pb_list[[gene_key]]
  lib_rows[[paste(mode, "cluster12")]] <- data.frame(
    mode = mode, group = "cluster12",
    sample_id = names(colSums(gene_pb$cluster12)),
    gene_lib_size = unname(colSums(gene_pb$cluster12)))
  lib_rows[[paste(mode, "other")]] <- data.frame(
    mode = mode, group = "other",
    sample_id = names(colSums(gene_pb$other)),
    gene_lib_size = unname(colSums(gene_pb$other)))
}
save_csv(bind_rows(lib_rows), "pseudobulk_gene_lib_sizes")
```

# Empirical control selection

Per-mode naive edgeR QL F-test on the pseudo-bulk gene matrix. Design is `~ sample_block + group`; the top `n_controls` genes with the largest p-values become empirical controls.

```{r rank-controls}
ruv_state_by_mode <- list()
naive_de_by_mode <- list()
for (mode in starsolo_modes) {
  gene_key <- paste(mode, "genes", sep = "_")
  gene_pb <- pb_list[[gene_key]]
  sample_cols <- colnames(gene_pb$cluster12)
  pb_samples <- c(paste0(sample_cols, "__c12"),
                  paste0(sample_cols, "__other"))
  group <- factor(c(rep("cluster12", length(sample_cols)),
                    rep("other",     length(sample_cols))),
                  levels = c("other", "cluster12"))
  sample_block <- factor(rep(sample_cols, 2))

  genes_mat <- cbind(gene_pb$cluster12, gene_pb$other)
  colnames(genes_mat) <- pb_samples
  storage.mode(genes_mat) <- "integer"
  keep <- rowSums(genes_mat >= 10) >= floor(ncol(genes_mat) / 2)
  genes_expr <- genes_mat[keep, , drop = FALSE]
  cat(mode, ": filtered gene matrix", nrow(genes_expr), "x", ncol(genes_expr), "\n")

  n_ctrl <- min(n_controls, nrow(genes_expr) - 10)
  stopifnot(n_ctrl >= 500)

  design0 <- model.matrix(~ sample_block + group)
  stopifnot(qr(design0)$rank == ncol(design0))
  ctrl_res <- rank_empirical_controls(
    counts = genes_expr, design = design0,
    test = list(coef = "groupcluster12"), n_controls = n_ctrl
  )
  naive_de_by_mode[[mode]] <- ctrl_res$naive_de
  writeLines(ctrl_res$controls,
             file.path(report_outdir, paste0("ruv_empirical_controls_", mode, ".txt")))

  ruv_state_by_mode[[mode]] <- list(
    pb_samples = pb_samples, group = group, sample_block = sample_block,
    genes_expr = genes_expr, controls = ctrl_res$controls, n_ctrl = n_ctrl
  )
}
save_rds(naive_de_by_mode, "ruv_naive_gene_de")
```

# RUVg factor estimation

k runs from 1 to 2 (capped by residual df of the pseudo-bulk design). The operating k is fixed via `k_selected` (default 1); k = 2 is marginal because `~ sample_block + W + group` has k + 4 coefficients over 6 columns.

```{r ruvg}
for (mode in starsolo_modes) {
  st <- ruv_state_by_mode[[mode]]
  ruv_fit_res <- fit_ruv_range(
    genes_expr = st$genes_expr, controls = st$controls,
    pheno_df = data.frame(st$group, row.names = st$pb_samples),
    ks = 1:2
  )
  st$diag_rle <- ruv_fit_res$diag_rle
  st$fits <- ruv_fit_res$fits
  write.table(ruv_fit_res$diag_rle,
              file.path(report_outdir, paste0("ruv_rle_diagnostic_", mode, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  knitr::kable(ruv_fit_res$diag_rle, digits = 3,
               caption = paste("RLE diagnostic, mode =", mode)) |> print()

  W <- pick_W_matrix(ruv_fit_res$fits, k_selected, st$pb_samples)
  sample_block <- st$sample_block
  group <- st$group
  design <- model.matrix(~ sample_block + W + group)
  stopifnot(qr(design)$rank == ncol(design))
  resid_df <- ncol(st$genes_expr) - qr(design)$rank
  cat(mode, "| design rank", qr(design)$rank,
      "| residual df", resid_df, "\n")
  st$W <- W
  st$design <- design
  st$resid_df <- resid_df
  ruv_state_by_mode[[mode]] <- st
}
save_rds(ruv_state_by_mode, "ruv_state")
```

```{r w-factors-csv}
w_rows <- list()
rle_rows <- list()
for (mode in names(ruv_state_by_mode)) {
  st <- ruv_state_by_mode[[mode]]
  wdf <- as.data.frame(st$W)
  wdf$sample_column <- rownames(st$W)
  wdf$mode <- mode
  wdf$group <- as.character(st$group)
  wdf$sample_block <- as.character(st$sample_block)
  w_rows[[mode]] <- wdf
  rle_df <- st$diag_rle
  rle_df$mode <- mode
  rle_rows[[mode]] <- rle_df
}
save_csv(bind_rows(w_rows), "ruv_W_factors")
save_csv(bind_rows(rle_rows), "ruv_rle_diagnostic")
```

# Repeat DE with RUV covariates

Per (mode, featureset), contrast `groupcluster12` on the pseudo-bulk repeat counts under `~ sample_block + W + group`.

```{r repeat-de}
all_results <- list()
for (mode in starsolo_modes) {
  st <- ruv_state_by_mode[[mode]]
  for (fs in repeat_fsets) {
    key <- paste(mode, fs, sep = "_")
    pb <- pb_list[[key]]
    counts <- cbind(pb$cluster12, pb$other)
    colnames(counts) <- st$pb_samples
    storage.mode(counts) <- "integer"
    stopifnot("groupcluster12" %in% colnames(st$design))
    tt <- run_ruv_repeat_de(counts = counts, design = st$design,
                            test = list(coef = "groupcluster12"))
    tag <- paste(key, "cluster12_vs_other", sep = ":")
    out_tsv <- file.path(report_outdir,
                         sprintf("ruv_k%d_%s_%s_cluster12_vs_other.tsv",
                                 k_selected, mode, fs))
    write.table(tt, out_tsv, sep = "\t", quote = FALSE, row.names = FALSE)
    all_results[[tag]] <- tt
  }
}
stopifnot(length(all_results) == length(starsolo_modes) * length(repeat_fsets))
save_rds(all_results, "ruv_repeat_de")
```

```{r de-long-csv}
long_rows <- list()
for (tag in names(all_results)) {
  r <- all_results[[tag]]
  parts <- strsplit(tag, ":", fixed = TRUE)[[1]]
  r$tag <- tag
  r$mode <- sub("_.*$", "", parts[1])
  r$featureset <- sub("^[^_]+_", "", parts[1])
  r$contrast <- parts[2]
  r$k <- k_selected
  long_rows[[tag]] <- r
}
save_csv(bind_rows(long_rows), "ruv_de_long")
```

# Summary

```{r summary}
summary_rows <- do.call(rbind, lapply(names(all_results), function(tag) {
  r <- all_results[[tag]]
  data.frame(
    tag = tag,
    k = k_selected,
    n_features = nrow(r),
    n_fdr_05 = sum(r$FDR < 0.05, na.rm = TRUE),
    n_fdr_10 = sum(r$FDR < 0.10, na.rm = TRUE),
    top_feature = r$feature_id[1],
    top_logfc = round(r$logFC[1], 3),
    top_fdr = signif(r$FDR[1], 3)
  )
}))
knitr::kable(summary_rows)
write.table(summary_rows, file.path(report_outdir, "ruv_summary.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
save_csv(summary_rows, "ruv_summary")
```
