---
title: "repeat element quantification - evaluation report"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  eval_dir: "."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 12)
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(scales)
library(patchwork)

eval_dir <- params$eval_dir

# consistent colour palette that scales to many categories
make_palette <- function(n) {
  if (n <= 8)  return(RColorBrewer::brewer.pal(max(3, n), "Dark2")[seq_len(n)])
  colorRampPalette(RColorBrewer::brewer.pal(8, "Dark2"))(n)
}
```

```{r load-data}
# -- global metrics --
summary_path <- file.path(eval_dir, "summary_global_metrics.tsv")
global <- read_tsv(summary_path, show_col_types = FALSE)

num_cols <- c("pearson_r", "spearman_r", "log1p_rmse",
              "precision", "recall", "f1", "jaccard", "specificity",
              "n_cells", "n_features",
              "wall_time_s", "cpu_time_s", "max_rss_mb", "io_in_mb", "io_out_mb")
for (col in num_cols) {
  if (col %in% names(global))
    global[[col]] <- suppressWarnings(as.numeric(global[[col]]))
}

# ensure factor columns exist
if (!"granularity" %in% names(global))
  global$granularity <- "gene_id"
if (!"feature_set" %in% names(global))
  global$feature_set <- "repeats"
if (!"multimapper_mode" %in% names(global))
  global$multimapper_mode <- "unique"

global <- global %>%
  mutate(
    aligner_mode = paste0(aligner, " (", multimapper_mode, ")"),
    gran_label   = factor(granularity,
                          levels = c("locus", "gene_id", "family_id", "class_id"))
  )

# -- per-cell metrics --
load_tsv_files <- function(pattern) {
  files <- list.files(eval_dir, pattern = pattern, full.names = TRUE)
  if (length(files) == 0) return(NULL)
  purrr::map_dfr(files, function(f) {
    df <- read_tsv(f, show_col_types = FALSE)
    df
  })
}

per_cell   <- tryCatch(load_tsv_files("_per_cell_metrics\\.tsv$"),   error = function(e) NULL)
per_family <- tryCatch(load_tsv_files("_per_family_metrics\\.tsv$"), error = function(e) NULL)

if (!is.null(per_family) && nrow(per_family) > 0) {
  for (col in c("pearson_r", "spearman_r", "f1", "precision", "recall",
                "jaccard", "specificity")) {
    if (col %in% names(per_family))
      per_family[[col]] <- suppressWarnings(as.numeric(per_family[[col]]))
  }
  if (!"multimapper_mode" %in% names(per_family))
    per_family$multimapper_mode <- "unique"
  per_family <- per_family %>%
    mutate(aligner_mode = paste0(aligner, " (", multimapper_mode, ")"))
}
```

## global accuracy metrics

All metrics compare the aligner count matrix against the simulation ground truth at the same aggregation level (granularity).

- **Pearson r / Spearman r**: correlation of observed vs ground-truth counts across all (feature, cell) pairs.
- **log1p RMSE**: root-mean-square error of log1p-transformed counts; lower is better.
- **Precision**: fraction of (feature, cell) pairs called expressed by the aligner that are truly expressed in the ground truth.
- **Recall**: fraction of truly expressed (feature, cell) pairs detected by the aligner.
- **F1**: harmonic mean of precision and recall.
- **Jaccard**: intersection over union of expressed (feature, cell) pairs; penalises both false positives and false negatives.
- **Specificity**: fraction of truly unexpressed (feature, cell) pairs correctly assigned zero by the aligner.


Rows = metric, columns = feature_set. Each point is one aligner at one granularity;
x-axis = granularity, colour = aligner (+ multimapper mode).

```{r global-accuracy, fig.height=18, fig.width=14}
acc_metrics <- c("pearson_r", "spearman_r", "f1",
                 "precision", "recall", "jaccard", "specificity")

acc_long <- global %>%
  select(aligner_mode, feature_set, gran_label, all_of(acc_metrics)) %>%
  pivot_longer(all_of(acc_metrics), names_to = "metric", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(metric = factor(metric, levels = acc_metrics))

n_al <- length(unique(acc_long$aligner_mode))
pal  <- make_palette(n_al)

ggplot(acc_long, aes(x = gran_label, y = value,
                     colour = aligner_mode, group = aligner_mode)) +
  geom_point(size = 2.5, alpha = 0.85) +
  geom_line(linetype = "dashed", linewidth = 0.5) +
  facet_grid(metric ~ feature_set, scales = "free_y") +
  scale_colour_manual(values = pal, name = "aligner (mode)") +
  labs(x = "granularity", y = NULL,
       title = "global accuracy metrics",
       subtitle = "rows = metric, columns = feature_set; x = granularity; colour = aligner") +
  theme_bw(base_size = 10) +
  theme(aspect.ratio = 1,
        axis.text.x  = element_text(angle = 40, hjust = 1, size = 8),
        strip.text.y = element_text(size = 8),
        strip.text.x = element_text(size = 8),
        legend.position = "bottom",
        legend.text = element_text(size = 8))
```

## precision-recall

```{r pr-plot, fig.height=6, fig.width=10}
det <- global %>% filter(!is.na(recall), !is.na(precision))

if (nrow(det) > 0) {
  n_lab <- length(unique(det$aligner_mode))
  pal_pr <- make_palette(n_lab)

  ggplot(det, aes(x = recall, y = precision,
                  colour = aligner_mode, shape = feature_set)) +
    geom_point(alpha = 0.85) +
    scale_colour_manual(values = pal_pr, name = "Aligner (mode)") +
    scale_shape_manual(values = c(16, 17, 15, 3, 7, 8)[seq_len(
                         length(unique(det$feature_set)))],
                       name = "Feature set") +
    facet_wrap(~gran_label, nrow = 1) +
    xlim(0, 1) + ylim(0, 1) +
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", colour = "grey60") +
    labs(title = "precision vs recall",
         subtitle = "facets = granularity",
         x = "Recall", y = "Precision") +
    theme_bw(base_size = 11) +
    theme(aspect.ratio = 1) +
    guides(colour = guide_legend(override.aes = list(size = 4)))
} else {
  cat("No detection metrics available.\n")
}
```

## correlation metrics

```{r correlations, fig.height=5}
corr <- global %>%
  select(aligner_mode, feature_set, gran_label, pearson_r, spearman_r) %>%
  pivot_longer(c(pearson_r, spearman_r), names_to = "metric", values_to = "r") %>%
  filter(!is.na(r))

n_al <- length(unique(corr$aligner_mode))
pal_corr <- make_palette(n_al)

ggplot(corr, aes(x = gran_label, y = r,
                  colour = aligner_mode, group = aligner_mode)) +
  geom_line(linetype = "dashed", linewidth = 0.6) +
  geom_point(size = 3) +
  facet_grid(metric ~ feature_set) +
  scale_colour_manual(values = pal_corr, name = "Aligner (mode)") +
  ylim(0, 1) +
  labs(x = "granularity", y = "correlation coefficient",
       title = "Pearson and Spearman correlations",
       subtitle = "rows = metric, columns = feature_set") +
  theme_bw(base_size = 11) +
  theme(aspect.ratio = 1,
        axis.text.x = element_text(angle = 25, hjust = 1))
```

## log1p RMSE

log1p RMSE is computed over all (feature, cell) pairs in the feature universe.
At locus level there are thousands of features, most with zero counts in both
truth and observed; the dominance of (0, 0) pairs suppresses the average squared
error so RMSE looks low. At class_id level there are only ~10-20 features, each
accumulating large aggregated counts, so per-entry log1p differences are larger
and RMSE rises. The gradient locus < gene_id < family_id < class_id is therefore
expected and not a mistake - lower locus RMSE reflects sparsity, not better
quantification accuracy.

```{r rmse, fig.height=5}
if ("log1p_rmse" %in% names(global) && any(!is.na(global$log1p_rmse))) {
  n_al2 <- length(unique(global$aligner_mode))
  pal_rmse <- make_palette(n_al2)

  ggplot(global, aes(x = gran_label, y = log1p_rmse,
                      colour = aligner_mode, group = aligner_mode)) +
    geom_line(linetype = "dashed", linewidth = 0.6) +
    geom_point(size = 3) +
    facet_wrap(~feature_set, nrow = 1) +
    scale_colour_manual(values = pal_rmse, name = "Aligner (mode)") +
    labs(x = "granularity", y = "log1p RMSE",
         title = "Count quantification error (lower = better)") +
    theme_bw(base_size = 11) +
    theme(aspect.ratio = 1,
          axis.text.x = element_text(angle = 25, hjust = 1))
} else {
  cat("log1p_rmse not available.\n")
}
```

## per-cell correlations

```{r per-cell, fig.height=7, fig.width=12}
if (!is.null(per_cell) && nrow(per_cell) > 0) {
  per_cell_num <- per_cell %>%
    mutate(across(c(pearson_r, spearman_r), ~ suppressWarnings(as.numeric(.)))) %>%
    filter(!is.na(aligner))

  if (!"multimapper_mode" %in% names(per_cell_num))
    per_cell_num$multimapper_mode <- "unique"
  if (!"feature_set" %in% names(per_cell_num))
    per_cell_num$feature_set <- "repeats"

  per_cell_num <- per_cell_num %>%
    mutate(aligner_mode = paste0(aligner, " (", multimapper_mode, ")"),
           gran_label = factor(granularity,
                               levels = c("locus", "gene_id", "family_id", "class_id")))

  pc_long <- per_cell_num %>%
    pivot_longer(c(pearson_r, spearman_r), names_to = "metric", values_to = "r") %>%
    filter(!is.na(r))

  n_al3 <- length(unique(pc_long$aligner_mode))
  pal_pc <- make_palette(n_al3)

  ggplot(pc_long, aes(x = aligner_mode, y = r, fill = aligner_mode)) +
    geom_violin(alpha = 0.7, draw_quantiles = c(0.25, 0.5, 0.75)) +
    geom_jitter(width = 0.1, size = 0.8, alpha = 0.5) +
    facet_grid(metric ~ feature_set + gran_label) +
    scale_fill_manual(values = pal_pc) +
    labs(x = NULL, y = "correlation",
         title = "Per-cell correlations",
         subtitle = "rows = metric, columns = feature_set x granularity") +
    theme_bw(base_size = 9) +
    theme(aspect.ratio = 1,
          axis.text.x  = element_text(angle = 40, hjust = 1, size = 7),
          legend.position = "none",
          strip.text = element_text(size = 7))
} else {
  cat("No per-cell metrics found.\n")
}
```

## per-repeat-class metrics

```{r per-family, fig.height=14, fig.width=10}
if (!is.null(per_family) && nrow(per_family) > 0 && "class_id" %in% names(per_family)) {
  pf <- per_family %>%
    filter(!is.na(class_id), class_id != "unknown") %>%
    mutate(gran_label = factor(granularity,
                               levels = c("locus", "gene_id", "family_id", "class_id")))

  n_al4 <- length(unique(pf$aligner_mode))
  pal_pf <- make_palette(n_al4)

  p_f1 <- ggplot(pf, aes(x = class_id, y = f1,
                           colour = aligner_mode, group = aligner_mode)) +
    geom_line(linewidth = 0.5, linetype = "dashed") +
    geom_point(size = 2, alpha = 0.85) +
    facet_grid(gran_label ~ feature_set) +
    scale_y_sqrt() +
    scale_colour_manual(values = pal_pf, name = "Aligner (mode)") +
    labs(x = "repeat class", y = "F1",
         title = "F1 per repeat class",
         subtitle = "rows = granularity, columns = feature_set") +
    theme_bw(base_size = 9) +
    theme(aspect.ratio = 1,
          axis.text.x = element_text(angle = 40, hjust = 1, size = 7),
          legend.position = "bottom",
          legend.text = element_text(size = 8))

  p_r <- ggplot(pf, aes(x = class_id, y = pearson_r,
                          colour = aligner_mode, group = aligner_mode)) +
    geom_line(linewidth = 0.5, linetype = "dashed") +
    geom_point(size = 2, alpha = 0.85) +
    facet_grid(gran_label ~ feature_set) +
    scale_y_sqrt() +
    scale_colour_manual(values = pal_pf, name = "Aligner (mode)") +
    labs(x = "repeat class", y = "pearson r",
         title = "Pearson r per repeat class") +
    theme_bw(base_size = 9) +
    theme(aspect.ratio = 1,
          axis.text.x = element_text(angle = 40, hjust = 1, size = 7),
          legend.position = "none",
          strip.text = element_text(size = 7))

  p_f1 / p_r
} else {
  cat("No per-family metrics with class_id found.\n")
}
```

## compute resources

```{r resources, fig.height=10, fig.width=14}
res_cols <- intersect(c("wall_time_s", "cpu_time_s", "max_rss_mb", "io_in_mb", "io_out_mb"), names(global))
if (length(res_cols) > 0 && any(!is.na(global[res_cols]))) {
  n_al5 <- length(unique(global$aligner_mode))
  pal_res <- make_palette(n_al5)

  res <- global %>%
    select(aligner_mode, feature_set, gran_label, all_of(res_cols)) %>%
    pivot_longer(all_of(res_cols), names_to = "resource", values_to = "value") %>%
    filter(!is.na(value))

  ggplot(res, aes(x = gran_label, y = value,
                   colour = aligner_mode, group = aligner_mode)) +
    geom_line(linetype = "dashed", linewidth = 0.6) +
    geom_point(size = 2) +
    facet_grid(feature_set ~ resource, scales = "free_y") +
    scale_colour_manual(values = pal_res, name = "Aligner (mode)") +
    labs(x = "granularity", y = NULL, title = "compute resource usage") +
    theme_bw(base_size = 10) +
    theme(aspect.ratio = 1,
          axis.text.x = element_text(angle = 25, hjust = 1))
} else {
  cat("Benchmark data not available.\n")
}
```

---
*Generated by [evaluation_report.Rmd](../workflow/scripts/evaluation_report.Rmd)*
