---
title: "GSE126543 bulk RNA-seq repeat quantification (gene-normalized)"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  counts_dir: "."
  report_outdir: "."
  indices_base: "."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(edgeR)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(tidyr)
library(readr)
library(patchwork)
library(pheatmap)
source(file.path("scripts", "ggtheme.R"))

counts_dir = params$counts_dir
report_outdir = params$report_outdir
indices_base = params$indices_base

aligners = c("star", "kallisto", "salmon")
granularities = c("gene_id", "family_id")
feature_sets = c("repeats", "genic_repeats", "intergenic_repeats")

donors = paste0("s", 1:7)
main_samples = c(paste0("tdp43pos_", donors), paste0("tdp43neg_", donors))
unsorted_samples = c("unsorted_s3", "unsorted_s4")

main_meta = data.frame(
  sample = main_samples,
  condition = rep(c("tdp43pos", "tdp43neg"), each = 7),
  donor = rep(donors, 2),
  row.names = main_samples,
  stringsAsFactors = FALSE
)

save_rds = function(obj, name) {
  path = file.path(report_outdir, paste0(name, ".rds"))
  saveRDS(obj, path)
  invisible(path)
}
```

## Biology annotation for detection and DE stratification

```{r biology-setup}
source(file.path("scripts", "biology_checks.R"))

biology_annotation_path = file.path(indices_base, "repeat_biology_annotation.tsv")
biology_annotation = load_biology_annotation(biology_annotation_path)
if (is.null(biology_annotation)) {
  cat("No biology annotation at", biology_annotation_path,
      "- biology-stratified panels will be skipped.\n")
} else {
  cat("Loaded biology annotation:", nrow(biology_annotation), "subfamilies\n")
}
save_rds(biology_annotation, "biology_annotation")

annotate_de_with_biology = function(tt) {
  tt = annotate_biology(tt, biology_annotation, key = "gene_id")
  tt$polya = polya_label(tt$class_id)
  tt$age = family_age_label(tt$family_age)
  tt$strand = strand_vs_gene_label(tt$strand_vs_gene)
  tt$a_run = a_run_label(tt$frac_downstream_a_run)
  tt
}

de_stratify_counts = function(tt, key, fdr = 0.05) {
  tt = annotate_de_with_biology(tt)
  tt$direction = ifelse(is.na(tt$FDR) | tt$FDR >= fdr, "ns",
                        ifelse(tt$logFC > 0, "up", "down"))
  facet_cols = c(polya = "polya", family_age = "age",
                 strand_vs_gene = "strand", downstream_a_run = "a_run")
  rows = list()
  for (facet_nm in names(facet_cols)) {
    col = facet_cols[[facet_nm]]
    rows[[facet_nm]] = tt %>%
      group_by(label = .data[[col]], direction) %>%
      summarize(n = dplyr::n(), .groups = "drop") %>%
      mutate(facet = facet_nm, key = key)
  }
  bind_rows(rows)
}

plot_de_stratification = function(strat_df, title) {
  if (is.null(strat_df) || nrow(strat_df) == 0) return(invisible(NULL))
  strat_df$direction = factor(strat_df$direction, levels = c("up", "down", "ns"))
  ggplot(strat_df, aes(x = label, y = n, fill = direction)) +
    geom_col() +
    facet_grid(facet ~ key, scales = "free", space = "free_x") +
    scale_fill_manual(values = c(up = "firebrick", down = "steelblue", ns = "grey70")) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = title, x = NULL, y = "features")
}
```

## About this report

```{r provenance}
git_commit = tryCatch(
  system("git rev-parse --short HEAD", intern = TRUE),
  error = function(e) "unknown"
)
cat("GEO accession: GSE126543\n")
cat("SRA study: SRP185789\n")
cat("Rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
cat("Git commit:", git_commit, "\n")
cat("counts_dir:", counts_dir, "\n")
cat("report_outdir:", report_outdir, "\n")
```

GSE126543 (SRP185789): NeuN-positive neuronal nuclei from human frontal cortex
(FTD/ALS post-mortem tissue), sorted by TDP-43 nuclear presence across 7 donors.

- tdp43pos: TDP-43 retained in nucleus, n=7 donors
- tdp43neg: TDP-43 lost from nucleus, n=7 donors
- unsorted: mixed TDP-43 status, donors S3 and S4 only (shown in PCA for context)

Repeat counts are normalized by gene library sizes (total gene read counts per
sample) so that fold changes reflect changes in the repeat-to-gene expression
ratio. Donor-matched design; edgeR QLF with donor as blocking factor
(~donor + condition).

## PolyA-competence of detected repeats

Detection and differential-expression results can be split into two groups
based on the intrinsic ability of each repeat class to produce a
polyadenylated transcript. This split uses only the `class_id` column from
`{feature_set}_locus_map.tsv` (equivalent to RepeatMasker `repClass`); no
sequence-level polyA scanning is performed.

- with polyA (polyA-competent): `LINE`, `SINE`, `LTR`, `Retroposon`.
  These classes produce polyadenylated transcripts when expressed from their
  own promoters (LINE-1 and ERV internal transcripts carry canonical AATAAA
  poly(A) sites, SINE/Alu relies on LINE-1 machinery and poly(A) tails,
  Retroposon/SVA also polyadenylates).
- without polyA (not polyA-competent): every other class, including
  `Satellite`, `Simple_repeat`, `Low_complexity`, `DNA`, `RC`, and the
  structural-RNA-derived classes (`rRNA`, `tRNA`, `snRNA`, `scRNA`,
  `srpRNA`). These lack a canonical polyadenylation signal in their body.
- unknown: loci with no class annotation in the locus map.

The genic vs intergenic feature_set split is complementary: genic repeats can
be captured through a polyadenylated host transcript regardless of intrinsic
class, while intergenic repeats need their own poly(A) tail to be detected by
polyA-selection assays. The class-based facet therefore matters most on the
`intergenic_repeats` feature set.

## Biology-driven sanity checks for detection

Three additional commonsense checks help reason about why a given repeat is
or is not detected. Classification logic is shared across reports in
`workflow/scripts/biology_annotations.py` (Python, pure functions, unit
tested in `test/unit/test_biology_annotations.py`) and the R-side join and
labelling helpers live in `workflow/scripts/biology_checks.R`.

- Family age (from RepeatMasker subfamily name or milliDiv). Younger
  subfamilies are more often transcriptionally active, older ones are
  generally silenced. Brackets:

      young   L1HS, L1PA2-7, AluY, SVA_A-F, HERVK/HERVH/LTR5
      medium  L1PA8-16, AluS
      old     L1PA17+, L1PB, L1M*, AluJ, FLAM, FRAM
      unknown anything not matching, including DNA/RC/satellite classes

  When milliDiv is available the same brackets are produced from
  divergence thresholds (<=50 young, <=200 medium, >200 old).

- Strand concordance with overlapping genes. By construction
  `genic_repeats` is built with a same-strand intersect, so loci in that
  set are sense-with-host. The shared annotation labels every locus as
  `sense`, `antisense`, `both`, or `intergenic` against a strand-aware
  gene-body overlap so reports can also surface antisense loci that the
  current `intergenic_repeats` set silently absorbs.

- Internal A-tract proxy for polyA priming. Length of the longest run of
  As in the genomic region downstream of each locus (in mRNA-sense
  orientation). A long downstream A-run flags a repeat that may be
  detected by internal priming rather than by a genuine poly(A) tail,
  which matters most for intergenic non-polyA-competent classes.

The shared annotation is loaded from
`{indices_base}/indices/{genome_tag}/repeat_biology_annotation.tsv` when
present; otherwise the helpers no-op and reports fall back to the
class-only polyA facet.

## Load counts

```{r load-counts}
load_counts = function(aligner, feature_set, granularity) {
  path = file.path(counts_dir,
                   paste0(aligner, "_", feature_set, "_", granularity, "_unique.tsv"))
  if (!file.exists(path)) return(NULL)
  mat = read_tsv(path, show_col_types = FALSE)
  feat_col = colnames(mat)[1]
  feat_ids = mat[[feat_col]]
  mat = as.matrix(mat[, -1, drop = FALSE])
  rownames(mat) = feat_ids
  storage.mode(mat) = "integer"
  mat
}

counts_cache = list()
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      counts_cache[[key]] = load_counts(al, fs, gr)
    }
  }
}
cat("Available count matrices:", sum(!sapply(counts_cache, is.null)), "\n")
save_rds(counts_cache, "gse126543_counts_cache")
```

## Gene library sizes for normalization

Gene counts are loaded per aligner to compute gene library sizes. These replace
the default repeat-derived library sizes in edgeR so that repeat expression is
measured relative to gene expression depth.

```{r load-gene-lib-sizes}
gene_lib_cache = list()
for (al in aligners) {
  gmat = load_counts(al, "genes", "gene_id")
  if (!is.null(gmat)) gene_lib_cache[[al]] = colSums(gmat)
}
cat("Gene library sizes available for:", paste(names(gene_lib_cache), collapse = ", "), "\n")
save_rds(gene_lib_cache, "gse126543_gene_lib_sizes")
```

```{r edger-function}
run_edger_paired = function(counts, meta, gene_lib_sizes) {
  samples = intersect(meta$sample, colnames(counts))
  if (length(samples) < 6) return(NULL)
  mat = counts[, samples, drop = FALSE]
  meta = meta[samples, , drop = FALSE]
  cond = factor(meta$condition, levels = c("tdp43neg", "tdp43pos"))
  donor = factor(meta$donor)
  y = DGEList(counts = mat, group = cond)
  keep = filterByExpr(y)
  y = y[keep, ]
  y$samples$lib.size = gene_lib_sizes[colnames(y)]
  y$samples$norm.factors = rep(1, ncol(y))
  design = model.matrix(~donor + cond)
  y = estimateDisp(y, design)
  fit = glmQLFit(y, design)
  qlf = glmQLFTest(fit, coef = ncol(design))
  list(y = y, fit = fit, qlf = qlf,
       tt = topTags(qlf, n = Inf, sort.by = "PValue")$table)
}
```

## TDP-43 nuclear positive vs negative

edgeR QLF with donor as blocking factor.
Repeat counts normalized by gene library sizes.

```{r de-edger}
de_results = list()
for (al in aligners) {
  gene_ls = gene_lib_cache[[al]]
  if (is.null(gene_ls)) next
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      mat = counts_cache[[key]]
      if (is.null(mat)) next
      res = run_edger_paired(mat, main_meta, gene_ls)
      if (!is.null(res)) de_results[[key]] = res
    }
  }
}
save_rds(de_results, "gse126543_de_edger")
cat("DE edgeR fits completed:", length(de_results), "\n")
```

```{r pca, fig.width=10, fig.height=10}
pca_plots = list()
for (al in aligners) {
  for (fs in feature_sets) {
    key = paste(al, fs, "gene_id", sep = "_")
    fit = de_results[[key]]
    if (is.null(fit)) next
    mat = counts_cache[[key]]
    all_sids = intersect(c(main_meta$sample, unsorted_samples), colnames(mat))
    mat_sub = mat[rowSums(mat[, all_sids, drop = FALSE]) > 0, all_sids, drop = FALSE]
    gene_ls = gene_lib_cache[[al]]
    gene_ls_sub = gene_ls[all_sids]
    lcpm = log2(sweep(mat_sub, 2, gene_ls_sub / 1e6, "/") + 0.5)
    pca = prcomp(t(lcpm), scale. = FALSE)
    pct = round(100 * pca$sdev^2 / sum(pca$sdev^2))
    cond_map = setNames(main_meta$condition, main_meta$sample)
    df = data.frame(
      PC1 = pca$x[, 1],
      PC2 = pca$x[, 2],
      condition = ifelse(rownames(pca$x) %in% names(cond_map),
                         cond_map[rownames(pca$x)], "unsorted"),
      sample = rownames(pca$x)
    )
    pca_plots[[key]] = ggplot(df, aes(PC1, PC2, colour = condition, label = sample)) +
      geom_point(size = 2) +
      geom_text_repel(size = 2.5, show.legend = FALSE) +
      xlab(paste0("PC1: ", pct[1], "%")) +
      ylab(paste0("PC2: ", pct[2], "%")) +
      scale_colour_manual(values = cbbPalette) +
      theme_ng() +
      ggtitle(paste(al, fs, "(gene-normalized)"))
  }
}
if (length(pca_plots) > 0) wrap_plots(pca_plots, ncol = 3)
save_rds(pca_plots, "gse126543_pca_plots")
```

```{r volcano, fig.width=10, fig.height=10}
for (gr in granularities) {
  vol_plots = list()
  for (al in aligners) {
    for (fs in feature_sets) {
      key = paste(al, fs, gr, sep = "_")
      fit = de_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      tt$sig = tt$FDR < 0.05 & abs(tt$logFC) > 1
      vol_plots[[key]] = ggplot(tt, aes(logFC, -log10(PValue), colour = sig)) +
        geom_point(size = 0.6, alpha = 0.6) +
        scale_colour_manual(values = c("grey70", cbbPalette[2]),
                            labels = c("ns", "FDR<0.05 & |lfc|>1")) +
        geom_vline(xintercept = c(-1, 1), linetype = "dashed", linewidth = 0.3) +
        xlab("log2 FC (repeat/gene ratio)") +
        theme_ng() +
        ggtitle(paste(al, fs, gr))
    }
  }
  if (length(vol_plots) > 0) print(wrap_plots(vol_plots, ncol = 3))
}
```

## DE stratified by biology (polyA / age / strand / A-run)

```{r biology-de-strat-gse126543, fig.width=14, fig.height=10}
if (!is.null(biology_annotation) && length(de_results) > 0) {
  strat_all = bind_rows(lapply(names(de_results), function(key) {
    de_stratify_counts(de_results[[key]]$tt, key)
  }))
  save_rds(strat_all, "gse126543_de_biology_stratification")
  for (facet_nm in unique(strat_all$facet)) {
    print(plot_de_stratification(filter(strat_all, facet == facet_nm),
                                 paste("TDP43 pos vs neg, stratified by", facet_nm)))
  }
}
```

```{r ma, fig.width=10, fig.height=10}
for (gr in granularities) {
  ma_plots = list()
  for (al in aligners) {
    for (fs in feature_sets) {
      key = paste(al, fs, gr, sep = "_")
      fit = de_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      tt$sig = tt$FDR < 0.05
      ma_plots[[key]] = ggplot(tt, aes(logCPM, logFC, colour = sig)) +
        geom_point(size = 0.6, alpha = 0.6) +
        scale_colour_manual(values = c("grey70", cbbPalette[2]),
                            labels = c("ns", "FDR<0.05")) +
        geom_hline(yintercept = 0, linewidth = 0.3) +
        xlab("avg log2 gene-normalized CPM") +
        ylab("log2 FC (repeat/gene ratio)") +
        theme_ng() +
        ggtitle(paste(al, fs, gr))
    }
  }
  if (length(ma_plots) > 0) print(wrap_plots(ma_plots, ncol = 3))
}
```

```{r top-hits}
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = de_results[[key]]
      if (is.null(fit)) next
      top = fit$tt[fit$tt$FDR < 0.05, , drop = FALSE]
      if (nrow(top) == 0) next
      cat("\nTop hits -", al, fs, gr, "\n")
      print(knitr::kable(head(top[, c("logFC", "logCPM", "F", "PValue", "FDR")], 20),
                         digits = 4))
    }
  }
}
```

```{r heatmap, fig.width=10, fig.height=10}
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = de_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      sig_feats = rownames(tt)[tt$FDR < 0.05 & abs(tt$logFC) > 1]
      if (length(sig_feats) < 2) next
      sig_feats = head(sig_feats, 40)
      lcpm = cpm(fit$y, log = TRUE)[sig_feats, , drop = FALSE]
      sids = intersect(colnames(lcpm), main_meta$sample)
      ann = data.frame(
        condition = main_meta[sids, "condition"],
        donor = main_meta[sids, "donor"],
        row.names = sids
      )
      pheatmap(lcpm[, sids, drop = FALSE],
               annotation_col = ann,
               scale = "row",
               show_rownames = nrow(lcpm) <= 40,
               main = paste(al, fs, gr, "(gene-normalized)"),
               fontsize = 7,
               silent = TRUE)
    }
  }
}
```

```{r normalized-counts, fig.width=10, fig.height=10, results = 'asis'}
plot_cpm = function(y, feature, meta, title) {
  if (!feature %in% rownames(y$counts)) return(invisible(NULL))
  lcpm = cpm(y, log = TRUE)
  sids = intersect(colnames(lcpm), meta$sample)
  df = data.frame(
    lcpm = lcpm[feature, sids],
    condition = meta[sids, "condition"],
    donor = meta[sids, "donor"]
  )
  ggplot(df, aes(condition, lcpm, colour = condition)) +
    geom_boxplot(outlier.shape = NA, width = 0.4) +
    geom_jitter(aes(shape = donor), width = 0.1, size = 2) +
    scale_colour_manual(values = cbbPalette) +
    scale_shape_manual(values = 0:24) +
    theme_ng() +
    ylab("log2 gene-normalized CPM") +
    ggtitle(feature)
}

for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = de_results[[key]]
      if (is.null(fit)) next
      top6 = head(rownames(fit$tt), 6)
      plots = Filter(Negate(is.null),
                     lapply(top6, function(f) plot_cpm(fit$y, f, main_meta, f)))
      if (length(plots) > 0) {
        cat("\n## Top 6 normalized counts -", al, fs, gr, "\n")
        print(wrap_plots(plots, ncol = 3))
      }
    }
  }
}
```

## Repeat detection summary

Detection is defined as CPM > 1 in at least half the replicates of a condition.
CPM here uses raw repeat library sizes (not gene-normalized) to reflect absolute
detectability in the sequencing data. The gene_id granularity is used to count
individual repeat elements.

```{r detection-helpers}
detect_features = function(mat, sample_ids, cpm_thr = 1, min_frac = 0.5) {
  sids = intersect(sample_ids, colnames(mat))
  if (length(sids) == 0) return(character(0))
  lib_sizes = colSums(mat[, sids, drop = FALSE])
  lib_sizes[lib_sizes == 0] = 1
  cpm_mat = sweep(mat[, sids, drop = FALSE], 2, lib_sizes / 1e6, "/")
  n_above = rowSums(cpm_mat > cpm_thr)
  names(n_above)[n_above >= ceiling(min_frac * length(sids))]
}

load_class_map = function(indices_base, feature_set) {
  path = file.path(indices_base, paste0(feature_set, "_locus_map.tsv"))
  if (!file.exists(path)) return(NULL)
  lm = read_tsv(path,
                 col_names = c("transcript_id", "gene_id", "family_id", "class_id"),
                 show_col_types = FALSE)
  dedup = lm %>% select(gene_id, class_id) %>% distinct()
  setNames(dedup$class_id, dedup$gene_id)
}
```

```{r detection-data}
class_maps = list()
for (fs in feature_sets) {
  class_maps[[fs]] = load_class_map(indices_base, fs)
}

unsorted_meta = data.frame(
  sample = unsorted_samples,
  condition = "unsorted",
  row.names = unsorted_samples,
  stringsAsFactors = FALSE
)
all_meta = rbind(main_meta[, c("sample", "condition")], unsorted_meta)

det_summary = list()
det_class = list()

for (al in aligners) {
  for (fs in feature_sets) {
    mat = counts_cache[[paste(al, fs, "gene_id", sep = "_")]]
    if (is.null(mat)) next
    cm = class_maps[[fs]]
    for (cond_name in unique(all_meta$condition)) {
      sids = all_meta$sample[all_meta$condition == cond_name]
      detected = detect_features(mat, sids)
      det_summary[[length(det_summary) + 1]] = data.frame(
        aligner = al, feature_set = fs, condition = cond_name,
        n_detected = length(detected), stringsAsFactors = FALSE)
      if (!is.null(cm) && length(detected) > 0) {
        classes = cm[detected]
        classes[is.na(classes)] = "unknown"
        class_tab = as.data.frame(table(classes), stringsAsFactors = FALSE)
        colnames(class_tab) = c("class_id", "n")
        class_tab$aligner = al
        class_tab$feature_set = fs
        class_tab$condition = cond_name
        det_class[[length(det_class) + 1]] = class_tab
      }
    }
  }
}

det_summary_df = bind_rows(det_summary)
det_class_df = bind_rows(det_class)

if (nrow(det_class_df) > 0) {
  top_classes = det_class_df %>%
    group_by(class_id) %>%
    summarize(total = sum(n), .groups = "drop") %>%
    arrange(desc(total)) %>%
    head(7) %>%
    pull(class_id)
  det_class_df$class_label = ifelse(det_class_df$class_id %in% top_classes,
                                    det_class_df$class_id, "other")
  det_class_df = det_class_df %>%
    group_by(aligner, feature_set, condition, class_label) %>%
    summarize(n = sum(n), .groups = "drop")
}

save_rds(det_summary_df, "gse126543_detection_summary")
save_rds(det_class_df, "gse126543_detection_by_class")
```

### Features detected per condition

```{r detection-barplot, fig.width=10, fig.height=6}
if (nrow(det_summary_df) > 0) {
  ggplot(det_summary_df, aes(x = condition, y = n_detected, fill = feature_set)) +
    geom_col(position = "dodge") +
    facet_wrap(~aligner) +
    scale_fill_manual(values = cbbPalette) +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(y = "detected features (CPM > 1)", fill = "feature set",
         title = "Repeat features detected per condition (gene_id level)")
}
```

### Features detected per condition by repeat class

```{r detection-class-barplot, fig.width=12, fig.height=8}
if (nrow(det_class_df) > 0) {
  ggplot(det_class_df, aes(x = condition, y = n, fill = class_label)) +
    geom_col() +
    facet_grid(feature_set ~ aligner) +
    scale_fill_brewer(palette = "Set2") +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(y = "detected features (CPM > 1)", fill = "repeat class",
         title = "Detected repeats by class per condition (gene_id level)")
}
```

## Session info

```{r session-info}
sessionInfo()
```
