---
title: "GSE126543 bulk RNA-seq repeat quantification (gene-normalized)"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  counts_dir: "."
  report_outdir: "."
  indices_base: "."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(edgeR)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(tidyr)
library(readr)
library(patchwork)
library(pheatmap)
source(file.path("scripts", "ggtheme.R"))

counts_dir = params$counts_dir
report_outdir = params$report_outdir
indices_base = params$indices_base

aligners = c("star", "kallisto", "salmon")
granularities = c("gene_id", "family_id")
feature_sets = c("repeats", "genic_repeats", "intergenic_repeats")

donors = paste0("s", 1:7)
main_samples = c(paste0("tdp43pos_", donors), paste0("tdp43neg_", donors))
unsorted_samples = c("unsorted_s3", "unsorted_s4")

main_meta = data.frame(
  sample = main_samples,
  condition = rep(c("tdp43pos", "tdp43neg"), each = 7),
  donor = rep(donors, 2),
  row.names = main_samples,
  stringsAsFactors = FALSE
)

save_rds = function(obj, name) {
  path = file.path(report_outdir, paste0(name, ".rds"))
  saveRDS(obj, path)
  invisible(path)
}
```

## About this report

```{r provenance}
git_commit = tryCatch(
  system("git rev-parse --short HEAD", intern = TRUE),
  error = function(e) "unknown"
)
cat("GEO accession: GSE126543\n")
cat("SRA study: SRP185789\n")
cat("Rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
cat("Git commit:", git_commit, "\n")
cat("counts_dir:", counts_dir, "\n")
cat("report_outdir:", report_outdir, "\n")
```

GSE126543 (SRP185789): NeuN-positive neuronal nuclei from human frontal cortex
(FTD/ALS post-mortem tissue), sorted by TDP-43 nuclear presence across 7 donors.

- tdp43pos: TDP-43 retained in nucleus, n=7 donors
- tdp43neg: TDP-43 lost from nucleus, n=7 donors
- unsorted: mixed TDP-43 status, donors S3 and S4 only (shown in PCA for context)

Repeat counts are normalized by gene library sizes (total gene read counts per
sample) so that fold changes reflect changes in the repeat-to-gene expression
ratio. Donor-matched design; edgeR QLF with donor as blocking factor
(~donor + condition).

## Load counts

```{r load-counts}
load_counts = function(aligner, feature_set, granularity) {
  path = file.path(counts_dir,
                   paste0(aligner, "_", feature_set, "_", granularity, "_unique.tsv"))
  if (!file.exists(path)) return(NULL)
  mat = read_tsv(path, show_col_types = FALSE)
  feat_col = colnames(mat)[1]
  feat_ids = mat[[feat_col]]
  mat = as.matrix(mat[, -1, drop = FALSE])
  rownames(mat) = feat_ids
  storage.mode(mat) = "integer"
  mat
}

counts_cache = list()
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      counts_cache[[key]] = load_counts(al, fs, gr)
    }
  }
}
cat("Available count matrices:", sum(!sapply(counts_cache, is.null)), "\n")
save_rds(counts_cache, "gse126543_counts_cache")
```

## Gene library sizes for normalization

Gene counts are loaded per aligner to compute gene library sizes. These replace
the default repeat-derived library sizes in edgeR so that repeat expression is
measured relative to gene expression depth.

```{r load-gene-lib-sizes}
gene_lib_cache = list()
for (al in aligners) {
  gmat = load_counts(al, "genes", "gene_id")
  if (!is.null(gmat)) gene_lib_cache[[al]] = colSums(gmat)
}
cat("Gene library sizes available for:", paste(names(gene_lib_cache), collapse = ", "), "\n")
save_rds(gene_lib_cache, "gse126543_gene_lib_sizes")
```

```{r edger-function}
run_edger_paired = function(counts, meta, gene_lib_sizes) {
  samples = intersect(meta$sample, colnames(counts))
  if (length(samples) < 6) return(NULL)
  mat = counts[, samples, drop = FALSE]
  meta = meta[samples, , drop = FALSE]
  cond = factor(meta$condition, levels = c("tdp43neg", "tdp43pos"))
  donor = factor(meta$donor)
  y = DGEList(counts = mat, group = cond)
  keep = filterByExpr(y)
  y = y[keep, ]
  y$samples$lib.size = gene_lib_sizes[colnames(y)]
  y$samples$norm.factors = rep(1, ncol(y))
  design = model.matrix(~donor + cond)
  y = estimateDisp(y, design)
  fit = glmQLFit(y, design)
  qlf = glmQLFTest(fit, coef = ncol(design))
  list(y = y, fit = fit, qlf = qlf,
       tt = topTags(qlf, n = Inf, sort.by = "PValue")$table)
}
```

## TDP-43 nuclear positive vs negative

edgeR QLF with donor as blocking factor.
Repeat counts normalized by gene library sizes.

```{r de-edger}
de_results = list()
for (al in aligners) {
  gene_ls = gene_lib_cache[[al]]
  if (is.null(gene_ls)) next
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      mat = counts_cache[[key]]
      if (is.null(mat)) next
      res = run_edger_paired(mat, main_meta, gene_ls)
      if (!is.null(res)) de_results[[key]] = res
    }
  }
}
save_rds(de_results, "gse126543_de_edger")
cat("DE edgeR fits completed:", length(de_results), "\n")
```

```{r pca, fig.width=10, fig.height=10}
pca_plots = list()
for (al in aligners) {
  for (fs in feature_sets) {
    key = paste(al, fs, "gene_id", sep = "_")
    fit = de_results[[key]]
    if (is.null(fit)) next
    mat = counts_cache[[key]]
    all_sids = intersect(c(main_meta$sample, unsorted_samples), colnames(mat))
    mat_sub = mat[rowSums(mat[, all_sids, drop = FALSE]) > 0, all_sids, drop = FALSE]
    gene_ls = gene_lib_cache[[al]]
    gene_ls_sub = gene_ls[all_sids]
    lcpm = log2(sweep(mat_sub, 2, gene_ls_sub / 1e6, "/") + 0.5)
    pca = prcomp(t(lcpm), scale. = FALSE)
    pct = round(100 * pca$sdev^2 / sum(pca$sdev^2))
    cond_map = setNames(main_meta$condition, main_meta$sample)
    df = data.frame(
      PC1 = pca$x[, 1],
      PC2 = pca$x[, 2],
      condition = ifelse(rownames(pca$x) %in% names(cond_map),
                         cond_map[rownames(pca$x)], "unsorted"),
      sample = rownames(pca$x)
    )
    pca_plots[[key]] = ggplot(df, aes(PC1, PC2, colour = condition, label = sample)) +
      geom_point(size = 2) +
      geom_text_repel(size = 2.5, show.legend = FALSE) +
      xlab(paste0("PC1: ", pct[1], "%")) +
      ylab(paste0("PC2: ", pct[2], "%")) +
      scale_colour_manual(values = cbbPalette) +
      theme_ng() +
      ggtitle(paste(al, fs, "(gene-normalized)"))
  }
}
if (length(pca_plots) > 0) wrap_plots(pca_plots, ncol = 3)
save_rds(pca_plots, "gse126543_pca_plots")
```

```{r volcano, fig.width=10, fig.height=10}
for (gr in granularities) {
  vol_plots = list()
  for (al in aligners) {
    for (fs in feature_sets) {
      key = paste(al, fs, gr, sep = "_")
      fit = de_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      tt$sig = tt$FDR < 0.05 & abs(tt$logFC) > 1
      vol_plots[[key]] = ggplot(tt, aes(logFC, -log10(PValue), colour = sig)) +
        geom_point(size = 0.6, alpha = 0.6) +
        scale_colour_manual(values = c("grey70", cbbPalette[2]),
                            labels = c("ns", "FDR<0.05 & |lfc|>1")) +
        geom_vline(xintercept = c(-1, 1), linetype = "dashed", linewidth = 0.3) +
        xlab("log2 FC (repeat/gene ratio)") +
        theme_ng() +
        ggtitle(paste(al, fs, gr))
    }
  }
  if (length(vol_plots) > 0) print(wrap_plots(vol_plots, ncol = 3))
}
```

```{r ma, fig.width=10, fig.height=10}
for (gr in granularities) {
  ma_plots = list()
  for (al in aligners) {
    for (fs in feature_sets) {
      key = paste(al, fs, gr, sep = "_")
      fit = de_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      tt$sig = tt$FDR < 0.05
      ma_plots[[key]] = ggplot(tt, aes(logCPM, logFC, colour = sig)) +
        geom_point(size = 0.6, alpha = 0.6) +
        scale_colour_manual(values = c("grey70", cbbPalette[2]),
                            labels = c("ns", "FDR<0.05")) +
        geom_hline(yintercept = 0, linewidth = 0.3) +
        xlab("avg log2 gene-normalized CPM") +
        ylab("log2 FC (repeat/gene ratio)") +
        theme_ng() +
        ggtitle(paste(al, fs, gr))
    }
  }
  if (length(ma_plots) > 0) print(wrap_plots(ma_plots, ncol = 3))
}
```

```{r top-hits}
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = de_results[[key]]
      if (is.null(fit)) next
      top = fit$tt[fit$tt$FDR < 0.05, , drop = FALSE]
      if (nrow(top) == 0) next
      cat("\nTop hits -", al, fs, gr, "\n")
      print(knitr::kable(head(top[, c("logFC", "logCPM", "F", "PValue", "FDR")], 20),
                         digits = 4))
    }
  }
}
```

```{r heatmap, fig.width=10, fig.height=10}
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = de_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      sig_feats = rownames(tt)[tt$FDR < 0.05 & abs(tt$logFC) > 1]
      if (length(sig_feats) < 2) next
      sig_feats = head(sig_feats, 40)
      lcpm = cpm(fit$y, log = TRUE)[sig_feats, , drop = FALSE]
      sids = intersect(colnames(lcpm), main_meta$sample)
      ann = data.frame(
        condition = main_meta[sids, "condition"],
        donor = main_meta[sids, "donor"],
        row.names = sids
      )
      pheatmap(lcpm[, sids, drop = FALSE],
               annotation_col = ann,
               scale = "row",
               show_rownames = nrow(lcpm) <= 40,
               main = paste(al, fs, gr, "(gene-normalized)"),
               fontsize = 7,
               silent = TRUE)
    }
  }
}
```

```{r normalized-counts, fig.width=10, fig.height=10, results = 'asis'}
plot_cpm = function(y, feature, meta, title) {
  if (!feature %in% rownames(y$counts)) return(invisible(NULL))
  lcpm = cpm(y, log = TRUE)
  sids = intersect(colnames(lcpm), meta$sample)
  df = data.frame(
    lcpm = lcpm[feature, sids],
    condition = meta[sids, "condition"],
    donor = meta[sids, "donor"]
  )
  ggplot(df, aes(condition, lcpm, colour = condition)) +
    geom_boxplot(outlier.shape = NA, width = 0.4) +
    geom_jitter(aes(shape = donor), width = 0.1, size = 2) +
    scale_colour_manual(values = cbbPalette) +
    scale_shape_manual(values = 0:24) +
    theme_ng() +
    ylab("log2 gene-normalized CPM") +
    ggtitle(feature)
}

for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = de_results[[key]]
      if (is.null(fit)) next
      top6 = head(rownames(fit$tt), 6)
      plots = Filter(Negate(is.null),
                     lapply(top6, function(f) plot_cpm(fit$y, f, main_meta, f)))
      if (length(plots) > 0) {
        cat("\n## Top 6 normalized counts -", al, fs, gr, "\n")
        print(wrap_plots(plots, ncol = 3))
      }
    }
  }
}
```

## Repeat detection summary

Detection is defined as CPM > 1 in at least half the replicates of a condition.
CPM here uses raw repeat library sizes (not gene-normalized) to reflect absolute
detectability in the sequencing data. The gene_id granularity is used to count
individual repeat elements.

```{r detection-helpers}
detect_features = function(mat, sample_ids, cpm_thr = 1, min_frac = 0.5) {
  sids = intersect(sample_ids, colnames(mat))
  if (length(sids) == 0) return(character(0))
  lib_sizes = colSums(mat[, sids, drop = FALSE])
  lib_sizes[lib_sizes == 0] = 1
  cpm_mat = sweep(mat[, sids, drop = FALSE], 2, lib_sizes / 1e6, "/")
  n_above = rowSums(cpm_mat > cpm_thr)
  names(n_above)[n_above >= ceiling(min_frac * length(sids))]
}

load_class_map = function(indices_base, feature_set) {
  path = file.path(indices_base, paste0(feature_set, "_locus_map.tsv"))
  if (!file.exists(path)) return(NULL)
  lm = read_tsv(path,
                 col_names = c("transcript_id", "gene_id", "family_id", "class_id"),
                 show_col_types = FALSE)
  dedup = lm %>% select(gene_id, class_id) %>% distinct()
  setNames(dedup$class_id, dedup$gene_id)
}
```

```{r detection-data}
class_maps = list()
for (fs in feature_sets) {
  class_maps[[fs]] = load_class_map(indices_base, fs)
}

unsorted_meta = data.frame(
  sample = unsorted_samples,
  condition = "unsorted",
  row.names = unsorted_samples,
  stringsAsFactors = FALSE
)
all_meta = rbind(main_meta[, c("sample", "condition")], unsorted_meta)

det_summary = list()
det_class = list()

for (al in aligners) {
  for (fs in feature_sets) {
    mat = counts_cache[[paste(al, fs, "gene_id", sep = "_")]]
    if (is.null(mat)) next
    cm = class_maps[[fs]]
    for (cond_name in unique(all_meta$condition)) {
      sids = all_meta$sample[all_meta$condition == cond_name]
      detected = detect_features(mat, sids)
      det_summary[[length(det_summary) + 1]] = data.frame(
        aligner = al, feature_set = fs, condition = cond_name,
        n_detected = length(detected), stringsAsFactors = FALSE)
      if (!is.null(cm) && length(detected) > 0) {
        classes = cm[detected]
        classes[is.na(classes)] = "unknown"
        class_tab = as.data.frame(table(classes), stringsAsFactors = FALSE)
        colnames(class_tab) = c("class_id", "n")
        class_tab$aligner = al
        class_tab$feature_set = fs
        class_tab$condition = cond_name
        det_class[[length(det_class) + 1]] = class_tab
      }
    }
  }
}

det_summary_df = bind_rows(det_summary)
det_class_df = bind_rows(det_class)

if (nrow(det_class_df) > 0) {
  top_classes = det_class_df %>%
    group_by(class_id) %>%
    summarize(total = sum(n), .groups = "drop") %>%
    arrange(desc(total)) %>%
    head(7) %>%
    pull(class_id)
  det_class_df$class_label = ifelse(det_class_df$class_id %in% top_classes,
                                    det_class_df$class_id, "other")
  det_class_df = det_class_df %>%
    group_by(aligner, feature_set, condition, class_label) %>%
    summarize(n = sum(n), .groups = "drop")
}

save_rds(det_summary_df, "gse126543_detection_summary")
save_rds(det_class_df, "gse126543_detection_by_class")
```

### Features detected per condition

```{r detection-barplot, fig.width=10, fig.height=6}
if (nrow(det_summary_df) > 0) {
  ggplot(det_summary_df, aes(x = condition, y = n_detected, fill = feature_set)) +
    geom_col(position = "dodge") +
    facet_wrap(~aligner) +
    scale_fill_manual(values = cbbPalette) +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(y = "detected features (CPM > 1)", fill = "feature set",
         title = "Repeat features detected per condition (gene_id level)")
}
```

### Features detected per condition by repeat class

```{r detection-class-barplot, fig.width=12, fig.height=8}
if (nrow(det_class_df) > 0) {
  ggplot(det_class_df, aes(x = condition, y = n, fill = class_label)) +
    geom_col() +
    facet_grid(feature_set ~ aligner) +
    scale_fill_brewer(palette = "Set2") +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(y = "detected features (CPM > 1)", fill = "repeat class",
         title = "Detected repeats by class per condition (gene_id level)")
}
```

## Session info

```{r session-info}
sessionInfo()
```
