---
title: "GSE230647 bulk RNA-seq repeat quantification (gene-normalized)"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  counts_dir: "."
  report_outdir: "."
  indices_base: "."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(edgeR)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(tidyr)
library(readr)
library(patchwork)
library(pheatmap)
source(file.path("scripts", "ggtheme.R"))

counts_dir = params$counts_dir
report_outdir = params$report_outdir
indices_base = params$indices_base

aligners = c("star", "kallisto", "salmon")
granularities = c("gene_id", "family_id")
feature_sets = c("repeats", "genic_repeats", "intergenic_repeats")

oe_samples = c(paste0("tdp43_oe_r", 1:4), paste0("oe_ctrl_r", 1:4))
kd_samples = c(paste0("tdp43_kd_r", 1:4), paste0("nt_kd_ctrl_r", 1:4))

oe_meta = data.frame(
  sample = oe_samples,
  condition = rep(c("tdp43_oe", "oe_ctrl"), each = 4),
  replicate = rep(1:4, 2),
  row.names = oe_samples,
  stringsAsFactors = FALSE
)

kd_meta = data.frame(
  sample = kd_samples,
  condition = rep(c("tdp43_kd", "nt_kd_ctrl"), each = 4),
  replicate = rep(1:4, 2),
  row.names = kd_samples,
  stringsAsFactors = FALSE
)

save_rds = function(obj, name) {
  path = file.path(report_outdir, paste0(name, ".rds"))
  saveRDS(obj, path)
  invisible(path)
}
```

## Biology annotation for detection and DE stratification

```{r biology-setup}
source(file.path("scripts", "biology_checks.R"))

biology_annotation_path = file.path(indices_base, "repeat_biology_annotation.tsv")
biology_annotation = load_biology_annotation(biology_annotation_path)
if (is.null(biology_annotation)) {
  cat("No biology annotation at", biology_annotation_path,
      "- biology-stratified panels will be skipped.\n")
} else {
  cat("Loaded biology annotation:", nrow(biology_annotation), "subfamilies\n")
}
save_rds(biology_annotation, "biology_annotation")

annotate_de_with_biology = function(tt) {
  tt = annotate_biology(tt, biology_annotation, key = "gene_id")
  tt$polya = polya_label(tt$class_id)
  tt$age = family_age_label(tt$family_age)
  tt$strand = strand_vs_gene_label(tt$strand_vs_gene)
  tt$a_run = a_run_label(tt$frac_downstream_a_run, class_ids = tt$class_id)
  tt
}

de_stratify_counts = function(tt, key, fdr = 0.05) {
  tt = annotate_de_with_biology(tt)
  tt$direction = ifelse(is.na(tt$FDR) | tt$FDR >= fdr, "ns",
                        ifelse(tt$logFC > 0, "up", "down"))
  facet_cols = c(polya = "polya", family_age = "age",
                 strand_vs_gene = "strand", downstream_a_run = "a_run")
  rows = list()
  for (facet_nm in names(facet_cols)) {
    col = facet_cols[[facet_nm]]
    rows[[facet_nm]] = tt %>%
      group_by(label = .data[[col]], direction) %>%
      summarize(n = dplyr::n(), .groups = "drop") %>%
      mutate(facet = facet_nm, key = key)
  }
  bind_rows(rows)
}

plot_de_stratification = function(strat_df, title) {
  if (is.null(strat_df) || nrow(strat_df) == 0) return(invisible(NULL))
  strat_df$direction = factor(strat_df$direction, levels = c("up", "down", "ns"))
  ggplot(strat_df, aes(x = label, y = n, fill = direction)) +
    geom_col() +
    facet_grid(facet ~ key, scales = "free", space = "free_x") +
    scale_fill_manual(values = c(up = "firebrick", down = "steelblue", ns = "grey70")) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = title, x = NULL, y = "features")
}
```

## About this report

```{r provenance}
git_commit = tryCatch(
  system("git rev-parse --short HEAD", intern = TRUE),
  error = function(e) "unknown"
)
cat("GEO accession: GSE230647\n")
cat("SRA study: SRP434658\n")
cat("Rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
cat("Git commit:", git_commit, "\n")
cat("counts_dir:", counts_dir, "\n")
cat("report_outdir:", report_outdir, "\n")
```

GSE230647 (SRP434658): iPSC-derived neural cultures with TDP-43 perturbations.
Two experiments:

- overexpression (OE): doxycycline-inducible HA-TDP43 on (tdp43_oe, n=4) vs DOX off control (oe_ctrl, n=4)
- knockdown (KD): TDP-43 shRNA (tdp43_kd, n=4) vs non-targeting shRNA control (nt_kd_ctrl, n=4)

Quantification by STAR, kallisto, and salmon at gene_id and family_id granularities
for three repeat feature sets: repeats, genic_repeats, and intergenic_repeats.

Repeat counts are normalized by gene library sizes (total gene read counts per
sample) so that fold changes reflect changes in the repeat-to-gene expression
ratio. Differential expression tested with edgeR QLF (quasi-likelihood F-test).

## PolyA-competence of detected repeats

Detection and differential-expression results can be split into two groups
based on the intrinsic ability of each repeat class to produce a
polyadenylated transcript. This split uses only the `class_id` column from
`{feature_set}_locus_map.tsv` (equivalent to RepeatMasker `repClass`); no
sequence-level polyA scanning is performed.

- with polyA (polyA-competent): `LINE`, `SINE`, `LTR`, `Retroposon`.
  These classes produce polyadenylated transcripts when expressed from their
  own promoters (LINE-1 and ERV internal transcripts carry canonical AATAAA
  poly(A) sites, SINE/Alu relies on LINE-1 machinery and poly(A) tails,
  Retroposon/SVA also polyadenylates).
- without polyA (not polyA-competent): every other class, including
  `Satellite`, `Simple_repeat`, `Low_complexity`, `DNA`, `RC`, and the
  structural-RNA-derived classes (`rRNA`, `tRNA`, `snRNA`, `scRNA`,
  `srpRNA`). These lack a canonical polyadenylation signal in their body.
- unknown: loci with no class annotation in the locus map.

The genic vs intergenic feature_set split is complementary: genic repeats can
be captured through a polyadenylated host transcript regardless of intrinsic
class, while intergenic repeats need their own poly(A) tail to be detected by
polyA-selection assays. The class-based facet therefore matters most on the
`intergenic_repeats` feature set.

## Biology-driven sanity checks for detection

Three additional commonsense checks help reason about why a given repeat is
or is not detected. Classification logic is shared across reports in
`workflow/scripts/biology_annotations.py` (Python, pure functions, unit
tested in `test/unit/test_biology_annotations.py`) and the R-side join and
labelling helpers live in `workflow/scripts/biology_checks.R`.

- Family age (from RepeatMasker subfamily name or milliDiv). Younger
  subfamilies are more often transcriptionally active, older ones are
  generally silenced. Brackets:

      young   L1HS, L1PA2-7, AluY, SVA_A-F, HERVK/HERVH/LTR5
      medium  L1PA8-16, AluS
      old     L1PA17+, L1PB, L1M*, AluJ, FLAM, FRAM
      unknown anything not matching, including DNA/RC/satellite classes

  When milliDiv is available the same brackets are produced from
  divergence thresholds (<=50 young, <=200 medium, >200 old).

- Strand concordance with overlapping genes. By construction
  `genic_repeats` is built with a same-strand intersect, so loci in that
  set are sense-with-host. The shared annotation labels every locus as
  `sense`, `antisense`, `both`, or `intergenic` against a strand-aware
  gene-body overlap so reports can also surface antisense loci that the
  current `intergenic_repeats` set silently absorbs.

- Internal A-tract proxy for polyA priming. Length of the longest run of
  As in the genomic region downstream of each locus (in mRNA-sense
  orientation). A long downstream A-run flags a repeat that may be
  detected by internal priming rather than by a genuine poly(A) tail,
  which matters most for intergenic non-polyA-competent classes.

The shared annotation is loaded from
`{indices_base}/indices/{genome_tag}/repeat_biology_annotation.tsv` when
present; otherwise the helpers no-op and reports fall back to the
class-only polyA facet.

## Load counts

```{r load-counts}
load_counts = function(aligner, feature_set, granularity) {
  path = file.path(counts_dir,
                   paste0(aligner, "_", feature_set, "_", granularity, "_unique.tsv"))
  if (!file.exists(path)) return(NULL)
  mat = read_tsv(path, show_col_types = FALSE)
  feat_col = colnames(mat)[1]
  feat_ids = mat[[feat_col]]
  mat = as.matrix(mat[, -1, drop = FALSE])
  rownames(mat) = feat_ids
  storage.mode(mat) = "integer"
  mat
}

counts_cache = list()
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      counts_cache[[key]] = load_counts(al, fs, gr)
    }
  }
}
cat("Available count matrices:", sum(!sapply(counts_cache, is.null)), "\n")
save_rds(counts_cache, "gse230647_counts_cache")
```

## Gene library sizes for normalization

Gene counts are loaded per aligner to compute gene library sizes. These replace
the default repeat-derived library sizes in edgeR so that repeat expression is
measured relative to gene expression depth.

```{r load-gene-lib-sizes}
gene_lib_cache = list()
for (al in aligners) {
  gmat = load_counts(al, "genes", "gene_id")
  if (!is.null(gmat)) gene_lib_cache[[al]] = colSums(gmat)
}
cat("Gene library sizes available for:", paste(names(gene_lib_cache), collapse = ", "), "\n")
save_rds(gene_lib_cache, "gse230647_gene_lib_sizes")
```

```{r edger-function}
run_edger = function(counts, meta, treat_level, ref_level, gene_lib_sizes) {
  samples = intersect(meta$sample, colnames(counts))
  if (length(samples) < 4) return(NULL)
  mat = counts[, samples, drop = FALSE]
  cond = factor(meta[samples, "condition"], levels = c(ref_level, treat_level))
  y = DGEList(counts = mat, group = cond)
  keep = filterByExpr(y)
  y = y[keep, ]
  y$samples$lib.size = gene_lib_sizes[colnames(y)]
  y$samples$norm.factors = rep(1, ncol(y))
  design = model.matrix(~cond)
  y = estimateDisp(y, design)
  fit = glmQLFit(y, design)
  qlf = glmQLFTest(fit, coef = 2)
  list(y = y, fit = fit, qlf = qlf,
       tt = topTags(qlf, n = Inf, sort.by = "PValue")$table)
}
```

## Overexpression experiment

TDP-43-HA ON (tdp43_oe) vs DOX off control (oe_ctrl).
Repeat counts normalized by gene library sizes.

```{r oe-edger}
oe_results = list()
for (al in aligners) {
  gene_ls = gene_lib_cache[[al]]
  if (is.null(gene_ls)) next
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      mat = counts_cache[[key]]
      if (is.null(mat)) next
      res = run_edger(mat, oe_meta, treat_level = "tdp43_oe",
                      ref_level = "oe_ctrl", gene_ls)
      if (!is.null(res)) oe_results[[key]] = res
    }
  }
}
save_rds(oe_results, "gse230647_oe_edger")
cat("OE edgeR fits completed:", length(oe_results), "\n")
```

```{r oe-pca, fig.width=10, fig.height=10}
pca_plots = list()
for (al in aligners) {
  for (fs in feature_sets) {
    key = paste(al, fs, "gene_id", sep = "_")
    fit = oe_results[[key]]
    if (is.null(fit)) next
    lcpm = cpm(fit$y, log = TRUE)
    pca = prcomp(t(lcpm), scale. = FALSE)
    pct = round(100 * pca$sdev^2 / sum(pca$sdev^2))
    df = data.frame(
      PC1 = pca$x[, 1],
      PC2 = pca$x[, 2],
      condition = fit$y$samples$group,
      sample = rownames(pca$x)
    )
    pca_plots[[key]] = ggplot(df, aes(PC1, PC2, colour = condition, label = sample)) +
      geom_point(size = 2) +
      geom_text_repel(size = 2.5, show.legend = FALSE) +
      xlab(paste0("PC1: ", pct[1], "%")) +
      ylab(paste0("PC2: ", pct[2], "%")) +
      scale_colour_manual(values = cbbPalette) +
      theme_ng() +
      ggtitle(paste(al, fs, "(gene-normalized)"))
  }
}
if (length(pca_plots) > 0) wrap_plots(pca_plots, ncol = 3)
save_rds(pca_plots, "gse230647_oe_pca_plots")
```

```{r oe-volcano, fig.width=10, fig.height=10}
for (gr in granularities) {
  vol_plots = list()
  for (al in aligners) {
    for (fs in feature_sets) {
      key = paste(al, fs, gr, sep = "_")
      fit = oe_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      tt$sig = tt$FDR < 0.05 & abs(tt$logFC) > 1
      vol_plots[[key]] = ggplot(tt, aes(logFC, -log10(PValue), colour = sig)) +
        geom_point(size = 0.6, alpha = 0.6) +
        scale_colour_manual(values = c("grey70", cbbPalette[2]),
                            labels = c("ns", "FDR<0.05 & |lfc|>1")) +
        geom_vline(xintercept = c(-1, 1), linetype = "dashed", linewidth = 0.3) +
        xlab("log2 FC (repeat/gene ratio)") +
        theme_ng() +
        ggtitle(paste(al, fs, gr))
    }
  }
  if (length(vol_plots) > 0) print(wrap_plots(vol_plots, ncol = 3))
}
```

```{r oe-ma, fig.width=10, fig.height=10}
for (gr in granularities) {
  ma_plots = list()
  for (al in aligners) {
    for (fs in feature_sets) {
      key = paste(al, fs, gr, sep = "_")
      fit = oe_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      tt$sig = tt$FDR < 0.05
      ma_plots[[key]] = ggplot(tt, aes(logCPM, logFC, colour = sig)) +
        geom_point(size = 0.6, alpha = 0.6) +
        scale_colour_manual(values = c("grey70", cbbPalette[2]),
                            labels = c("ns", "FDR<0.05")) +
        geom_hline(yintercept = 0, linewidth = 0.3) +
        xlab("avg log2 gene-normalized CPM") +
        ylab("log2 FC (repeat/gene ratio)") +
        theme_ng() +
        ggtitle(paste(al, fs, gr))
    }
  }
  if (length(ma_plots) > 0) print(wrap_plots(ma_plots, ncol = 3))
}
```

### OE DE stratified by biology (polyA / age / strand / A-run)

```{r biology-de-strat-oe, fig.width=14, fig.height=10}
if (!is.null(biology_annotation) && length(oe_results) > 0) {
  strat_oe = bind_rows(lapply(names(oe_results), function(key) {
    de_stratify_counts(oe_results[[key]]$tt, key)
  }))
  save_rds(strat_oe, "gse230647_oe_biology_stratification")
  for (facet_nm in unique(strat_oe$facet)) {
    print(plot_de_stratification(filter(strat_oe, facet == facet_nm),
                                 paste("OE, stratified by", facet_nm)))
  }
}
```

```{r oe-top-hits}
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = oe_results[[key]]
      if (is.null(fit)) next
      top = fit$tt[fit$tt$FDR < 0.05, , drop = FALSE]
      if (nrow(top) == 0) next
      cat("\nOE top hits -", al, fs, gr, "\n")
      print(knitr::kable(head(top[, c("logFC", "logCPM", "F", "PValue", "FDR")], 20),
                         digits = 4))
    }
  }
}
```

```{r oe-heatmap, fig.width=ng_fig_size(2, 1, panel_mm = 60)$w, fig.height=ng_fig_size(2, 1, panel_mm = 60)$h}
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = oe_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      sig_feats = rownames(tt)[tt$FDR < 0.05 & abs(tt$logFC) > 1]
      if (length(sig_feats) < 2) next
      sig_feats = head(sig_feats, 40)
      lcpm = cpm(fit$y, log = TRUE)[sig_feats, , drop = FALSE]
      ann = data.frame(condition = fit$y$samples$group,
                       row.names = colnames(lcpm))
      pheatmap(lcpm,
               annotation_col = ann,
               scale = "row",
               show_rownames = nrow(lcpm) <= 40,
               main = paste("OE", al, fs, gr, "(gene-normalized)"),
               fontsize = 7,
               silent = TRUE)
    }
  }
}
```

```{r oe-counts, fig.width=10, fig.height=10}
plot_cpm = function(y, feature, title) {
  if (!feature %in% rownames(y$counts)) return(invisible(NULL))
  lcpm = cpm(y, log = TRUE)
  df = data.frame(
    lcpm = lcpm[feature, ],
    condition = y$samples$group,
    replicate = seq_len(ncol(lcpm))
  )
  ggplot(df, aes(condition, lcpm, colour = condition)) +
    geom_boxplot(outlier.shape = NA, width = 0.4) +
    geom_jitter(aes(shape = factor(replicate)), width = 0.1, size = 2) +
    scale_colour_manual(values = cbbPalette) +
    theme_ng() +
    ylab("log2 gene-normalized CPM") +
    ggtitle(feature)
}

for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = oe_results[[key]]
      if (is.null(fit)) next
      top6 = head(rownames(fit$tt), 6)
      plots = Filter(Negate(is.null),
                     lapply(top6, function(f) plot_cpm(fit$y, f, f)))
      if (length(plots) > 0) {
        cat("\nOE top 6 normalized counts -", al, fs, gr, "\n")
        print(wrap_plots(plots, ncol = 3))
      }
    }
  }
}
```

## Knockdown experiment

TDP-43 shRNA (tdp43_kd) vs non-targeting shRNA control (nt_kd_ctrl).
Repeat counts normalized by gene library sizes.

```{r kd-edger}
kd_results = list()
for (al in aligners) {
  gene_ls = gene_lib_cache[[al]]
  if (is.null(gene_ls)) next
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      mat = counts_cache[[key]]
      if (is.null(mat)) next
      res = run_edger(mat, kd_meta, treat_level = "tdp43_kd",
                      ref_level = "nt_kd_ctrl", gene_ls)
      if (!is.null(res)) kd_results[[key]] = res
    }
  }
}
save_rds(kd_results, "gse230647_kd_edger")
cat("KD edgeR fits completed:", length(kd_results), "\n")
```

```{r kd-pca, fig.width=10, fig.height=10}
pca_plots = list()
for (al in aligners) {
  for (fs in feature_sets) {
    key = paste(al, fs, "gene_id", sep = "_")
    fit = kd_results[[key]]
    if (is.null(fit)) next
    lcpm = cpm(fit$y, log = TRUE)
    pca = prcomp(t(lcpm), scale. = FALSE)
    pct = round(100 * pca$sdev^2 / sum(pca$sdev^2))
    df = data.frame(
      PC1 = pca$x[, 1],
      PC2 = pca$x[, 2],
      condition = fit$y$samples$group,
      sample = rownames(pca$x)
    )
    pca_plots[[key]] = ggplot(df, aes(PC1, PC2, colour = condition, label = sample)) +
      geom_point(size = 2) +
      geom_text_repel(size = 2.5, show.legend = FALSE) +
      xlab(paste0("PC1: ", pct[1], "%")) +
      ylab(paste0("PC2: ", pct[2], "%")) +
      scale_colour_manual(values = cbbPalette) +
      theme_ng() +
      ggtitle(paste(al, fs, "(gene-normalized)"))
  }
}
if (length(pca_plots) > 0) wrap_plots(pca_plots, ncol = 3)
save_rds(pca_plots, "gse230647_kd_pca_plots")
```

```{r kd-volcano, fig.width=10, fig.height=10}
for (gr in granularities) {
  vol_plots = list()
  for (al in aligners) {
    for (fs in feature_sets) {
      key = paste(al, fs, gr, sep = "_")
      fit = kd_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      tt$sig = tt$FDR < 0.05 & abs(tt$logFC) > 1
      vol_plots[[key]] = ggplot(tt, aes(logFC, -log10(PValue), colour = sig)) +
        geom_point(size = 0.6, alpha = 0.6) +
        scale_colour_manual(values = c("grey70", cbbPalette[2]),
                            labels = c("ns", "FDR<0.05 & |lfc|>1")) +
        geom_vline(xintercept = c(-1, 1), linetype = "dashed", linewidth = 0.3) +
        xlab("log2 FC (repeat/gene ratio)") +
        theme_ng() +
        ggtitle(paste(al, fs, gr))
    }
  }
  if (length(vol_plots) > 0) print(wrap_plots(vol_plots, ncol = 3))
}
```

```{r kd-ma, fig.width=10, fig.height=10}
for (gr in granularities) {
  ma_plots = list()
  for (al in aligners) {
    for (fs in feature_sets) {
      key = paste(al, fs, gr, sep = "_")
      fit = kd_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      tt$sig = tt$FDR < 0.05
      ma_plots[[key]] = ggplot(tt, aes(logCPM, logFC, colour = sig)) +
        geom_point(size = 0.6, alpha = 0.6) +
        scale_colour_manual(values = c("grey70", cbbPalette[2]),
                            labels = c("ns", "FDR<0.05")) +
        geom_hline(yintercept = 0, linewidth = 0.3) +
        xlab("avg log2 gene-normalized CPM") +
        ylab("log2 FC (repeat/gene ratio)") +
        theme_ng() +
        ggtitle(paste(al, fs, gr))
    }
  }
  if (length(ma_plots) > 0) print(wrap_plots(ma_plots, ncol = 3))
}
```

### KD DE stratified by biology (polyA / age / strand / A-run)

```{r biology-de-strat-kd, fig.width=14, fig.height=10}
if (!is.null(biology_annotation) && length(kd_results) > 0) {
  strat_kd = bind_rows(lapply(names(kd_results), function(key) {
    de_stratify_counts(kd_results[[key]]$tt, key)
  }))
  save_rds(strat_kd, "gse230647_kd_biology_stratification")
  for (facet_nm in unique(strat_kd$facet)) {
    print(plot_de_stratification(filter(strat_kd, facet == facet_nm),
                                 paste("KD, stratified by", facet_nm)))
  }
}
```

```{r kd-top-hits}
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = kd_results[[key]]
      if (is.null(fit)) next
      top = fit$tt[fit$tt$FDR < 0.05, , drop = FALSE]
      if (nrow(top) == 0) next
      cat("\nKD top hits -", al, fs, gr, "\n")
      print(knitr::kable(head(top[, c("logFC", "logCPM", "F", "PValue", "FDR")], 20),
                         digits = 4))
    }
  }
}
```

```{r kd-heatmap, fig.width=10, fig.height=10}
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = kd_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      sig_feats = rownames(tt)[tt$FDR < 0.05 & abs(tt$logFC) > 1]
      if (length(sig_feats) < 2) next
      sig_feats = head(sig_feats, 40)
      lcpm = cpm(fit$y, log = TRUE)[sig_feats, , drop = FALSE]
      ann = data.frame(condition = fit$y$samples$group,
                       row.names = colnames(lcpm))
      pheatmap(lcpm,
               annotation_col = ann,
               scale = "row",
               show_rownames = nrow(lcpm) <= 40,
               main = paste("KD", al, fs, gr, "(gene-normalized)"),
               fontsize = 7,
               silent = TRUE)
    }
  }
}
```

```{r kd-counts, fig.width=10, fig.height=10}
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = kd_results[[key]]
      if (is.null(fit)) next
      top6 = head(rownames(fit$tt), 6)
      plots = Filter(Negate(is.null),
                     lapply(top6, function(f) plot_cpm(fit$y, f, f)))
      if (length(plots) > 0) {
        cat("\nKD top 6 normalized counts -", al, fs, gr, "\n")
        print(wrap_plots(plots, ncol = 3))
      }
    }
  }
}
```

## Repeat detection summary

Detection is defined as CPM > 1 in at least half the replicates of a condition.
CPM here uses raw repeat library sizes (not gene-normalized) to reflect absolute
detectability in the sequencing data. The gene_id granularity is used to count
individual repeat elements.

```{r detection-helpers}
detect_features = function(mat, sample_ids, cpm_thr = 1, min_frac = 0.5) {
  sids = intersect(sample_ids, colnames(mat))
  if (length(sids) == 0) return(character(0))
  lib_sizes = colSums(mat[, sids, drop = FALSE])
  lib_sizes[lib_sizes == 0] = 1
  cpm_mat = sweep(mat[, sids, drop = FALSE], 2, lib_sizes / 1e6, "/")
  n_above = rowSums(cpm_mat > cpm_thr)
  names(n_above)[n_above >= ceiling(min_frac * length(sids))]
}

load_class_map = function(indices_base, feature_set) {
  path = file.path(indices_base, paste0(feature_set, "_locus_map.tsv"))
  if (!file.exists(path)) return(NULL)
  lm = read_tsv(path,
                 col_names = c("transcript_id", "gene_id", "family_id", "class_id"),
                 show_col_types = FALSE)
  dedup = lm %>% select(gene_id, class_id) %>% distinct()
  setNames(dedup$class_id, dedup$gene_id)
}
```

```{r detection-data}
class_maps = list()
for (fs in feature_sets) {
  class_maps[[fs]] = load_class_map(indices_base, fs)
}

all_meta = rbind(oe_meta[, c("sample", "condition")],
                 kd_meta[, c("sample", "condition")])

det_summary = list()
det_class = list()

for (al in aligners) {
  for (fs in feature_sets) {
    mat = counts_cache[[paste(al, fs, "gene_id", sep = "_")]]
    if (is.null(mat)) next
    cm = class_maps[[fs]]
    for (cond_name in unique(all_meta$condition)) {
      sids = all_meta$sample[all_meta$condition == cond_name]
      detected = detect_features(mat, sids)
      det_summary[[length(det_summary) + 1]] = data.frame(
        aligner = al, feature_set = fs, condition = cond_name,
        n_detected = length(detected), stringsAsFactors = FALSE)
      if (!is.null(cm) && length(detected) > 0) {
        classes = cm[detected]
        classes[is.na(classes)] = "unknown"
        class_tab = as.data.frame(table(classes), stringsAsFactors = FALSE)
        colnames(class_tab) = c("class_id", "n")
        class_tab$aligner = al
        class_tab$feature_set = fs
        class_tab$condition = cond_name
        det_class[[length(det_class) + 1]] = class_tab
      }
    }
  }
}

det_summary_df = bind_rows(det_summary)
det_class_df = bind_rows(det_class)

if (nrow(det_class_df) > 0) {
  top_classes = det_class_df %>%
    group_by(class_id) %>%
    summarize(total = sum(n), .groups = "drop") %>%
    arrange(desc(total)) %>%
    head(7) %>%
    pull(class_id)
  det_class_df$class_label = ifelse(det_class_df$class_id %in% top_classes,
                                    det_class_df$class_id, "other")
  det_class_df = det_class_df %>%
    group_by(aligner, feature_set, condition, class_label) %>%
    summarize(n = sum(n), .groups = "drop")
}

save_rds(det_summary_df, "gse230647_detection_summary")
save_rds(det_class_df, "gse230647_detection_by_class")
```

### Features detected per condition

```{r detection-barplot, fig.width=10, fig.height=6}
if (nrow(det_summary_df) > 0) {
  ggplot(det_summary_df, aes(x = condition, y = n_detected, fill = feature_set)) +
    geom_col(position = "dodge") +
    facet_wrap(~aligner) +
    scale_fill_manual(values = cbbPalette) +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(y = "detected features (CPM > 1)", fill = "feature set",
         title = "Repeat features detected per condition (gene_id level)")
}
```

### Features detected per condition by repeat class

```{r detection-class-barplot, fig.width=12, fig.height=8}
if (nrow(det_class_df) > 0) {
  ggplot(det_class_df, aes(x = condition, y = n, fill = class_label)) +
    geom_col() +
    facet_grid(feature_set ~ aligner) +
    scale_fill_brewer(palette = "Set2") +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(y = "detected features (CPM > 1)", fill = "repeat class",
         title = "Detected repeats by class per condition (gene_id level)")
}
```

## Session info

```{r session-info}
sessionInfo()
```
