---
title: "GSE230647 bulk RNA-seq repeat quantification"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  counts_dir: "."
  report_outdir: "."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(edgeR)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(tidyr)
library(readr)
library(patchwork)
library(pheatmap)
source(file.path("scripts", "ggtheme.R"))

counts_dir = params$counts_dir
report_outdir = params$report_outdir

aligners = c("star", "kallisto", "salmon")
granularities = c("gene_id", "family_id")
feature_sets = c("repeats", "genic_repeats", "intergenic_repeats")

oe_samples = c(paste0("tdp43_oe_r", 1:4), paste0("oe_ctrl_r", 1:4))
kd_samples = c(paste0("tdp43_kd_r", 1:4), paste0("nt_kd_ctrl_r", 1:4))

oe_meta = data.frame(
  sample = oe_samples,
  condition = rep(c("tdp43_oe", "oe_ctrl"), each = 4),
  replicate = rep(1:4, 2),
  row.names = oe_samples,
  stringsAsFactors = FALSE
)

kd_meta = data.frame(
  sample = kd_samples,
  condition = rep(c("tdp43_kd", "nt_kd_ctrl"), each = 4),
  replicate = rep(1:4, 2),
  row.names = kd_samples,
  stringsAsFactors = FALSE
)

save_rds = function(obj, name) {
  path = file.path(report_outdir, paste0(name, ".rds"))
  saveRDS(obj, path)
  invisible(path)
}
```

## About this report

```{r provenance}
git_commit = tryCatch(
  system("git rev-parse --short HEAD", intern = TRUE),
  error = function(e) "unknown"
)
cat("GEO accession: GSE230647\n")
cat("SRA study: SRP434658\n")
cat("Rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
cat("Git commit:", git_commit, "\n")
cat("counts_dir:", counts_dir, "\n")
cat("report_outdir:", report_outdir, "\n")
```

GSE230647 (SRP434658): iPSC-derived neural cultures with TDP-43 perturbations.
Two experiments:

- overexpression (OE): doxycycline-inducible HA-TDP43 on (tdp43_oe, n=4) vs DOX off control (oe_ctrl, n=4)
- knockdown (KD): TDP-43 shRNA (tdp43_kd, n=4) vs non-targeting shRNA control (nt_kd_ctrl, n=4)

Quantification by STAR, kallisto, and salmon at gene_id and family_id granularities
for three repeat feature sets: repeats, genic_repeats, and intergenic_repeats.
Differential expression tested with edgeR QLF (quasi-likelihood F-test).

## Load counts

```{r load-counts}
load_counts = function(aligner, feature_set, granularity) {
  path = file.path(counts_dir,
                   paste0(aligner, "_", feature_set, "_", granularity, "_unique.tsv"))
  if (!file.exists(path)) return(NULL)
  mat = read_tsv(path, show_col_types = FALSE)
  feat_col = colnames(mat)[1]
  feat_ids = mat[[feat_col]]
  mat = as.matrix(mat[, -1, drop = FALSE])
  rownames(mat) = feat_ids
  storage.mode(mat) = "integer"
  mat
}

counts_cache = list()
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      counts_cache[[key]] = load_counts(al, fs, gr)
    }
  }
}
cat("Available count matrices:", sum(!sapply(counts_cache, is.null)), "\n")
save_rds(counts_cache, "gse230647_counts_cache")
```

```{r edger-function}
run_edger = function(counts, meta, treat_level, ref_level) {
  samples = intersect(meta$sample, colnames(counts))
  if (length(samples) < 4) return(NULL)
  mat = counts[, samples, drop = FALSE]
  cond = factor(meta[samples, "condition"], levels = c(ref_level, treat_level))
  y = DGEList(counts = mat, group = cond)
  keep = filterByExpr(y)
  y = y[keep, , keep.lib.sizes = FALSE]
  y = normLibSizes(y)
  design = model.matrix(~cond)
  y = estimateDisp(y, design)
  fit = glmQLFit(y, design)
  qlf = glmQLFTest(fit, coef = 2)
  list(y = y, fit = fit, qlf = qlf,
       tt = topTags(qlf, n = Inf, sort.by = "PValue")$table)
}
```

## Overexpression experiment

TDP-43-HA ON (tdp43_oe) vs DOX off control (oe_ctrl).

```{r oe-edger}
oe_results = list()
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      mat = counts_cache[[key]]
      if (is.null(mat)) next
      res = run_edger(mat, oe_meta, treat_level = "tdp43_oe", ref_level = "oe_ctrl")
      if (!is.null(res)) oe_results[[key]] = res
    }
  }
}
save_rds(oe_results, "gse230647_oe_edger")
cat("OE edgeR fits completed:", length(oe_results), "\n")
```

```{r oe-pca, fig.width=10, fig.height=10}
pca_plots = list()
for (al in aligners) {
  for (fs in feature_sets) {
    key = paste(al, fs, "gene_id", sep = "_")
    fit = oe_results[[key]]
    if (is.null(fit)) next
    lcpm = cpm(fit$y, log = TRUE)
    pca = prcomp(t(lcpm), scale. = FALSE)
    pct = round(100 * pca$sdev^2 / sum(pca$sdev^2))
    df = data.frame(
      PC1 = pca$x[, 1],
      PC2 = pca$x[, 2],
      condition = fit$y$samples$group,
      sample = rownames(pca$x)
    )
    pca_plots[[key]] = ggplot(df, aes(PC1, PC2, colour = condition, label = sample)) +
      geom_point(size = 2) +
      geom_text_repel(size = 2.5, show.legend = FALSE) +
      xlab(paste0("PC1: ", pct[1], "%")) +
      ylab(paste0("PC2: ", pct[2], "%")) +
      scale_colour_manual(values = cbbPalette) +
      theme_ng() +
      ggtitle(paste(al, fs))
  }
}
if (length(pca_plots) > 0) wrap_plots(pca_plots, ncol = 3)
save_rds(pca_plots, "gse230647_oe_pca_plots")
```

```{r oe-volcano, fig.width=10, fig.height=10}
for (gr in granularities) {
  vol_plots = list()
  for (al in aligners) {
    for (fs in feature_sets) {
      key = paste(al, fs, gr, sep = "_")
      fit = oe_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      tt$sig = tt$FDR < 0.05 & abs(tt$logFC) > 1
      vol_plots[[key]] = ggplot(tt, aes(logFC, -log10(PValue), colour = sig)) +
        geom_point(size = 0.6, alpha = 0.6) +
        scale_colour_manual(values = c("grey70", cbbPalette[2]),
                            labels = c("ns", "FDR<0.05 & |lfc|>1")) +
        geom_vline(xintercept = c(-1, 1), linetype = "dashed", linewidth = 0.3) +
        theme_ng() +
        ggtitle(paste(al, fs, gr))
    }
  }
  if (length(vol_plots) > 0) print(wrap_plots(vol_plots, ncol = 3))
}
```

```{r oe-ma, fig.width=10, fig.height=10}
for (gr in granularities) {
  ma_plots = list()
  for (al in aligners) {
    for (fs in feature_sets) {
      key = paste(al, fs, gr, sep = "_")
      fit = oe_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      tt$sig = tt$FDR < 0.05
      ma_plots[[key]] = ggplot(tt, aes(logCPM, logFC, colour = sig)) +
        geom_point(size = 0.6, alpha = 0.6) +
        scale_colour_manual(values = c("grey70", cbbPalette[2]),
                            labels = c("ns", "FDR<0.05")) +
        geom_hline(yintercept = 0, linewidth = 0.3) +
        theme_ng() +
        ggtitle(paste(al, fs, gr))
    }
  }
  if (length(ma_plots) > 0) print(wrap_plots(ma_plots, ncol = 3))
}
```

```{r oe-top-hits}
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = oe_results[[key]]
      if (is.null(fit)) next
      top = fit$tt[fit$tt$FDR < 0.05, , drop = FALSE]
      if (nrow(top) == 0) next
      cat("\nOE top hits -", al, fs, gr, "\n")
      print(knitr::kable(head(top[, c("logFC", "logCPM", "F", "PValue", "FDR")], 20),
                         digits = 4))
    }
  }
}
```

```{r oe-heatmap, fig.width=ng_fig_size(2, 1, panel_mm = 60)$w, fig.height=ng_fig_size(2, 1, panel_mm = 60)$h}
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = oe_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      sig_feats = rownames(tt)[tt$FDR < 0.05 & abs(tt$logFC) > 1]
      if (length(sig_feats) < 2) next
      sig_feats = head(sig_feats, 40)
      lcpm = cpm(fit$y, log = TRUE)[sig_feats, , drop = FALSE]
      ann = data.frame(condition = fit$y$samples$group,
                       row.names = colnames(lcpm))
      pheatmap(lcpm,
               annotation_col = ann,
               scale = "row",
               show_rownames = nrow(lcpm) <= 40,
               main = paste("OE", al, fs, gr),
               fontsize = 7,
               silent = TRUE)
    }
  }
}
```

```{r oe-counts, fig.width=10, fig.height=10}
plot_cpm = function(y, feature, title) {
  if (!feature %in% rownames(y$counts)) return(invisible(NULL))
  lcpm = cpm(y, log = TRUE)
  df = data.frame(
    lcpm = lcpm[feature, ],
    condition = y$samples$group,
    replicate = seq_len(ncol(lcpm))
  )
  ggplot(df, aes(condition, lcpm, colour = condition)) +
    geom_boxplot(outlier.shape = NA, width = 0.4) +
    geom_jitter(aes(shape = factor(replicate)), width = 0.1, size = 2) +
    scale_colour_manual(values = cbbPalette) +
    theme_ng() +
    ylab("log2 CPM") +
    ggtitle(feature)
}

for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = oe_results[[key]]
      if (is.null(fit)) next
      top6 = head(rownames(fit$tt), 6)
      plots = Filter(Negate(is.null),
                     lapply(top6, function(f) plot_cpm(fit$y, f, f)))
      if (length(plots) > 0) {
        cat("\nOE top 6 normalized counts -", al, fs, gr, "\n")
        print(wrap_plots(plots, ncol = 3))
      }
    }
  }
}
```

## Knockdown experiment

TDP-43 shRNA (tdp43_kd) vs non-targeting shRNA control (nt_kd_ctrl).

```{r kd-edger}
kd_results = list()
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      mat = counts_cache[[key]]
      if (is.null(mat)) next
      res = run_edger(mat, kd_meta, treat_level = "tdp43_kd", ref_level = "nt_kd_ctrl")
      if (!is.null(res)) kd_results[[key]] = res
    }
  }
}
save_rds(kd_results, "gse230647_kd_edger")
cat("KD edgeR fits completed:", length(kd_results), "\n")
```

```{r kd-pca, fig.width=10, fig.height=10}
pca_plots = list()
for (al in aligners) {
  for (fs in feature_sets) {
    key = paste(al, fs, "gene_id", sep = "_")
    fit = kd_results[[key]]
    if (is.null(fit)) next
    lcpm = cpm(fit$y, log = TRUE)
    pca = prcomp(t(lcpm), scale. = FALSE)
    pct = round(100 * pca$sdev^2 / sum(pca$sdev^2))
    df = data.frame(
      PC1 = pca$x[, 1],
      PC2 = pca$x[, 2],
      condition = fit$y$samples$group,
      sample = rownames(pca$x)
    )
    pca_plots[[key]] = ggplot(df, aes(PC1, PC2, colour = condition, label = sample)) +
      geom_point(size = 2) +
      geom_text_repel(size = 2.5, show.legend = FALSE) +
      xlab(paste0("PC1: ", pct[1], "%")) +
      ylab(paste0("PC2: ", pct[2], "%")) +
      scale_colour_manual(values = cbbPalette) +
      theme_ng() +
      ggtitle(paste(al, fs))
  }
}
if (length(pca_plots) > 0) wrap_plots(pca_plots, ncol = 3)
save_rds(pca_plots, "gse230647_kd_pca_plots")
```

```{r kd-volcano, fig.width=10, fig.height=10}
for (gr in granularities) {
  vol_plots = list()
  for (al in aligners) {
    for (fs in feature_sets) {
      key = paste(al, fs, gr, sep = "_")
      fit = kd_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      tt$sig = tt$FDR < 0.05 & abs(tt$logFC) > 1
      vol_plots[[key]] = ggplot(tt, aes(logFC, -log10(PValue), colour = sig)) +
        geom_point(size = 0.6, alpha = 0.6) +
        scale_colour_manual(values = c("grey70", cbbPalette[2]),
                            labels = c("ns", "FDR<0.05 & |lfc|>1")) +
        geom_vline(xintercept = c(-1, 1), linetype = "dashed", linewidth = 0.3) +
        theme_ng() +
        ggtitle(paste(al, fs, gr))
    }
  }
  if (length(vol_plots) > 0) print(wrap_plots(vol_plots, ncol = 3))
}
```

```{r kd-ma, fig.width=10, fig.height=10}
for (gr in granularities) {
  ma_plots = list()
  for (al in aligners) {
    for (fs in feature_sets) {
      key = paste(al, fs, gr, sep = "_")
      fit = kd_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      tt$sig = tt$FDR < 0.05
      ma_plots[[key]] = ggplot(tt, aes(logCPM, logFC, colour = sig)) +
        geom_point(size = 0.6, alpha = 0.6) +
        scale_colour_manual(values = c("grey70", cbbPalette[2]),
                            labels = c("ns", "FDR<0.05")) +
        geom_hline(yintercept = 0, linewidth = 0.3) +
        theme_ng() +
        ggtitle(paste(al, fs, gr))
    }
  }
  if (length(ma_plots) > 0) print(wrap_plots(ma_plots, ncol = 3))
}
```

```{r kd-top-hits}
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = kd_results[[key]]
      if (is.null(fit)) next
      top = fit$tt[fit$tt$FDR < 0.05, , drop = FALSE]
      if (nrow(top) == 0) next
      cat("\nKD top hits -", al, fs, gr, "\n")
      print(knitr::kable(head(top[, c("logFC", "logCPM", "F", "PValue", "FDR")], 20),
                         digits = 4))
    }
  }
}
```

```{r kd-heatmap, fig.width=10, fig.height=10}
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = kd_results[[key]]
      if (is.null(fit)) next
      tt = fit$tt
      sig_feats = rownames(tt)[tt$FDR < 0.05 & abs(tt$logFC) > 1]
      if (length(sig_feats) < 2) next
      sig_feats = head(sig_feats, 40)
      lcpm = cpm(fit$y, log = TRUE)[sig_feats, , drop = FALSE]
      ann = data.frame(condition = fit$y$samples$group,
                       row.names = colnames(lcpm))
      pheatmap(lcpm,
               annotation_col = ann,
               scale = "row",
               show_rownames = nrow(lcpm) <= 40,
               main = paste("KD", al, fs, gr),
               fontsize = 7,
               silent = TRUE)
    }
  }
}
```

```{r kd-counts, fig.width=10, fig.height=10}
for (al in aligners) {
  for (fs in feature_sets) {
    for (gr in granularities) {
      key = paste(al, fs, gr, sep = "_")
      fit = kd_results[[key]]
      if (is.null(fit)) next
      top6 = head(rownames(fit$tt), 6)
      plots = Filter(Negate(is.null),
                     lapply(top6, function(f) plot_cpm(fit$y, f, f)))
      if (length(plots) > 0) {
        cat("\nKD top 6 normalized counts -", al, fs, gr, "\n")
        print(wrap_plots(plots, ncol = 3))
      }
    }
  }
}
```

## Session info

```{r session-info}
sessionInfo()
```
