---
title: "GSE230647 single-cell repeat profiling: cluster 12 vs other cells (gene-normalized)"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  starsolo_base: "."
  kallisto_base: "."
  indices_base: "."
  metadata: "."
  sample_metadata: "."
  report_outdir: "."
  feature_sets: "genes genic_repeats intergenic_repeats"
  starsolo_modes: "unique"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(Matrix)
library(edgeR)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(tidyr)
library(readr)
library(pheatmap)

starsolo_base = params$starsolo_base
kallisto_base = params$kallisto_base
indices_base = params$indices_base
report_outdir = params$report_outdir

feature_sets = strsplit(params$feature_sets, " ")[[1]]
starsolo_modes = strsplit(params$starsolo_modes, " ")[[1]]
repeat_fsets = feature_sets[feature_sets != "genes"]

save_rds = function(obj, name) {
  path = file.path(report_outdir, paste0(name, ".rds"))
  saveRDS(obj, path)
  invisible(path)
}
```

# About this report

Repeat counts are normalized by gene library sizes (total gene read counts per
pseudo-bulk sample) so that fold changes reflect changes in the repeat-to-gene
expression ratio.

# Sample metadata

```{r sample-meta}
sample_meta = read_tsv(params$sample_metadata, show_col_types = FALSE)
sample_ids = sample_meta$sample_id
knitr::kable(sample_meta)
```

# Cell-type annotations from GEO

The GEO supplementary file maps each cell barcode to its cluster assignment.
Cluster 12 contains cells that express TDP-43-HA.

```{r load-geo-meta}
geo_meta = read_tsv(params$metadata, show_col_types = FALSE)
save_rds(geo_meta, "geo_meta_raw")

cat("columns:", paste(colnames(geo_meta), collapse = ", "), "\n")
cat("rows:", nrow(geo_meta), "\n")
```

```{r inspect-geo-meta}
head(geo_meta)
```

```{r cluster-column}
# Prefer v2 cluster assignment from the biorxiv reanalysis for both group
# assignment and plotting labels; fall back to any cluster column if absent.
cluster_num_col = if ("biorxiv_v2_cluster_number" %in% colnames(geo_meta)) {
  "biorxiv_v2_cluster_number"
} else {
  grep("cluster_number", colnames(geo_meta), ignore.case = TRUE, value = TRUE)[1]
}
cluster_name_col = if ("biorxiv_v2_cluster_name" %in% colnames(geo_meta)) {
  "biorxiv_v2_cluster_name"
} else {
  grep("cluster_name|cluster$", colnames(geo_meta), ignore.case = TRUE, value = TRUE)[1]
}
barcode_col = grep("barcode|cell|cb", colnames(geo_meta), ignore.case = TRUE, value = TRUE)[1]
if (is.na(cluster_num_col)) stop("no cluster number column found; available: ", paste(colnames(geo_meta), collapse = ", "))
if (is.na(cluster_name_col)) stop("no cluster name column found; available: ", paste(colnames(geo_meta), collapse = ", "))
if (is.na(barcode_col)) stop("no barcode column found; available: ", paste(colnames(geo_meta), collapse = ", "))
cat("using barcode column:", barcode_col, "\n")
cat("using cluster number column:", cluster_num_col, "\n")
cat("using cluster name column:", cluster_name_col, "\n")

geo_meta$cluster_num = as.integer(geo_meta[[cluster_num_col]])
geo_meta$cluster = geo_meta[[cluster_name_col]]

# GEO cell id format: <16bp_barcode>-<lane>.<sample_tag>, e.g. AAACCCACATGTCTAG-1.TDP2wON
# STARsolo raw barcodes are plain 16bp. Strip lane and sample tag to get the 16bp barcode,
# and keep the sample tag so we can match per-sample (16bp barcodes are not unique across samples).
geo_meta$barcode_16bp = sub("-[0-9]+\\..*$", "", geo_meta[[barcode_col]])
geo_meta$sample_tag = sub("^[^.]*\\.", "", geo_meta[[barcode_col]])

# Map sample_tag to sample_id using sample_meta (culture_weeks + replicate).
# Observed tags: TDP2wON (2w, no replicate), TDP4wONa (4w rep A), TDP4wONb (4w rep B).
rep_suffix = ifelse(is.na(sample_meta$replicate) | sample_meta$replicate == "",
                    "", tolower(sample_meta$replicate))
tag_to_sid = setNames(sample_meta$sample_id,
                      paste0("TDP", sample_meta$culture_weeks, "wON", rep_suffix))
geo_meta$sample_id = unname(tag_to_sid[geo_meta$sample_tag])

cat("geo sample_tag -> sample_id mapping:\n")
print(tag_to_sid)
cat("cells per sample_id in GEO metadata:\n")
print(table(geo_meta$sample_id, useNA = "ifany"))

table(geo_meta$cluster)
```

```{r cluster12-flag}
geo_meta$group = ifelse(!is.na(geo_meta$cluster_num) & geo_meta$cluster_num == 12L,
                        "cluster12", "other")
c12_labels = unique(geo_meta$cluster[geo_meta$group == "cluster12"])
cat("cluster 12 label(s):", paste(c12_labels, collapse = "; "), "\n")
cat("cluster 12 cells:", sum(geo_meta$group == "cluster12", na.rm = TRUE), "\n")
cat("other cells:", sum(geo_meta$group == "other", na.rm = TRUE), "\n")
save_rds(geo_meta, "geo_meta_annotated")
```

# Load STARsolo matrices and create pseudo-bulk counts

For each sample and each feature set (including genes), load the raw CB x feature
matrix from STARsolo, intersect barcodes with GEO metadata, and sum counts within
cluster 12 vs all other annotated cells.

```{r load-starsolo}
load_and_split = function(raw_dir, geo_meta_sample) {
  if (!dir.exists(raw_dir)) {
    warning("STARsolo raw directory not found: ", raw_dir)
    return(NULL)
  }
  barcodes = readLines(file.path(raw_dir, "barcodes.tsv"))
  features = read_tsv(file.path(raw_dir, "features.tsv"),
                      col_names = c("id", "name", "type"),
                      show_col_types = FALSE)
  mat = readMM(file.path(raw_dir, "matrix.mtx"))
  rownames(mat) = features$id
  colnames(mat) = barcodes

  matched_idx = match(barcodes, geo_meta_sample$barcode_16bp)
  matched = !is.na(matched_idx)
  cat("  barcodes in matrix:", length(barcodes),
      "| matched to GEO metadata:", sum(matched), "\n")

  if (sum(matched) == 0) {
    warning("no barcodes matched; check barcode format in GEO metadata")
    return(NULL)
  }

  group_matched = geo_meta_sample$group[matched_idx]
  idx12 = matched & !is.na(group_matched) & group_matched == "cluster12"
  idx_other = matched & !is.na(group_matched) & group_matched == "other"

  list(
    cluster12 = Matrix::rowSums(mat[, idx12, drop = FALSE]),
    other     = Matrix::rowSums(mat[, idx_other, drop = FALSE]),
    n12       = sum(idx12),
    n_other   = sum(idx_other)
  )
}

pb_list = list()

for (mode in starsolo_modes) {
  for (fs in feature_sets) {
    key = paste(mode, fs, sep = "_")
    counts12 = list()
    counts_other = list()

    for (sid in sample_ids) {
      raw_dir = file.path(starsolo_base, sid,
                          paste0(mode, "_", fs),
                          "Solo.out", "Gene", "raw")
      cat("loading", sid, key, "\n")
      geo_sample = geo_meta[!is.na(geo_meta$sample_id) & geo_meta$sample_id == sid, ]
      if (nrow(geo_sample) == 0) {
        cat("  no GEO annotations for sample", sid, "- skipping\n")
        next
      }
      res = load_and_split(raw_dir, geo_sample)
      if (!is.null(res)) {
        counts12[[sid]]    = res$cluster12
        counts_other[[sid]] = res$other
        cat("  cluster12:", res$n12, "cells | other:", res$n_other, "cells\n")
      }
    }

    if (length(counts12) > 0) {
      features_union = Reduce(union, lapply(counts12, names))
      mat12 = sapply(counts12, function(x) x[features_union])
      mat_other = sapply(counts_other, function(x) x[features_union])
      rownames(mat12) = features_union
      rownames(mat_other) = features_union
      mat12[is.na(mat12)] = 0
      mat_other[is.na(mat_other)] = 0
      pb_list[[key]] = list(cluster12 = mat12, other = mat_other)
    }
  }
}

save_rds(pb_list, "pseudobulk_counts")
```

# Gene library sizes for normalization

Gene pseudo-bulk library sizes are used to normalize repeat counts so that
fold changes reflect the repeat-to-gene expression ratio.

```{r gene-lib-sizes}
gene_lib_cache = list()
for (mode in starsolo_modes) {
  gene_key = paste(mode, "genes", sep = "_")
  if (gene_key %in% names(pb_list)) {
    gene_pb = pb_list[[gene_key]]
    gene_lib_cache[[mode]] = c(colSums(gene_pb$cluster12),
                               colSums(gene_pb$other))
    cat(mode, "gene lib sizes (cluster12 | other):\n")
    cat("  cluster12:", colSums(gene_pb$cluster12), "\n")
    cat("  other:", colSums(gene_pb$other), "\n")
  }
}
save_rds(gene_lib_cache, "gene_lib_sizes")
```

# edgeR: cluster 12 vs other cells (gene-normalized)

Pseudo-bulk: each sample contributes one column per group (cluster12 / other).
A paired design (sample as blocking factor) accounts for between-sample variation.
Library sizes are set to gene pseudo-bulk totals so that the test measures changes
in the repeat-to-gene expression ratio.

```{r edger-function}
run_edger = function(mat12, mat_other, label, gene_lib_sizes) {
  stopifnot(ncol(mat12) == ncol(mat_other))
  n_samples = ncol(mat12)
  counts = cbind(mat12, mat_other)
  group = factor(c(rep("cluster12", n_samples), rep("other", n_samples)),
                 levels = c("other", "cluster12"))
  sample_block = factor(rep(seq_len(n_samples), 2))

  design = model.matrix(~ sample_block + group)

  dge = DGEList(counts = counts, group = group)
  dge = dge[filterByExpr(dge, design), ]
  dge$samples$lib.size = gene_lib_sizes
  dge$samples$norm.factors = rep(1, ncol(dge))
  dge = estimateDisp(dge, design)

  fit = glmQLFit(dge, design)
  res = glmQLFTest(fit, coef = ncol(design))
  tt = topTags(res, n = Inf)$table
  tt$feature = rownames(tt)
  tt$feature_set = label
  tt
}
```

```{r run-edger}
edger_results = list()

for (mode in starsolo_modes) {
  gene_ls = gene_lib_cache[[mode]]
  if (is.null(gene_ls)) {
    cat("skipping mode", mode, "- no gene library sizes\n")
    next
  }
  for (fs in repeat_fsets) {
    key = paste(mode, fs, sep = "_")
    if (!key %in% names(pb_list)) next
    cat("edgeR:", key, "(gene-normalized)\n")
    pb = pb_list[[key]]
    tt = tryCatch(
      run_edger(pb$cluster12, pb$other, label = key,
                gene_lib_sizes = gene_ls),
      error = function(e) { message("edgeR failed for ", key, ": ", e$message); NULL }
    )
    if (!is.null(tt)) edger_results[[key]] = tt
  }
}

save_rds(edger_results, "edger_results")
```

# Results

```{r sig-summary}
for (key in names(edger_results)) {
  tt = edger_results[[key]]
  cat(key, ": FDR < 0.05 features:", sum(tt$FDR < 0.05, na.rm = TRUE),
      "/ total tested:", nrow(tt), "\n")
}
```

```{r volcano-plots, fig.width=7, fig.height=5}
for (key in names(edger_results)) {
  tt = edger_results[[key]]
  tt$sig = tt$FDR < 0.05
  p = ggplot(tt, aes(x = logFC, y = -log10(FDR), colour = sig)) +
    geom_point(size = 0.8, alpha = 0.6) +
    geom_text_repel(
      data = filter(tt, FDR < 0.01) %>% arrange(FDR) %>% head(20),
      aes(label = feature), size = 2.5, max.overlaps = 15
    ) +
    scale_colour_manual(values = c("FALSE" = "grey60", "TRUE" = "firebrick")) +
    labs(title = paste(key, "(gene-normalized)"),
         x = "log2 FC repeat/gene ratio (cluster 12 / other)",
         y = "-log10 FDR",
         colour = "FDR < 0.05") +
    theme_bw()
  print(p)
}
```

```{r top-tables}
for (key in names(edger_results)) {
  tt = edger_results[[key]]
  cat("\n###", key, "\n")
  top = tt %>% filter(FDR < 0.05) %>% arrange(FDR) %>% head(30)
  if (nrow(top) > 0) print(knitr::kable(top, digits = 4))
}
```

```{r heatmap-top, fig.width=8, fig.height=10}
for (mode in starsolo_modes) {
  gene_key = paste0(mode, "_genes")
  if (!gene_key %in% names(pb_list)) next
  gene_pb = pb_list[[gene_key]]
  gene_ls = c(colSums(gene_pb$cluster12), colSums(gene_pb$other))

  for (fs in repeat_fsets) {
    key = paste0(mode, "_", fs)
    if (!key %in% names(edger_results) || !key %in% names(pb_list)) next

    tt = edger_results[[key]]
    top_feats = tt %>% filter(FDR < 0.05) %>% arrange(FDR) %>% head(40) %>% pull(feature)
    if (length(top_feats) < 2) next

    pb = pb_list[[key]]
    mat = cbind(pb$cluster12, pb$other)
    colnames(mat) = c(paste0(colnames(pb$cluster12), "_c12"),
                      paste0(colnames(pb$other), "_other"))
    mat_sub = mat[top_feats, , drop = FALSE]

    cpm_mat = sweep(mat_sub, 2, gene_ls / 1e6, "/")
    cpm_log = log1p(cpm_mat)

    ann_col = data.frame(
      group = factor(c(rep("cluster12", ncol(pb$cluster12)),
                       rep("other", ncol(pb$other)))),
      row.names = colnames(cpm_log)
    )

    pheatmap(cpm_log,
             annotation_col = ann_col,
             cluster_rows = TRUE,
             cluster_cols = FALSE,
             show_rownames = TRUE,
             fontsize_row = 7,
             main = paste("top", fs, ":", key, "(gene-normalized CPM)"))
  }
}
```

# Per-cluster repeat expression dotplot (gene-normalized)

For the top genic repeats (by FDR from cluster 12 vs other), show expression
across all clusters as a dotplot. Each dot: mean gene-normalized log-CPM and
fraction of cells with at least one count. Gene library sizes are computed per
cluster to normalize the repeat signal.

```{r per-cluster-pseudobulk}
key_dot = paste0(starsolo_modes[1], "_genic_repeats")

dotplot_df = NULL

if (key_dot %in% names(pb_list) && key_dot %in% names(edger_results)) {
  tt = edger_results[[key_dot]]
  top_feats = tt %>% filter(FDR < 0.05) %>% arrange(FDR) %>% head(25) %>% pull(feature)

  cluster_counts = list()
  cluster_ncells = list()
  cluster_expr = list()
  gene_cluster_counts = list()

  for (sid in sample_ids) {
    raw_dir = file.path(starsolo_base, sid,
                        paste0(starsolo_modes[1], "_genic_repeats"),
                        "Solo.out", "Gene", "raw")
    gene_raw_dir = file.path(starsolo_base, sid,
                             paste0(starsolo_modes[1], "_genes"),
                             "Solo.out", "Gene", "raw")
    if (!dir.exists(raw_dir) || !dir.exists(gene_raw_dir)) next

    barcodes = readLines(file.path(raw_dir, "barcodes.tsv"))
    features = read_tsv(file.path(raw_dir, "features.tsv"),
                        col_names = c("id", "name", "type"),
                        show_col_types = FALSE)
    mat = readMM(file.path(raw_dir, "matrix.mtx"))
    rownames(mat) = features$id
    colnames(mat) = sub("-[0-9]+$", "", barcodes)

    gene_barcodes = readLines(file.path(gene_raw_dir, "barcodes.tsv"))
    gene_features = read_tsv(file.path(gene_raw_dir, "features.tsv"),
                             col_names = c("id", "name", "type"),
                             show_col_types = FALSE)
    gene_mat = readMM(file.path(gene_raw_dir, "matrix.mtx"))
    rownames(gene_mat) = gene_features$id
    colnames(gene_mat) = sub("-[0-9]+$", "", gene_barcodes)

    geo_sub = geo_meta %>%
      filter(!is.na(sample_id) & sample_id == sid) %>%
      mutate(cb = barcode_16bp) %>%
      filter(cb %in% colnames(mat))

    for (cl in unique(geo_sub$cluster)) {
      cbs = geo_sub$cb[geo_sub$cluster == cl]
      cbs_rep = intersect(cbs, colnames(mat))
      cbs_gene = intersect(cbs, colnames(gene_mat))
      if (length(cbs_rep) == 0 || length(cbs_gene) == 0) next
      key_cl = as.character(cl)
      m_sub = mat[top_feats, cbs_rep, drop = FALSE]
      prev_counts = if (is.null(cluster_counts[[key_cl]])) 0 else cluster_counts[[key_cl]]
      cluster_counts[[key_cl]] = prev_counts + Matrix::rowSums(m_sub)
      prev_ncells = if (is.null(cluster_ncells[[key_cl]])) 0 else cluster_ncells[[key_cl]]
      cluster_ncells[[key_cl]] = prev_ncells + length(cbs_rep)
      prev_expr = if (is.null(cluster_expr[[key_cl]])) 0 else cluster_expr[[key_cl]]
      cluster_expr[[key_cl]] = prev_expr + Matrix::rowSums(m_sub > 0)
      prev_gene = if (is.null(gene_cluster_counts[[key_cl]])) 0 else gene_cluster_counts[[key_cl]]
      gene_cluster_counts[[key_cl]] = prev_gene + sum(Matrix::colSums(gene_mat[, cbs_gene, drop = FALSE]))
    }
  }

  rows = list()
  for (cl in names(cluster_counts)) {
    n_cells = cluster_ncells[[cl]]
    cnts = cluster_counts[[cl]]
    gene_lib = gene_cluster_counts[[cl]]
    if (gene_lib == 0) gene_lib = 1
    cpm = cnts / (gene_lib / 1e6)
    rows[[cl]] = data.frame(
      cluster = cl,
      feature = names(cpm),
      mean_log_cpm = log1p(cpm),
      pct_expressed = cluster_expr[[cl]] / n_cells,
      stringsAsFactors = FALSE
    )
  }
  dotplot_df = bind_rows(rows)
  cluster_order_df = geo_meta %>%
    filter(!is.na(cluster_num)) %>%
    distinct(cluster, cluster_num) %>%
    arrange(cluster_num)
  cluster_levels = cluster_order_df$cluster[cluster_order_df$cluster %in% dotplot_df$cluster]
  dotplot_df$cluster = factor(dotplot_df$cluster, levels = cluster_levels)
  save_rds(dotplot_df, "dotplot_df")
}
```

```{r dotplot, fig.width=10, fig.height=7}
if (!is.null(dotplot_df)) {
  ggplot(dotplot_df, aes(x = cluster, y = feature, colour = mean_log_cpm,
                         size = pct_expressed)) +
    geom_point() +
    scale_colour_viridis_c(option = "plasma") +
    scale_size_continuous(range = c(0.5, 5)) +
    labs(title = "Genic repeat expression per cluster (gene-normalized)",
         subtitle = "top FDR-significant features, cluster 12 vs other",
         x = "cluster", y = NULL,
         colour = "mean log gene-norm CPM", size = "fraction expressed") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
          axis.text.y = element_text(size = 8))
}
```

# Repeat detection summary

Detection is defined as CPM > 1 in at least half the pseudo-bulk samples of a
group. CPM here uses raw repeat library sizes (not gene-normalized) to reflect
absolute detectability in the sequencing data.

```{r detection-helpers}
detect_features = function(mat, cpm_thr = 1, min_frac = 0.5) {
  if (is.null(mat) || ncol(mat) == 0) return(character(0))
  lib_sizes = colSums(mat)
  lib_sizes[lib_sizes == 0] = 1
  cpm_mat = sweep(mat, 2, lib_sizes / 1e6, "/")
  n_above = rowSums(cpm_mat > cpm_thr)
  names(n_above)[n_above >= ceiling(min_frac * ncol(mat))]
}

load_class_map = function(indices_base, feature_set) {
  path = file.path(indices_base, paste0(feature_set, "_locus_map.tsv"))
  if (!file.exists(path)) return(NULL)
  lm = read_tsv(path,
                 col_names = c("transcript_id", "gene_id", "family_id", "class_id"),
                 show_col_types = FALSE)
  dedup = lm %>% select(gene_id, class_id) %>% distinct()
  setNames(dedup$class_id, dedup$gene_id)
}
```

```{r detection-data}
class_maps = list()
for (fs in repeat_fsets) {
  class_maps[[fs]] = load_class_map(indices_base, fs)
}

det_summary = list()
det_class = list()

for (mode in starsolo_modes) {
  for (fs in repeat_fsets) {
    key = paste(mode, fs, sep = "_")
    if (!key %in% names(pb_list)) next
    pb = pb_list[[key]]
    cm = class_maps[[fs]]
    for (grp in c("cluster12", "other")) {
      grp_mat = pb[[grp]]
      detected = detect_features(grp_mat)
      det_summary[[length(det_summary) + 1]] = data.frame(
        mode = mode, feature_set = fs, group = grp,
        n_detected = length(detected), stringsAsFactors = FALSE)
      if (!is.null(cm) && length(detected) > 0) {
        classes = cm[detected]
        classes[is.na(classes)] = "unknown"
        class_tab = as.data.frame(table(classes), stringsAsFactors = FALSE)
        colnames(class_tab) = c("class_id", "n")
        class_tab$mode = mode
        class_tab$feature_set = fs
        class_tab$group = grp
        det_class[[length(det_class) + 1]] = class_tab
      }
    }
  }
}

det_summary_df = bind_rows(det_summary)
det_class_df = bind_rows(det_class)

if (nrow(det_class_df) > 0) {
  top_classes = det_class_df %>%
    group_by(class_id) %>%
    summarize(total = sum(n), .groups = "drop") %>%
    arrange(desc(total)) %>%
    head(7) %>%
    pull(class_id)
  det_class_df$class_label = ifelse(det_class_df$class_id %in% top_classes,
                                    det_class_df$class_id, "other")
  det_class_df = det_class_df %>%
    group_by(mode, feature_set, group, class_label) %>%
    summarize(n = sum(n), .groups = "drop")
}

save_rds(det_summary_df, "sc_detection_summary")
save_rds(det_class_df, "sc_detection_by_class")
```

## Features detected per group

```{r detection-barplot, fig.width=8, fig.height=5}
if (nrow(det_summary_df) > 0) {
  ggplot(det_summary_df, aes(x = group, y = n_detected, fill = feature_set)) +
    geom_col(position = "dodge") +
    facet_wrap(~mode) +
    scale_fill_brewer(palette = "Set2") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(y = "detected features (CPM > 1)", fill = "feature set",
         title = "Repeat features detected per group (pseudo-bulk)")
}
```

## Features detected per group by repeat class

```{r detection-class-barplot, fig.width=10, fig.height=7}
if (nrow(det_class_df) > 0) {
  ggplot(det_class_df, aes(x = group, y = n, fill = class_label)) +
    geom_col() +
    facet_grid(feature_set ~ mode) +
    scale_fill_brewer(palette = "Set2") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(y = "detected features (CPM > 1)", fill = "repeat class",
         title = "Detected repeats by class per group (pseudo-bulk)")
}
```

```{r session}
sessionInfo()
```
