---
title: "GSE230647 single-cell repeat profiling: cluster 12 vs other cells"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  starsolo_base: "."
  kallisto_base: "."
  indices_base: "."
  metadata: "."
  sample_metadata: "."
  report_outdir: "."
  feature_sets: "genes genic_repeats intergenic_repeats"
  starsolo_modes: "unique"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(Matrix)
library(edgeR)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(tidyr)
library(readr)
library(pheatmap)

starsolo_base = params$starsolo_base
kallisto_base = params$kallisto_base
indices_base = params$indices_base
report_outdir = params$report_outdir

feature_sets = strsplit(params$feature_sets, " ")[[1]]
starsolo_modes = strsplit(params$starsolo_modes, " ")[[1]]
repeat_fsets = feature_sets[feature_sets != "genes"]

save_rds = function(obj, name) {
  path = file.path(report_outdir, paste0(name, ".rds"))
  saveRDS(obj, path)
  invisible(path)
}
```

# Sample metadata

```{r sample-meta}
sample_meta = read_tsv(params$sample_metadata, show_col_types = FALSE)
sample_ids = sample_meta$sample_id
knitr::kable(sample_meta)
```

# Cell-type annotations from GEO

The GEO supplementary file maps each cell barcode to its cluster assignment.
Cluster 12 contains cells that express TDP-43-HA.

```{r load-geo-meta}
geo_meta = read_tsv(params$metadata, show_col_types = FALSE)
save_rds(geo_meta, "geo_meta_raw")

cat("columns:", paste(colnames(geo_meta), collapse = ", "), "\n")
cat("rows:", nrow(geo_meta), "\n")
```

```{r inspect-geo-meta}
# print first rows to confirm column names before hardcoding
head(geo_meta)
```

```{r cluster-column}
cluster_col = grep("cluster", colnames(geo_meta), ignore.case = TRUE, value = TRUE)[1]
barcode_col = grep("barcode|cell|cb", colnames(geo_meta), ignore.case = TRUE, value = TRUE)[1]
if (is.na(cluster_col)) stop("no cluster column found; available: ", paste(colnames(geo_meta), collapse = ", "))
if (is.na(barcode_col)) stop("no barcode column found; available: ", paste(colnames(geo_meta), collapse = ", "))
cat("using barcode column:", barcode_col, "\n")
cat("using cluster column:", cluster_col, "\n")

geo_meta$cluster = geo_meta[[cluster_col]]
geo_meta$barcode_raw = geo_meta[[barcode_col]]

table(geo_meta$cluster)
```

```{r cluster12-flag}
geo_meta$group = ifelse(as.integer(geo_meta$cluster) == 12L, "cluster12", "other")
cat("cluster 12 cells:", sum(geo_meta$group == "cluster12"), "\n")
cat("other cells:", sum(geo_meta$group == "other"), "\n")
save_rds(geo_meta, "geo_meta_annotated")
```

# Load STARsolo matrices and create pseudo-bulk counts

For each sample and each repeat feature set, load the raw CB x feature
matrix from STARsolo, intersect barcodes with GEO metadata, and sum
counts within cluster 12 vs all other annotated cells.

```{r load-starsolo}
# returns a named list: cluster12 and other, each a numeric vector of feature sums
load_and_split = function(raw_dir, geo_meta) {
  if (!dir.exists(raw_dir)) {
    warning("STARsolo raw directory not found: ", raw_dir)
    return(NULL)
  }
  barcodes = readLines(file.path(raw_dir, "barcodes.tsv"))
  features = read_tsv(file.path(raw_dir, "features.tsv"),
                      col_names = c("id", "name", "type"),
                      show_col_types = FALSE)
  mat = readMM(file.path(raw_dir, "matrix.mtx"))
  # STARsolo: rows = features, cols = barcodes
  rownames(mat) = features$id
  colnames(mat) = barcodes

  # strip any suffix like -1 from STARsolo barcodes for matching
  cb_stripped = sub("-[0-9]+$", "", barcodes)
  geo_stripped = sub("-[0-9]+$", "", geo_meta$barcode_raw)

  matched = cb_stripped %in% geo_stripped
  cat("  barcodes in matrix:", length(barcodes),
      "| matched to GEO metadata:", sum(matched), "\n")

  if (sum(matched) == 0) {
    warning("no barcodes matched; check barcode format in GEO metadata")
    return(NULL)
  }

  idx12 = matched & (geo_meta$group[match(cb_stripped, geo_stripped)] == "cluster12")
  idx_other = matched & (geo_meta$group[match(cb_stripped, geo_stripped)] == "other")
  idx12[is.na(idx12)] = FALSE
  idx_other[is.na(idx_other)] = FALSE

  list(
    cluster12 = Matrix::rowSums(mat[, idx12, drop = FALSE]),
    other     = Matrix::rowSums(mat[, idx_other, drop = FALSE]),
    n12       = sum(idx12),
    n_other   = sum(idx_other)
  )
}

# collect pseudo-bulk counts across samples x feature_sets x starsolo_modes
pb_list = list()

for (mode in starsolo_modes) {
  for (fs in feature_sets) {
    key = paste(mode, fs, sep = "_")
    counts12 = list()
    counts_other = list()

    for (sid in sample_ids) {
      raw_dir = file.path(starsolo_base, sid,
                          paste0(mode, "_", fs),
                          "Solo.out", "Gene", "raw")
      cat("loading", sid, key, "\n")
      res = load_and_split(raw_dir, geo_meta)
      if (!is.null(res)) {
        counts12[[sid]]    = res$cluster12
        counts_other[[sid]] = res$other
        cat("  cluster12:", res$n12, "cells | other:", res$n_other, "cells\n")
      }
    }

    if (length(counts12) > 0) {
      features_union = Reduce(union, lapply(counts12, names))
      mat12 = sapply(counts12, function(x) x[features_union])
      mat_other = sapply(counts_other, function(x) x[features_union])
      rownames(mat12) = features_union
      rownames(mat_other) = features_union
      mat12[is.na(mat12)] = 0
      mat_other[is.na(mat_other)] = 0
      pb_list[[key]] = list(cluster12 = mat12, other = mat_other)
    }
  }
}

save_rds(pb_list, "pseudobulk_counts")
```

# edgeR: cluster 12 vs other cells

Pseudo-bulk: each sample contributes one column per group (cluster12 / other).
A paired design (sample as blocking factor) accounts for between-sample variation.

```{r edger-function}
run_edger = function(mat12, mat_other, label) {
  stopifnot(ncol(mat12) == ncol(mat_other))
  n_samples = ncol(mat12)
  counts = cbind(mat12, mat_other)
  group = factor(c(rep("cluster12", n_samples), rep("other", n_samples)),
                 levels = c("other", "cluster12"))
  sample_block = factor(rep(seq_len(n_samples), 2))

  design = model.matrix(~ sample_block + group)

  dge = DGEList(counts = counts, group = group)
  dge = dge[filterByExpr(dge, design), ]
  dge = calcNormFactors(dge)
  dge = estimateDisp(dge, design)

  fit = glmQLFit(dge, design)
  res = glmQLFTest(fit, coef = ncol(design))
  tt = topTags(res, n = Inf)$table
  tt$feature = rownames(tt)
  tt$feature_set = label
  tt
}
```

```{r run-edger}
edger_results = list()

for (key in names(pb_list)) {
  cat("edgeR:", key, "\n")
  pb = pb_list[[key]]
  tt = tryCatch(
    run_edger(pb$cluster12, pb$other, label = key),
    error = function(e) { message("edgeR failed for ", key, ": ", e$message); NULL }
  )
  if (!is.null(tt)) edger_results[[key]] = tt
}

save_rds(edger_results, "edger_results")
```

# Results

```{r sig-summary}
for (key in names(edger_results)) {
  tt = edger_results[[key]]
  cat(key, ": FDR < 0.05 features:", sum(tt$FDR < 0.05, na.rm = TRUE),
      "/ total tested:", nrow(tt), "\n")
}
```

```{r volcano-plots, fig.width=7, fig.height=5}
for (key in names(edger_results)) {
  tt = edger_results[[key]]
  tt$sig = tt$FDR < 0.05
  p = ggplot(tt, aes(x = logFC, y = -log10(FDR), colour = sig)) +
    geom_point(size = 0.8, alpha = 0.6) +
    geom_text_repel(
      data = filter(tt, FDR < 0.01) %>% arrange(FDR) %>% head(20),
      aes(label = feature), size = 2.5, max.overlaps = 15
    ) +
    scale_colour_manual(values = c("FALSE" = "grey60", "TRUE" = "firebrick")) +
    labs(title = key,
         x = "log2 fold change (cluster 12 / other)",
         y = "-log10 FDR",
         colour = "FDR < 0.05") +
    theme_bw()
  print(p)
}
```

```{r top-tables}
for (key in names(edger_results)) {
  tt = edger_results[[key]]
  cat("\n###", key, "\n")
  top = tt %>% filter(FDR < 0.05) %>% arrange(FDR) %>% head(30)
  if (nrow(top) > 0) print(knitr::kable(top, digits = 4))
}
```

```{r heatmap-top, fig.width=8, fig.height=10}
# heatmap of top 40 FDR-significant features for the unique_genic_repeats result
key_heat = paste0(starsolo_modes[1], "_genic_repeats")
if (key_heat %in% names(edger_results) && key_heat %in% names(pb_list)) {
  tt = edger_results[[key_heat]]
  top_feats = tt %>% filter(FDR < 0.05) %>% arrange(FDR) %>% head(40) %>% pull(feature)

  if (length(top_feats) > 1) {
    pb = pb_list[[key_heat]]
    mat = cbind(pb$cluster12, pb$other)
    colnames(mat) = c(paste0(sample_ids, "_c12"), paste0(sample_ids, "_other"))
    mat_sub = mat[top_feats, , drop = FALSE]

    # CPM for display
    lib_sizes = colSums(mat)
    cpm_mat = sweep(mat_sub, 2, lib_sizes / 1e6, "/")
    cpm_log = log1p(cpm_mat)

    ann_col = data.frame(
      group = factor(c(rep("cluster12", length(sample_ids)),
                       rep("other", length(sample_ids)))),
      row.names = colnames(cpm_log)
    )

    pheatmap(cpm_log,
             annotation_col = ann_col,
             cluster_rows = TRUE,
             cluster_cols = FALSE,
             show_rownames = TRUE,
             fontsize_row = 7,
             main = paste("top genic repeats:", key_heat))
  }
}
```

# Per-cluster repeat expression dotplot

Extra analysis: for the top genic repeats (by FDR from cluster 12 vs other),
show expression across all 17 clusters as a dotplot. Each dot: mean log-CPM
and fraction of cells with at least one count.

```{r per-cluster-pseudobulk}
key_dot = paste0(starsolo_modes[1], "_genic_repeats")

dotplot_df = NULL

if (key_dot %in% names(pb_list) && key_dot %in% names(edger_results)) {
  tt = edger_results[[key_dot]]
  top_feats = tt %>% filter(FDR < 0.05) %>% arrange(FDR) %>% head(25) %>% pull(feature)

  # build per-cluster pseudo-bulk across all samples
  cluster_counts = list()
  cluster_ncells = list()
  cluster_expr = list()

  for (sid in sample_ids) {
    raw_dir = file.path(starsolo_base, sid,
                        paste0(starsolo_modes[1], "_genic_repeats"),
                        "Solo.out", "Gene", "raw")
    if (!dir.exists(raw_dir)) next

    barcodes = readLines(file.path(raw_dir, "barcodes.tsv"))
    features = read_tsv(file.path(raw_dir, "features.tsv"),
                        col_names = c("id", "name", "type"),
                        show_col_types = FALSE)
    mat = readMM(file.path(raw_dir, "matrix.mtx"))
    rownames(mat) = features$id
    colnames(mat) = sub("-[0-9]+$", "", barcodes)

    geo_sub = geo_meta %>%
      mutate(cb = sub("-[0-9]+$", "", barcode_raw)) %>%
      filter(cb %in% colnames(mat))

    for (cl in unique(geo_sub$cluster)) {
      cbs = geo_sub$cb[geo_sub$cluster == cl]
      cbs = intersect(cbs, colnames(mat))
      if (length(cbs) == 0) next
      key_cl = as.character(cl)
      m_sub = mat[top_feats, cbs, drop = FALSE]
      prev_counts = if (is.null(cluster_counts[[key_cl]])) 0 else cluster_counts[[key_cl]]
      cluster_counts[[key_cl]] = prev_counts + Matrix::rowSums(m_sub)
      prev_ncells = if (is.null(cluster_ncells[[key_cl]])) 0 else cluster_ncells[[key_cl]]
      cluster_ncells[[key_cl]] = prev_ncells + length(cbs)
      prev_expr = if (is.null(cluster_expr[[key_cl]])) 0 else cluster_expr[[key_cl]]
      cluster_expr[[key_cl]] = prev_expr + Matrix::rowSums(m_sub > 0)
    }
  }

  # build dotplot data frame
  rows = list()
  for (cl in names(cluster_counts)) {
    n_cells = cluster_ncells[[cl]]
    cnts = cluster_counts[[cl]]
    lib = sum(cnts)
    cpm = cnts / (lib / 1e6)
    rows[[cl]] = data.frame(
      cluster = cl,
      feature = names(cpm),
      mean_log_cpm = log1p(cpm),
      pct_expressed = cluster_expr[[cl]] / n_cells,
      stringsAsFactors = FALSE
    )
  }
  dotplot_df = bind_rows(rows)
  dotplot_df$cluster = factor(dotplot_df$cluster,
                              levels = sort(unique(as.integer(dotplot_df$cluster))))
  save_rds(dotplot_df, "dotplot_df")
}
```

```{r dotplot, fig.width=10, fig.height=7}
if (!is.null(dotplot_df)) {
  ggplot(dotplot_df, aes(x = cluster, y = feature, colour = mean_log_cpm,
                         size = pct_expressed)) +
    geom_point() +
    scale_colour_viridis_c(option = "plasma") +
    scale_size_continuous(range = c(0.5, 5)) +
    labs(title = "Genic repeat expression per cluster",
         subtitle = "top FDR-significant features, cluster 12 vs other",
         x = "cluster", y = NULL,
         colour = "mean log-CPM", size = "fraction expressed") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 0),
          axis.text.y = element_text(size = 8))
}
```

```{r session}
sessionInfo()
```
