---
title: "repeat element quantification - noise sweep report"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  eval_dirs: "."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 5)
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(scales)
library(patchwork)
source(file.path(dirname(knitr::current_input(full.names = TRUE)), "ggtheme.R"))

make_palette <- function(n) {
  if (n <= 8)  return(cbbPalette[seq_len(n)])
  colorRampPalette(cbbPalette)(n)
}

eval_dirs <- strsplit(params$eval_dirs, ",")[[1]]
eval_dirs <- trimws(eval_dirs)
```

```{r load-data}
load_metrics_from_dirs <- function(dirs, filename) {
  purrr::map_dfr(dirs, function(d) {
    p <- file.path(d, filename)
    if (!file.exists(p)) return(NULL)
    read_tsv(p, show_col_types = FALSE)
  })
}

load_pattern_from_dirs <- function(dirs, pattern) {
  purrr::map_dfr(dirs, function(d) {
    files <- list.files(d, pattern = pattern, full.names = TRUE)
    if (length(files) == 0) return(NULL)
    purrr::map_dfr(files, ~ read_tsv(.x, show_col_types = FALSE))
  })
}

global <- load_metrics_from_dirs(eval_dirs, "summary_global_metrics.tsv")

num_cols <- c("pearson_r", "spearman_r", "log1p_rmse",
              "precision", "recall", "f1", "jaccard", "specificity",
              "n_cells", "n_features", "mutation_rate",
              "wall_time_s", "cpu_time_s", "max_rss_mb", "io_in_mb", "io_out_mb")
for (col in num_cols) {
  if (col %in% names(global))
    global[[col]] <- suppressWarnings(as.numeric(global[[col]]))
}

if (!"granularity" %in% names(global))   global$granularity   <- "gene_id"
if (!"feature_set" %in% names(global))   global$feature_set   <- "repeats"
if (!"multimapper_mode" %in% names(global)) global$multimapper_mode <- "unique"
if (!"mutation_rate" %in% names(global)) global$mutation_rate <- NA_real_

global <- global %>%
  mutate(
    aligner_mode  = paste0(aligner, " (", multimapper_mode, ")"),
    gran_label    = factor(granularity,
                           levels = c("locus", "gene_id", "family_id", "class_id")),
    noise_label   = paste0(mutation_rate * 100, "%")
  )

per_cell   <- tryCatch(load_pattern_from_dirs(eval_dirs, "_per_cell_metrics\\.tsv$"),
                       error = function(e) NULL)
per_family <- tryCatch(load_pattern_from_dirs(eval_dirs, "_per_family_metrics\\.tsv$"),
                       error = function(e) NULL)

for (col in c("pearson_r", "spearman_r", "mutation_rate")) {
  if (!is.null(per_cell) && col %in% names(per_cell))
    per_cell[[col]] <- suppressWarnings(as.numeric(per_cell[[col]]))
  if (!is.null(per_family) && col %in% names(per_family))
    per_family[[col]] <- suppressWarnings(as.numeric(per_family[[col]]))
}

if (!is.null(per_family) && nrow(per_family) > 0) {
  for (col in c("f1", "precision", "recall", "jaccard", "specificity")) {
    if (col %in% names(per_family))
      per_family[[col]] <- suppressWarnings(as.numeric(per_family[[col]]))
  }
  if (!"multimapper_mode" %in% names(per_family)) per_family$multimapper_mode <- "unique"
  per_family <- per_family %>%
    mutate(aligner_mode = paste0(aligner, " (", multimapper_mode, ")"))
}
```

---

This report combines evaluation results across multiple noise levels (simulated nucleotide
substitution rates). Each `eval_dirs` entry corresponds to one pipeline run with a distinct
`mutation_rate` in the simulation config. The `mutation_rate` column in the metrics TSVs
identifies each run.

Noise levels found in the data: `r sort(unique(global$mutation_rate))`

---

## metric degradation by noise level

How each aligner's accuracy changes as the per-base mutation rate increases.
x-axis = mutation rate, y-axis = metric value, colour = aligner (+ multimapper mode).

```{r degradation-main, fig.height=15, fig.width=12}
acc_metrics <- c("pearson_r", "spearman_r", "f1", "precision", "recall",
                 "jaccard", "specificity")

deg <- global %>%
  filter(!is.na(mutation_rate)) %>%
  select(aligner_mode, feature_set, gran_label, mutation_rate, all_of(acc_metrics)) %>%
  pivot_longer(all_of(acc_metrics), names_to = "metric", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(metric = factor(metric, levels = acc_metrics))

n_al <- length(unique(deg$aligner_mode))
pal  <- make_palette(n_al)

ggplot(deg, aes(x = mutation_rate, y = value,
                colour = aligner_mode, group = aligner_mode)) +
  geom_point(size = 2.5, alpha = 0.85) +
  geom_line(linewidth = 0.6) +
  facet_grid(metric ~ feature_set + gran_label, scales = "free_y") +
  scale_colour_manual(values = pal, name = "aligner (mode)") +
  scale_x_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(x = "mutation rate", y = NULL,
       title = "metric degradation with increasing noise",
       subtitle = "rows = metric; columns = feature_set x granularity; x = mutation rate") +
  theme_ng() +
  theme(axis.text.x  = element_text(angle = 40, hjust = 1, size = 7),
        strip.text   = element_text(size = 7),
        legend.position = "bottom",
        legend.text  = element_text(size = 8))
```

## pearson r and F1 — noise overview

Summary view focusing on the two most interpretable metrics.

```{r overview-two-metrics, fig.height=5, fig.width=7}
ov <- global %>%
  filter(!is.na(mutation_rate)) %>%
  select(aligner_mode, feature_set, gran_label, mutation_rate, pearson_r, f1) %>%
  pivot_longer(c(pearson_r, f1), names_to = "metric", values_to = "value") %>%
  filter(!is.na(value))

n_al2 <- length(unique(ov$aligner_mode))
pal2  <- make_palette(n_al2)

ggplot(ov, aes(x = mutation_rate, y = value,
               colour = aligner_mode, group = aligner_mode)) +
  geom_point(size = 3, alpha = 0.85) +
  geom_line(linewidth = 0.8) +
  facet_grid(metric ~ feature_set, scales = "free_y") +
  scale_colour_manual(values = pal2, name = "aligner (mode)") +
  scale_x_continuous(labels = percent_format(accuracy = 0.1)) +
  ylim(0, 1) +
  labs(x = "mutation rate", y = NULL,
       title = "Pearson r and F1 across noise levels",
       subtitle = "rows = metric, columns = feature_set; lines connect same aligner across mutation rates") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1),
        legend.position = "bottom")
```

## global accuracy metrics per noise level

Full metric grid from the standard report, faceted by mutation rate.

```{r global-faceted, fig.height=18, fig.width=14}
full_metrics <- c("pearson_r", "spearman_r", "f1", "precision", "recall",
                  "jaccard", "specificity", "log1p_rmse")

full_long <- global %>%
  filter(!is.na(mutation_rate)) %>%
  select(aligner_mode, feature_set, gran_label, mutation_rate, noise_label,
         all_of(full_metrics)) %>%
  pivot_longer(all_of(full_metrics), names_to = "metric", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(metric = factor(metric, levels = full_metrics))

n_al3 <- length(unique(full_long$aligner_mode))
pal3  <- make_palette(n_al3)

ggplot(full_long, aes(x = gran_label, y = value,
                      colour = aligner_mode, group = aligner_mode)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_line(linetype = "dashed", linewidth = 0.5) +
  facet_grid(metric ~ noise_label + feature_set, scales = "free_y") +
  scale_colour_manual(values = pal3, name = "aligner (mode)") +
  labs(x = "granularity", y = NULL,
       title = "global accuracy metrics",
       subtitle = "rows = metric; columns = mutation rate x feature_set; x = granularity") +
  theme_ng() +
  theme(axis.text.x  = element_text(angle = 40, hjust = 1, size = 7),
        strip.text   = element_text(size = 6),
        legend.position = "bottom",
        legend.text  = element_text(size = 7))
```

## per-cell correlations by noise level

```{r per-cell-noise, fig.height=10, fig.width=14}
if (!is.null(per_cell) && nrow(per_cell) > 0 && "mutation_rate" %in% names(per_cell)) {
  pc <- per_cell %>%
    mutate(across(c(pearson_r, spearman_r), ~ suppressWarnings(as.numeric(.)))) %>%
    filter(!is.na(aligner), !is.na(mutation_rate))

  if (!"multimapper_mode" %in% names(pc)) pc$multimapper_mode <- "unique"
  if (!"feature_set" %in% names(pc))      pc$feature_set      <- "repeats"

  pc <- pc %>%
    mutate(aligner_mode = paste0(aligner, " (", multimapper_mode, ")"),
           gran_label   = factor(granularity,
                                 levels = c("locus", "gene_id", "family_id", "class_id")),
           noise_label  = paste0(mutation_rate * 100, "%"))

  pc_long <- pc %>%
    pivot_longer(c(pearson_r, spearman_r), names_to = "metric", values_to = "r") %>%
    filter(!is.na(r))

  n_al4 <- length(unique(pc_long$aligner_mode))
  pal4  <- make_palette(n_al4)

  ggplot(pc_long, aes(x = aligner_mode, y = r, fill = aligner_mode)) +
    geom_violin(alpha = 0.7, draw_quantiles = c(0.25, 0.5, 0.75)) +
    geom_jitter(width = 0.1, size = 0.5, alpha = 0.4) +
    facet_grid(metric + gran_label ~ noise_label + feature_set) +
    scale_fill_manual(values = pal4) +
    labs(x = NULL, y = "correlation",
         title = "per-cell correlations by noise level",
         subtitle = "rows = metric x granularity; columns = mutation rate x feature_set") +
    theme_ng() +
    theme(axis.text.x  = element_text(angle = 45, hjust = 1, size = 6),
          legend.position = "none",
          strip.text   = element_text(size = 6))
} else {
  cat("No per-cell metrics with mutation_rate column found.\n")
}
```

## per-repeat-class F1 by noise level

```{r per-family-noise, fig.height=14, fig.width=14}
if (!is.null(per_family) && nrow(per_family) > 0 &&
    "class_id" %in% names(per_family) && "mutation_rate" %in% names(per_family)) {

  pf <- per_family %>%
    filter(!is.na(class_id), class_id != "unknown", !is.na(mutation_rate)) %>%
    mutate(gran_label  = factor(granularity,
                                levels = c("locus", "gene_id", "family_id", "class_id")),
           noise_label = paste0(mutation_rate * 100, "%"))

  n_al5 <- length(unique(pf$aligner_mode))
  pal5  <- make_palette(n_al5)

  p_f1 <- ggplot(pf, aes(x = class_id, y = f1,
                          colour = aligner_mode, group = aligner_mode)) +
    geom_line(linewidth = 0.4, linetype = "dashed") +
    geom_point(size = 1.5, alpha = 0.85) +
    facet_grid(gran_label ~ noise_label + feature_set) +
    scale_y_sqrt() +
    scale_colour_manual(values = pal5, name = "Aligner (mode)") +
    labs(x = "repeat class", y = "F1",
         title = "F1 per repeat class by noise level",
         subtitle = "rows = granularity; columns = mutation rate x feature_set") +
    theme_ng() +
    theme(axis.text.x  = element_text(angle = 40, hjust = 1, size = 7),
          legend.position = "bottom",
          legend.text  = element_text(size = 7),
          strip.text   = element_text(size = 6))

  p_r <- ggplot(pf, aes(x = class_id, y = pearson_r,
                         colour = aligner_mode, group = aligner_mode)) +
    geom_line(linewidth = 0.4, linetype = "dashed") +
    geom_point(size = 1.5, alpha = 0.85) +
    facet_grid(gran_label ~ noise_label + feature_set) +
    scale_y_sqrt() +
    scale_colour_manual(values = pal5, name = "Aligner (mode)") +
    labs(x = "repeat class", y = "pearson r") +
    theme_ng() +
    theme(axis.text.x  = element_text(angle = 40, hjust = 1, size = 7),
          legend.position = "none",
          strip.text   = element_text(size = 6))

  p_f1 / p_r
} else {
  cat("No per-family metrics with class_id and mutation_rate found.\n")
}
```

---
*Generated by [noise_sweep_report.Rmd](../workflow/scripts/noise_sweep_report.Rmd)*
