# Dimred PCA comparing modalities and SCINA {.tabset}

Colour by SCINA

```{r, fig.width = 5.1, fig.height = 5.1, results = 'asis'}

for (flavour1 in names(d)) {

    for (flavour2 in setdiff(names(d), flavour1)) {
        cat('## ', flavour1, flavour2, '\n\n')
    
        common <- intersect(rownames(Embeddings(d[[flavour1]])),
                            rownames(Embeddings(d[[flavour2]])))

        palette <- brewer.pal(n = nlevels(d[[flavour1]]@meta.data$scina), name = "Paired")

        for (i in 1:5) {
            for (j in 1:5) {
                par(mar=c(5.1, 4.1, 7.1, 2.1))

                plot(x = Embeddings(d[[flavour1]], reduction = "pca")[common,i],
                     y = Embeddings(d[[flavour2]], reduction = "pca")[common,j],
                     col = pal[d[[flavour1]]@meta.data[common, 'scina'],'color'],
                     pch = 19, cex = 0.5,
                     xlab = sprintf('%s PC %s', flavour1, i),
                     ylab = sprintf('%s PC %s', flavour2, j))

                    legend('top',
                           legend = pal$identity,
                           col = pal$color,
                           pch=19, cex=0.8,
                           inset=c(0,-0.42),
                           ncol = 2,
                           xpd = TRUE)
            }
        }

        cat('\n\n')
    }
}


```


# Pairwise scatterplots colored by clustering {.tabset}

Tabset: coloring according to the clustering results.


```{r}
## retrieves the modality that measured genes
tag_genes <- grep('^genes', names(d), value = TRUE)[1]

```

```{r, results = 'asis', fig.width =8, fig.height = 8}

res <- 'RNA_snn_res.0.2'
common_cells <- Reduce(intersect, sapply(d, colnames))

for (item_clust in names(d)) {
    tag <- sprintf('%s\n%s', item_clust, res)
    cat('## ', tag, '\n\n')
    
    merged <- data.frame(cell = common_cells,
                         percent.mt = d[[tag_genes]]@meta.data[common_cells, 'percent.mt'])

    merged[,tag] <- d[[item_clust]]@meta.data[common_cells,res]

    for (item in names(d)) {
        d[[item]]@meta.data$cell <- rownames(d[[item]]@meta.data)

        merged <- merge(merged,
                        d[[item]]@meta.data[c('cell', 'nCount_RNA', 'nFeature_RNA')],
                        by = 'cell',
                        all.x = TRUE,
                        suffixes = c('', sprintf('\n%s', item)))

    }
    
    rownames(merged) <- merged$cell
    merged <- merged[,-1]
    colnames(merged) <- gsub('_RNA', '', colnames(merged))
    stop(colnames(merged))
    plot(merged, pch = 20, cex = 0.5, col = ac(as.numeric(merged[,tag]), 0.1))
    cat('\n\n')
}

rm(res)
rm(merged)
```

#  Pairwise scatterplots colored by cell identiy {.tabset} 

Tabset: coloring according to the clustering results.

```{r, results = 'asis', fig.width =8, fig.height = 8}

tag <- 'scina'

merged <- data.frame(cell = common_cells,
                     percent.mt = d[[tag_genes]]@meta.data[common_cells, 'percent.mt'])

## merged[,tag] <- d[[item_clust]]@meta.data[common_cells,res]
merged[,tag] <- d[[item]]@meta.data[common_cells, tag]

for (item in names(d)) {
        ## cat('## ', tag, '\n\n')
        d[[item]]@meta.data$cell <- rownames(d[[item]]@meta.data)

        merged <- merge(merged,
                        d[[item]]@meta.data[c('cell', 'nCount_RNA', 'nFeature_RNA')],
                        by = 'cell',
                        all.x = TRUE,
                        suffixes = c('', sprintf('\n%s', item)))

        ## cat("### ", item, "\n\n")
}

rownames(merged) <- merged$cell
merged <- merged[,-1]
colnames(merged) <- gsub('_RNA', '', colnames(merged))
plot(merged, pch = 20, cex = 0.5, col = ac(pal[merged$scina, 'color'], 0.1))
## plot(merged, pch = 20, cex = 0.5, col = ac(as.numeric(as.factor(merged[,tag])), 1))
## cat('\n\n')

rm(merged)
```


# Pairwise scatterplot highlighting principal components and colored by cell identity {.tabset}


Comparing PCs to SCINA

```{r, results = 'asis', fig.width = 10, fig.height = 10}

common_cells <- Reduce(intersect, sapply(d, colnames))

for (item in names(d)) {
    cat('### ', item, '\n\n')


    tag <- 'scina'

    merged <- data.frame(cell = common_cells,
                         percent.mt = d[[tag_genes]]@meta.data[common_cells, 'percent.mt'])

    merged[,tag] <- d[[item]]@meta.data[common_cells, tag]

    d[[item]]@meta.data$cell <- rownames(d[[item]]@meta.data)

    merged <- cbind(merged,
                   Embeddings(d[[1]], reduction = "pca")[merged$cell, 1:5])
       
    rownames(merged) <- merged$cell
    merged <- merged[,-1]
    colnames(merged) <- gsub('_RNA', '', colnames(merged))
    
    plot(merged, pch = 20, cex = 0.5, col = ac(pal[merged$scina, 'color'], 0.1))

    rm(merged)
    cat("\n\n")
}

```


# Cell identity vs unsupervised clustering: F1 scores

Largely derived from Lukas Weber's https://github.com/lmweber/cytometry-clustering-comparison/blob/master/helpers/helper_match_evaluate_multiple.R (Hungarian algorithm)

@todo document parameters

```{r}
map_levels <- function(x){
    ## x<- as.character(iris$Species)
    if(is.factor(x)) {
        nlevs <- nlevels(x)
        levs <- levels(x)
    } else {   # character
        levs <- unique(x)
        nlevs <- length(levs)
    }
    fd <- data.frame(level = levs, code = 1:nlevs)
    return(fd)
}

## data(iris)
## map_levels(iris$Species)
## map_levels(as.character(iris$Species))
```

```{r}

## inputs are factor vectors
hungarian_weber_with_labels <- function(clus_algorithm, clus_truth) {
    ## number of detected clusters
    n_clus <- length(table(clus_algorithm))

    clus_names <- map_levels(clus_algorithm)
    clus <- as.numeric(clus_algorithm)

    truth_names <- map_levels(clus_truth)
    clus_truth <- as.numeric(clus_truth)

    ## remove unassigned cells (NA's in clus_truth)
    unassigned <- is.na(clus_truth)
    clus <- clus[!unassigned]
    clus_truth <- clus_truth[!unassigned]
    if (length(clus) != length(clus_truth)) warning("vector lengths are not equal")
    
    tbl_algorithm <- table(clus)
    tbl_truth <- table(clus_truth)

    ## detected clusters in rows, true populations in columns
    pr_mat <- re_mat <- F1_mat <- matrix(NA, nrow = length(tbl_algorithm), ncol = length(tbl_truth))
    
    for (i in 1:length(tbl_algorithm)) {
        for (j in 1:length(tbl_truth)) {
            i_int <- as.integer(names(tbl_algorithm))[i]  ## cluster number from algorithm
            j_int <- as.integer(names(tbl_truth))[j]  ## cluster number from true labels
            j_label <- as.character(truth_names[truth_names$code== j_int, 'level'])
                
            true_positives <- sum(clus == i_int & clus_truth == j_int, na.rm = TRUE)
            detected <- sum(clus == i_int, na.rm = TRUE)
            truth <- sum(clus_truth == j_int, na.rm = TRUE)
            
            ## calculate precision, recall, and F1 score
            precision_ij <- true_positives / detected
            recall_ij <- true_positives / truth
            F1_ij <- 2 * (precision_ij * recall_ij) / (precision_ij + recall_ij)
            
            if (F1_ij == "NaN") F1_ij <- 0
            
            pr_mat[i, j] <- precision_ij
            re_mat[i, j] <- recall_ij
            F1_mat[i, j] <- F1_ij
        }
    }
    
    ## put back cluster labels (note some row names may be missing due to removal of unassigned cells)
    rownames(pr_mat) <- rownames(re_mat) <- rownames(F1_mat) <- names(tbl_algorithm)
    colnames(pr_mat) <- colnames(re_mat) <- colnames(F1_mat) <- names(tbl_truth)
  
    ## match labels using Hungarian algorithm applied to matrix of F1 scores (Hungarian
    ## algorithm calculates an optimal one-to-one assignment)
    
    ## use transpose matrix (Hungarian algorithm assumes n_rows <= n_cols)
    F1_mat_trans <- t(F1_mat)
    
    if (nrow(F1_mat_trans) <= ncol(F1_mat_trans)) {
        ## if fewer (or equal no.) true populations than detected clusters, can match all true populations
        labels_matched <- clue::solve_LSAP(F1_mat_trans, maximum = TRUE)
        ## use row and column names since some labels may have been removed due to unassigned cells
        labels_matched <- as.numeric(colnames(F1_mat_trans)[as.numeric(labels_matched)])
        names(labels_matched) <- rownames(F1_mat_trans)
        
    } else {
        ## if fewer detected clusters than true populations, use transpose matrix and assign
        ## NAs for true populations without any matching clusters
        labels_matched_flipped <- clue::solve_LSAP(F1_mat, maximum = TRUE)
        ## use row and column names since some labels may have been removed due to unassigned cells
        labels_matched_flipped <- as.numeric(rownames(F1_mat_trans)[as.numeric(labels_matched_flipped)])
        names(labels_matched_flipped) <- rownames(F1_mat)
        
        labels_matched <- rep(NA, ncol(F1_mat))
        names(labels_matched) <- rownames(F1_mat_trans)
        labels_matched[as.character(labels_matched_flipped)] <- as.numeric(names(labels_matched_flipped))
    }
    
    ## precision, recall, F1 score, and number of cells for each matched cluster
    pr <- re <- F1 <- n_cells_matched <- rep(NA, ncol(F1_mat))
    names(pr) <- names(re) <- names(F1) <- names(n_cells_matched) <- names(labels_matched)
    
    for (i in 1:ncol(F1_mat)) {
        ## set to 0 if no matching cluster (too few detected clusters); use character names
        ## for row and column indices in case subsampling completely removes some clusters
        pr[i] <- ifelse(is.na(labels_matched[i]), 0, pr_mat[as.character(labels_matched[i]), names(labels_matched)[i]])
        re[i] <- ifelse(is.na(labels_matched[i]), 0, re_mat[as.character(labels_matched[i]), names(labels_matched)[i]])
        F1[i] <- ifelse(is.na(labels_matched[i]), 0, F1_mat[as.character(labels_matched[i]), names(labels_matched)[i]])
        
        n_cells_matched[i] <- sum(clus == labels_matched[i], na.rm = TRUE)
    }
    
    ## means across populations
    mean_pr <- mean(pr)
    mean_re <- mean(re)
    mean_F1 <- mean(F1)

    ## add back original names
    for (item in c('pr', 're', 'F1', 'labels_matched', 'n_cells_matched')) {
        curr <- get(item)
        names(curr) <- truth_names[names(curr), 'level']
        assign(item, curr)
    }
   
    return(list(n_clus = n_clus,
                pr = pr,
                re = re,
                F1 = F1,
                labels_matched = labels_matched,
                n_cells_matched = n_cells_matched,
                mean_pr = mean_pr,
                mean_re = mean_re,
                mean_F1 = mean_F1))
}

```

```{r}
hung <- list()

for (flavour in names(d)) {
    resolutions <- grep('RNA_snn_res', colnames(d[[flavour]]@meta.data), value = TRUE)
    hung[[flavour]] <- list()
    for (res in resolutions) {
        rownames(d[[flavour]]@meta.data) <- d[[flavour]]@meta.data$cell
        known <- rownames(d[[flavour]]@meta.data[d[[flavour]]@meta.data$scina != 'unknown',])

        length(as.factor(d[[flavour]]@meta.data[known, res]))
        length(as.factor(d[[flavour]]@meta.data[known, 'scina']))
        hung[[flavour]][[res]] <- hungarian_weber_with_labels(d[[flavour]]@meta.data[known, res],
                                                         d[[flavour]]@meta.data[known, 'scina'])

    }
    
}

```

```{r}

cobra <- list()
i <- 1
for (flavour in names(hung)) {
    for (res in names(hung[[flavour]])) {
    cobra[[i]] <- data.frame(cell = names(hung[[flavour]][[res]]$pr),
                             pr = hung[[flavour]][[res]]$pr,
                             re = hung[[flavour]][[res]]$re,
                             F1 = hung[[flavour]][[res]]$F1,
                             res = res,
                             flavour = flavour)
    i = i +1
    }
}

cobra <- do.call(rbind.data.frame, cobra)
cobra$id <- params$identifier
```

```{r, fig.height = 14, fig.width = 7}
ggplot(data = cobra, aes(x = pr, y =  re, group = res, col = res)) +
    geom_point() +
    facet_grid(flavour~cell, margins = FALSE) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r, fig.height = 12, fig.width = 8}
ggplot(data = cobra, aes(x = pr, y =  re, group = cell, col = cell)) +
    geom_point() +
    facet_grid(flavour~res, margins = FALSE) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    scale_colour_manual(values = setNames(pal$color, as.character(pal$identity)))

```

```{r, fig.height = 7, fig.width = 7}
ggplot(data = cobra, aes(x = res, y =  F1, group=flavour, col = cell)) +
    geom_point() +
    facet_wrap(~flavour) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    scale_colour_manual(values = setNames(pal$color, as.character(pal$identity)))

```

```{r, fig.height = 14, fig.width = 8}
ggplot(data = cobra, aes(x = re, y =  F1, group=res, col =cell)) +
    geom_point() +
    facet_grid(flavour~res) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    scale_colour_manual(values = setNames(pal$color, as.character(pal$identity)))

```

For each cell type, get the max F1 stratifying by flavour *regardless of the resolution*.

```{r}

## cobra <- cobra[order(cobra$F1, cobra$cell, decreasing = TRUE),]
## maxf1 <- cobra[0,]
maxF1 <- list()

for (flavour in levels(cobra$flavour)) {
    for (cell in levels(cobra$cell)) {
        maxF1[[paste(flavour, cell)]] <- c(flavour, cell, max(cobra[cobra$flavour == flavour &
                                                                    cobra$cell == cell, 'F1']))
    }
    
}


maxF1 <- do.call(rbind.data.frame, maxF1)
colnames(maxF1) <- c('flavour', 'cell', 'max_F1')
maxF1$max_F1 <- as.numeric(as.character(maxF1$max_F1))
      
maxF1$cell <- factor(as.character(maxF1$cell), levels = unique(as.character(maxF1$cell[order(maxF1$max_F1)])))

```

```{r, fig.width = 7, fig.height = 7}
ggplot(data = maxF1, aes(x= cell, y = max_F1, col = flavour, group = flavour)) +
    geom_point() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    scale_colour_manual(values = setNames(pal$color, as.character(pal$identity)))
       
```
