---
title: "PBMCs preliminary"
author: "Izaskun Mallona"
output:
  html_document:
    keep_md: false
    toc: true
    toc_float: true
    toc_depth: 4
params:
  seed: 11
---


@todo regress out mitochondrial contribution

```{r}
## stop('todo: get top quantiles for chipseq signals and run pagoda on these')

```

# Aim

Exploratory analysis on Hubrecht plates Tina 

Tip for transparency: `.Rprofile` with

```{r}
setHook(packageEvent("grDevices", "onLoad"),
function(...) grDevices::X11.options(type='cairo'))
options(device='x11')

```
# Preprocessing

```{r settings, include = TRUE}

library(knitr)
library(Seurat)
library(dplyr)
library(scater)
library(dplyr)
## library(data.table)
## library(SingleCellExperiment)
## library(scran)
## library(org.Mm.eg.db)
## suppressPackageStartupMessages(library(dplyr))
## suppressPackageStartupMessages(library(DT))
## library(pheatmap)
## library(Cairo)
## library(RColorBrewer)
## library(corrplot)

opts_chunk$set(fig.width = 8,
               fig.height = 8,
               cache = TRUE,
               include = TRUE,
               cache.lazy = FALSE,
               warning = TRUE,
               message = TRUE)


options(bitmapType='cairo')
getOption('bitmapType')
```

```{r}
COUNTS_PATH='/home/Shared_s3it/imallona/repeats_sc/data2/featurecounts'
```

```{r functions}
## add alpha
ac <- function(col, alpha=1){
  apply(sapply(col, col2rgb)/255, 2, 
                     function(x) 
                       rgb(x[1], x[2], x[3], alpha=alpha))  
}

```

Data load. We'll focus on TranscriptCounts, but read them all just in case we needed to compare summarizing ratios.

# Expression

```{r}


# this is wget http://cf.10xgenomics.com/samples/cell-exp/3.0.2/5k_pbmc_v3/5k_pbmc_v3_filtered_feature_bc_matrix.tar.gz
pbmc.data <- Read10X(data.dir = "../data/filtered_feature_bc_matrix/")
# Initialize the Seurat object with the raw (non-normalized data).
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc5k", min.cells = 3,
                           min.features = 200)

rm(pbmc.data)
pbmc
```

```{r}
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot2
```

```{r}
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500)
```

```{r}
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(pbmc), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
CombinePlots(plots = list(plot1, plot2))
```

```{r}
pbmc <- ScaleData(pbmc)
```

```{r}
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
```

```{r}
DimPlot(pbmc, reduction = "pca")
```

```{r}
ElbowPlot(pbmc)
```

```{r}
pbmc <- FindNeighbors(pbmc, dims = 1:16)
pbmc <- FindClusters(pbmc, resolution = 0.5)
```

```{r}
pbmc <- RunUMAP(pbmc, dims = 1:16)
                
DimPlot(pbmc, reduction = "umap")
```


```{r}
pbmc <- RunTSNE(pbmc, dims = 1:16)
                
DimPlot(pbmc, reduction = "tsne")
```

```{r}
# find markers for every cluster compared to all remaining cells, report only the positive ones
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
pbmc.markers %>% group_by(cluster) %>% top_n(n = 2, wt = avg_logFC)
```

```{r}
top10 <- pbmc.markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_logFC)
DoHeatmap(pbmc, features = top10$gene) + NoLegend()
```

```
0	IL7R, CCR7	Naive CD4+ T
1	IL7R, S100A4	Memory CD4+
2	CD14, LYZ	CD14+ Mono
3	MS4A1	B
4	CD8A	CD8+ T
5	FCGR3A, MS4A7	FCGR3A+ Mono
6	GNLY, NKG7	NK
7	FCER1A, CST3	DC
8	PPBP	Platelet
```

```{r}
DoHeatmap(pbmc, features = c('IL7R', 'CCR7', 'S100A4', 'CD14', 'LYZ',
                             'MS4A1', 'CD8A', 'FCGR3A', 'MS4A7',
                             'GNLY', 'NKG7',
                             'FCER1A', 'CST3',
                             'PPBP')) + NoLegend()
```

# Repeats

@todo fix this is untested, indicate the path

```{r}

if (!file.exists(file.path('data', 'repeats_counts_01.RData'))) {
## setwd(COUNTS_PATH)
    fns <- list.files(path = COUNTS_PATH, pattern = '*.counts$', recursive = TRUE)

    d <- list()
    for (fn in fns) {
        print(fn)
        ## d[[gsub('_counts', '', basename(fn))]] <- read.table(
        ##     file.path(COUNTS_PATH, fn),
        ##     header = TRUE,
        ##     sep = "\t",
        ##     colClasses = c(rep("NULL", 6), "integer"),
        ##     skip = 1)

        d[[gsub('_counts', '', basename(fn))]] <- read.table(
            pipe(sprintf('cut -f7 %s', file.path(COUNTS_PATH, fn))),            
            header = TRUE,
            sep = "\t",
            skip = 1)
        
    }



    str(d)

    ## last item, to get the annotation
    annot <-  read.table( fn,
                         header = TRUE,
                         sep = "\t",
                         skip = 1)[,1:6]


    d_bak <- d
    d <- cbind(annot, d)
    rownames(d) <- d$Geneid
    str(d)
} else {
    d <- load(file.path('data', 'repeats_counts_01.RData'))
}
```

```{r save_repeats_object}
if (!file.exists(file.path('data', 'repeats_counts_01.RData'))){
    dir.create('data')
    save(d, file = file.path('data', 'repeats_counts_01.RData'))
}
    
```


```{r}
reps <- CreateSeuratObject(counts = d,
                           project = "pbmc5k",
                           min.cells = 3,
                           min.features = 50)


```


```{r}
plot2 <- FeatureScatter(reps, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot2
```

```{r}
reps <- subset(reps, subset = nFeature_RNA > 200 & nFeature_RNA < 2500)
```

```{r}
reps <- FindVariableFeatures(reps, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(reps), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(reps)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
CombinePlots(plots = list(plot1, plot2))
```

```{r}
reps <- ScaleData(reps)
```

```{r}
reps <- RunPCA(reps, features = VariableFeatures(object = reps))
```

```{r}
DimPlot(reps, reduction = "pca")
```

```{r}
ElbowPlot(reps)
```

Highlighting the repeats expression of the RNAseq clusters?



# Timestamp

```{r sessionInfo, cache = FALSE}
date()
sessionInfo()
devtools::session_info()

```
