---
title: "Across samples report"
author: "Izaskun Mallona at Robinsonlab"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: kable
    theme: lumen
    always_allow_html: yes
params:
  run1: ""   
  run2: ""
  run3: ""
  run4: ""    
---


Rendered for samples:

```{r}
## print(params)
```
# Config

```{r}
suppressPackageStartupMessages({
library(Matrix)

library(knitr)
library(Seurat)
## library(dplyr)
## library(scater)
library(data.table)
## library(SingleCellExperiment)
## library(scran)
## library(org.Mm.eg.db)
## suppressPackageStartupMessages(library(dplyr))
## suppressPackageStartupMessages(library(DT))
library(pheatmap)
## library(Cairo)
## library(RColorBrewer)
## library(corrplot)
library(ggplot2)
library(reshape2)
## library(xtable)
## library(mclust)
library(viridis)
})
```

```{r}
BASE <- file.path('/home', 'imallona', 'repeats_sc')
```

```{r}
opts_chunk$set(fig.width = 5,
               fig.height = 5,
               dev = 'png',
               cache = TRUE,
               echo = TRUE,
               include = TRUE,
               cache.lazy = FALSE,
               warning = TRUE,
               error = TRUE,
               message = FALSE)


options(bitmapType='cairo')
## getOption('bitmapType')
```


```{r functions}
## add alpha
ac <- function(col, alpha=1){
  apply(sapply(col, col2rgb)/255, 2, 
                     function(x) 
                       rgb(x[1], x[2], x[3], alpha=alpha))  
}

load_standard_cellranger_genes <- function(run_name) {
    ## barcodes_fn <- params$genes_cellranger ## should end with `barcodes.tsv.gz`
    barcodes_fn <- file.path(BASE, 'runs', run_name, 'cellranger_standard', 'outs',
                             'filtered_feature_bc_matrix', 'barcodes.tsv.gz')
    
    path <- dirname(barcodes_fn)

    genes <- CreateSeuratObject(counts = Read10X(data.dir = path),
                                project = 'cellranger_standard',
                                min.cells = 0,
                                min.features = 0)
    return(genes)
}

load_cellranger_repeats <-  function(run_name) {
    barcodes_fn <- file.path(BASE, 'runs', run_name, 'cellranger_repeats', 'outs',
                             'filtered_feature_bc_matrix', 'barcodes.tsv.gz')

    path <- dirname(barcodes_fn)

    repeats <- CreateSeuratObject(counts = Read10X(data.dir = path),
                                project = 'cellranger_repeats',
                                min.cells = 0,
                                min.features = 0)
    return(repeats)
}

load_featurecounts_repeats <-  function(run_name) {

    repeats <- file.path(BASE, "runs", run_name, 'count_repeats_on_cellranger_standard',
                                paste0(run_name, "_repeats.counts.gz"))

    tmp <- data.table::fread(repeats, sep = '\t', header = TRUE,
                             drop = c('Chr', 'Start', 'End', 'Strand', 'Length'))
    rownames(tmp) <- tmp$Geneid ## here

    colnames(tmp) <- gsub(".*:([A-Z]+)-1.bam","\\1", colnames(tmp))

    tmp <- as.data.frame(tmp)
    rownames(tmp) <- tmp$Geneid
    tmp <- tmp[,-1]

    featurecounts <- CreateSeuratObject(
        tmp,
        project = "seurat",
        assay = "RNA",
        min.cells = 0,
        min.features = 0,
        names.field = 1,
        names.delim = "_",
        meta.data = NULL)

    return(featurecounts)
}

coef_var <- function(x){
        (sd(x)/mean(x))*100
}


```

# Data load

```{r justfordebugging, eval = FALSE}
## params <- list(genes_cellranger = "/home/imallona/repeats_sc/runs/frozen_pbmc_donor_a/cellranger_standard/outs/filtered_feature_bc_matrix/barcodes.tsv.gz",
##                repeats_cellranger= "/home/imallona/repeats_sc/runs/frozen_pbmc_donor_a/cellranger_repeats/outs/filtered_feature_bc_matrix/barcodes.tsv.gz",
##                repeats_featurecounts= "/home/imallona/repeats_sc/runs/frozen_pbmc_donor_a/count_repeats_on_cellranger_standard/frozen_pbmc_donor_a_repeats.counts.gz")

## SRR10974769

params <- list(run1 = '/home/imallona/repeats_sc/runs/SRR10974769/summarize_cellranger_run.rds',
               run2 = '/home/imallona/repeats_sc/runs/pbmc_10k_v3/summarize_cellranger_run.rds',
               run3 = '/home/imallona/repeats_sc/runs/frozen_pbmc_donor_a/summarize_cellranger_run.rds')
```


```{r}
d <- list()
```

```{r}
## cr cellranger
## fc featurecounts
for (rds in params) {
    
    ## d[[run]] <- list()
    ## d[[run]]$genes_cr <- load_standard_cellranger_genes(run_name = run)
    ## d[[run]]$reps_cr <- load_cellranger_repeats(run_name = run)
    ## d[[run]]$reps_fc <- load_featurecounts_repeats(run_name = run)
    d[[sapply(strsplit(dirname(rds), .Platform$file.sep),
               function(x) return(x[6]))]] <- readRDS(file = rds)
}
```

# PBMC markers

As in https://satijalab.org/seurat/v3.1/pbmc3k_tutorial.html

```{r}
ids <- read.table(text='marker,type
IL7R,naive_t_cd4
CCR7,naivet_t_cd4
IL7R,memory_t_cd4
S100A4,memory_t_cd4
CD14,mono
LYZ,mono
MS4A1,b
CD8A,t_cd8
FCGR3A,mono_fcgr3a
MS4A7,mono_fcgr3a
GNLY,nk
NKG7,nk
FCER1A,dc
CST3,dc
PPBP,platelet', sep = ',', header = TRUE, stringsAsFactors = FALSE)
```



```{r, fig.height = 3, fig.width = 6, dpi = 50, results="asis", message = FALSE} 
for (run in names(d)) {
    cat("## ", run, " {.tabset}\n\n")
    ## print(run)
    ## for (marker in ids$marker) {    
    ## for (res in grep('RNA_snn', colnames(d[[run]]$genes@meta.data), value = TRUE)) {
    res <- 'RNA_snn_res.0.2'
    ## print(FeaturePlot(object = d[[run]]$genes,
    ##                   features = ids$marker,
    ##                   cols = c("green", "blue"), 
    ##                   combine = TRUE,
    ##                   pt.size = 1,
    ##                   shape.by = res,
    ##                   reduction = 'umap',
    ##                   label = TRUE)) ## + ggtitle(paste(run, marker, res)))
        ## }
    ## }

    for (type in unique(ids$type)) {
        cat("### ", type, " {.tabset}\n\n")
        ## print(paste(run, type, res))
        print(RidgePlot(object = d[[run]]$genes,
                  features = ids[ids$type == type, 'marker'],
                  sort = FALSE,
                  assay = NULL,
                  group.by = res,
                  y.max = NULL,
                  same.y.lims = FALSE ,
                  log = FALSE,
                  ncol = NULL,
                  slot = "data",
                  combine = FALSE))

        cat("\n\n")
    }
    cat("\n\n")
}
```

# Most variable repeats cellranger {.tabset}

Similarly for repeats expression (avg +- sd of repeats CPM by gene expression cluster?)

What's the point here?

```{r}
## normalize CPMs
for (run in names(d)) {
    d[[run]]$repeats <- NormalizeData(d[[run]]$repeats,
                                      normalization.method = "RC", scale.factor = 1e6)
    d[[run]]$featurecounts <- NormalizeData(d[[run]]$featurecounts,
                                            normalization.method = "RC", scale.factor = 1e6)
}
```

```{r, fig.height = 5, fig.width = 15, results = 'asis'}
## selected <- c('AluY', 'AluSx', 'MER49', 'MIR', 'L2a', 'L2b', 'L1PA11')
selected <- c(names(head(sort(apply(GetAssayData(d[[run]]$repeats),1, coef_var), decreasing = TRUE),
                       5)), c('AluY', 'AluSx', 'L1PA11', 'MIR', 'L2a'))

for (run in names(d)) {
    cat("## ", run, "\n\n")
    repeats <- as.data.frame(FetchData(d[[run]]$repeats, vars = selected, slot = 'data'))
    repeats$cell <- rownames(repeats)

    d[[run]]$genes@meta.data$cell <- rownames(d[[run]]$genes@meta.data)
    repeats <- merge(d[[run]]$genes@meta.data[, c(res, 'cell') ],
                     repeats, by = 'cell', all.x = TRUE)
    
    rownames(repeats) <- repeats$cell
    
    repeats <- melt(repeats, id.vars = c('cell', res))
    
    colnames(repeats) <- c('cell', res, 'repfamily', 'CPM')

    print(ggplot(repeats,
           aes(x = get(res), y = CPM)) + 
          geom_boxplot(outlier.size=1) +
          scale_y_continuous(trans='log10') +
        scale_fill_discrete(drop=FALSE) +
        facet_wrap(~ factor(repfamily), ncol = 10) +
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        ylab('repeat expression [CPM]') +
        xlab('gene-based clustering') + ggtitle(sprintf('%s %s\n%s', run, 'repeats', res)))
    cat('\n\n')
}

```

# Most variable repeats featurecounts {.tabset}

Same for featurecounts-based repeats

```{r, fig.height = 5, fig.width = 15, results = 'asis'}
## selected <- c('AluY', 'AluSx', 'MER49', 'MIR', 'L2a', 'L2b', 'L1PA11')

for (run in names(d)) {
    cat("## ", run, "\n\n")
    featurecounts <- as.data.frame(FetchData(d[[run]]$featurecounts, vars = selected, slot = 'data'))
    featurecounts$cell <- rownames(featurecounts)

    d[[run]]$genes@meta.data$cell <- rownames(d[[run]]$genes@meta.data)
    featurecounts <- merge(d[[run]]$genes@meta.data[, c(res, 'cell') ],
                     featurecounts, by = 'cell', all.x = TRUE)
    
    rownames(featurecounts) <- featurecounts$cell
    
    featurecounts <- melt(featurecounts, id.vars = c('cell', res))
    
    colnames(featurecounts) <- c('cell', res, 'repfamily', 'CPM')

    print(ggplot(featurecounts,
           aes(x = get(res), y = CPM)) + 
          geom_boxplot(outlier.size=1) +
          scale_y_continuous(trans='log10') +
        scale_fill_discrete(drop=FALSE) +
        facet_wrap(~ factor(repfamily), ncol = 10) +
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        ylab('repeat expression [CPM]') +
        xlab('gene-based clustering') + ggtitle(sprintf('%s %s\n%s', run, 'repeats', res)))

    cat('\n\n')
}

```

# Pheatmaps

Note the cell sampling: same cells in all plots; `annotation_col` comes from genes' clustering.

```{r, results = 'asis', dpi=50, fig.width = 15, fig.height = 15}
NCELLS <- 1000
NFEATURES <- 500

## stop()
for (run in names(d)) {
    cat("## ", run, "{.tabset}\n\n")
    
    set.seed(654)
    idx_cell <- sample(colnames(GetAssayData(d[[run]]$genes)), NCELLS, replace = FALSE)
    
    for (item in names(d[[run]])) {
        cat("### ", item, "{.tabset}\n\n")

        set.seed(464)
        idx_feat <- sample(VariableFeatures(d[[run]][[item]]), NFEATURES, replace = FALSE)
        
        print(pheatmap(GetAssayData(
            object = subset(d[[run]][[item]],
                            features = idx_feat,
                            cells = intersect(idx_cell, colnames(GetAssayData(d[[run]][[item]])))),
                            slot = 'scale.data'),
            color = viridis(100),
            cluster_rows = TRUE,
            cluster_cols = TRUE,
            show_rownames = FALSE,
            show_colnames = FALSE,
            clustering_distance_cols = "euclidean",
            clustering_method = "ward.D2",
            clustering_distance_rows = "euclidean",
            fontsize_row = 12,
            annotation_col = as.data.frame(d[[run]]$genes@meta.data[idx_cell, c(res),
                                                                    drop = FALSE]),
            scale = "none"))
        cat('\n\n')
    }
    cat('\n\n')
}
```

Maybe add the markers of some celltypes on top?


# Timestamp

```{r sessionInfo, cache = FALSE}
date()
sessionInfo()
devtools::session_info()

```

