---
title: "Zheng FACS ground truth - cellranger only"
author: "Izaskun Mallona at Robinsonlab"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: kable
    theme: lumen
params:
  identifier: ""   
  genes_cellranger: ""
  repeats_cellranger: ""
  seurat_output: ""
  aris_output: ""
  cobra_output: ""
  markers_output: ""
  config_fn: ""
  regress_genes_nCount: ""
  regress_genes_nFeature: ""
---


# Config

## Parameters

Rendered for sample:

```{r}
print(params)
```

## Libraries

```{r}
suppressPackageStartupMessages({
    library(Matrix)

    library(knitr)
    library(Seurat)
    library(dplyr)
    library(scater)
    ## library(scater)
    library(data.table)
    ## library(SingleCellExperiment)
    ## library(scran)
    ## library(org.Mm.eg.db)
    ## suppressPackageStartupMessages(library(dplyr))
    ## suppressPackageStartupMessages(library(DT))
    ## library(pheatmap)
    ## library(Cairo)
    ## library(RColorBrewer)
    ## library(corrplot)
    library(ggplot2)
    ## library(GGally)
    library(reshape2)
    ## library(xtable)
    library(mclust)
    library(pheatmap)
    ## library(SCINA)
    library(DT)
    library(clue)
    ## library(iCOBRA)
    library(RColorBrewer)
    library(tximeta)
    library(SummarizedExperiment)
    library(biomaRt)
})
```

```{r}
opts_chunk$set(fig.width = 5,
               fig.height = 5,
               dev = 'png',
               cache = TRUE,
               include = TRUE,
               cache.lazy = FALSE,
               warning = TRUE,
               echo = TRUE,
               message = FALSE,
               knitr.table.format = "html")


options(bitmapType='cairo')
## getOption('bitmapType')
```


## Functions

```{r functions}
## add alpha
ac <- function(col, alpha=1){
  apply(sapply(col, col2rgb)/255, 2, 
                     function(x) 
                       rgb(x[1], x[2], x[3], alpha=alpha))  
}

## coef_var <- function(x){
##         (sd(x)/mean(x))*100
## }

```

# Data load

```{r justfordebugging, eval = FALSE}
## params <- list(genes_cellranger = "/home/imallona/repeats_sc/runs/frozen_pbmc_donor_a/cellranger_standard/outs/filtered_feature_bc_matrix/barcodes.tsv.gz",
##                repeats_cellranger= "/home/imallona/repeats_sc/runs/frozen_pbmc_donor_a/cellranger_repeats/outs/filtered_feature_bc_matrix/barcodes.tsv.gz",
##                repeats_featurecounts= "/home/imallona/repeats_sc/runs/frozen_pbmc_donor_a/count_repeats_on_cellranger_standard/frozen_pbmc_donor_a_repeats.counts.gz")

## SRR10974769

## params <- list(genes_cellranger = "/home/imallona/repeats_sc/runs/SRR10974769/cellranger_standard/outs/filtered_feature_bc_matrix/barcodes.tsv.gz",
##                repeats_cellranger= "/home/imallona/repeats_sc/runs/SRR10974769/cellranger_repeats/outs/filtered_feature_bc_matrix/barcodes.tsv.gz",
##                repeats_featurecounts= "/home/imallona/repeats_sc/runs/SRR10974769/count_repeats_on_cellranger_standard/SRR10974769_repeats.counts.gz",
##                identifier = 'test')

## params <- list(identifier = "5k_pbmc_v3",
##                genes_cellranger = '/home/imallona/repeats_sc/runs/5k_pbmc_v3/cellranger_standard/outs/filtered_feature_bc_matrix/barcodes.tsv.gz',
##                repeats_cellranger = "/home/imallona/repeats_sc/runs/5k_pbmc_v3/cellranger_repeats/outs/filtered_feature_bc_matrix/barcodes.tsv.gz",
##                repeats_featurecounts_multi = "/home/imallona/repeats_sc/runs/5k_pbmc_v3/count_repeats_on_cellranger_standard/multimappers/5k_pbmc_v3_repeats.counts.gz",
##                repeats_only_featurecounts_control_multi = "/home/imallona/repeats_sc/runs/5k_pbmc_v3/count_repeats_on_cellranger_standard_not_overlapping_genes/multimappers/5k_pbmc_v3_repeats_not_overlapping_genes.counts.gz",
##                repeats_featurecounts_unique = "/home/imallona/repeats_sc/runs/5k_pbmc_v3/count_repeats_on_cellranger_standard/unique_reads/5k_pbmc_v3_repeats.counts.gz",
##                repeats_only_featurecounts_control_unique = "/home/imallona/repeats_sc/runs/5k_pbmc_v3/count_repeats_on_cellranger_standard_not_overlapping_genes/unique_reads/5k_pbmc_v3_repeats_not_overlapping_genes.counts.gz",
##                seurat_output = "/home/imallona/tmp/test.rds",
##                regress_genes_nCount = FALSE,
##                regress_genes_nFeatures = FALSE,
##                repeats_bowtie_unique = '/home/imallona/repeats_sc/runs/5k_pbmc_v3/bowtie_repeatome/unique_reads/all_cells_repeats.counts.gz',
##                repeats_bowtie_multi = '/home/imallona/repeats_sc/runs/5k_pbmc_v3/bowtie_repeatome/multimappers/all_cells_repeats.counts.gz',
##                genes_alevin = '/home/imallona/repeats_sc/runs/5k_pbmc_v3/alevin/genes/alevin/quants_mat.gz',
##                repeats_alevin = '/home/imallona/repeats_sc/runs/5k_pbmc_v3/alevin/repeats/alevin/quants_mat.gz')




params <- list(identifier = "zheng_truth",
               genes_cellranger = '/home/imallona/repeats_sc/runs/zheng_truth',
               repeats_cellranger = "/home/imallona/repeats_sc/runs/zheng_truth",
               seurat_output = "/home/imallona/tmp/test.rds",
               regress_genes_nCount = FALSE,
               regress_genes_nFeatures = FALSE)#,
               ## repeats_bowtie_unique = '/home/imallona/repeats_sc/runs/zheng_truth/bowtie_repeatome/unique_reads/all_cells_repeats.counts.gz',
               ## repeats_bowtie_multi = '/home/imallona/repeats_sc/runs/zheng_truth/bowtie_repeatome/multimappers/all_cells_repeats.counts.gz',
               ## genes_alevin = 'None',
               ## repeats_alevin = 'None')


```

```{r}
d <- list(genes_cellranger = list(),
          repeats_cellranger = list())
```

```{r}
celltypes <- c('naive_t',
               'cd14_monocytes',
               'cd4_t_helper',
               'b_cells',
               'regulatory_t',
               'memory_t',
               'cd56_nk',
               'naive_cytotoxic',
               'cytotoxic_t',
               'cd34')

```

Genes cellranger data read

```{r}

for (celltype in celltypes){
    ## Genes data read

    barcodes_fn <- file.path(params$genes_cellranger, celltype, 'cellranger_standard',
                             'outs', 'filtered_feature_bc_matrix', 'barcodes.tsv.gz')
    path <- dirname(barcodes_fn)


    d$genes_cellranger[[celltype]] <- CreateSeuratObject(counts = Read10X(data.dir = path),
                                                         project = sprintf('%s\n%s', celltype,
                                                                           'genes_cellranger'),
                                                         min.cells = 0,
                                                         min.features = 0)
}

```

Collapse them into an object

```{r byintegration, message = FALSE, eval = FALSE}
for (i in 1:length(d$genes_cellranger)) {
    d$genes_cellranger[[i]] <- NormalizeData(d$genes_cellranger[[i]], verbose = FALSE)
    d$genes_cellranger[[i]] <- FindVariableFeatures(d$genes_cellranger[[i]],
                                                    selection.method = "vst", 
        nfeatures = 2000, verbose = FALSE)
}

anchors <- FindIntegrationAnchors(object.list = d$genes_cellranger, dims = 1:30)

d$genes_cellranger <- IntegrateData(anchorset = anchors, dims = 1:30)
```

```{r bymerging}

## ?Seurat::merge
d$genes_cellranger <- merge(x = d$genes_cellranger[[1]],
                            y = d$genes_cellranger[2:length(d$genes_cellranger)],
                            add.cell.ids = names(d$genes_cellranger),
                            project = params$identifier)


rownames(d$genes_cellranger@meta.data) <- colnames(d$genes_cellranger)
```

Similarly for repeats



```{r}

for (celltype in celltypes){
    ## Repeats data read

    barcodes_fn <- file.path(params$repeats_cellranger, celltype, 'cellranger_repeats',
                             'outs', 'filtered_feature_bc_matrix/barcodes.tsv.gz')
    path <- dirname(barcodes_fn)


    d$repeats_cellranger[[celltype]] <- CreateSeuratObject(counts = Read10X(data.dir = path),
                                                         project = sprintf('%s\n%s', celltype,
                                                                           'repeats_cellranger'),
                                                         min.cells = 0,
                                                         min.features = 0)
}

```

Collapse them into an object

```{r byintegration2, message = FALSE, eval = FALSE}
for (i in 1:length(d$repeats_cellranger)) {
    d$repeats_cellranger[[i]] <- NormalizeData(d$repeats_cellranger[[i]], verbose = FALSE)
    d$repeats_cellranger[[i]] <- FindVariableFeatures(d$repeats_cellranger[[i]],
                                                    selection.method = "vst", 
        nfeatures = 2000, verbose = FALSE)
}

anchors <- FindIntegrationAnchors(object.list = d$repeats_cellranger, dims = 1:30)

d$repeats_cellranger <- IntegrateData(anchorset = anchors, dims = 1:30)
```

```{r bymerging2}

d$repeats_cellranger <- merge(x = d$repeats_cellranger[[1]],
                            y = d$repeats_cellranger[2:length(d$repeats_cellranger)],
                            add.cell.ids = names(d$repeats_cellranger),
                            project = params$identifier)

rownames(d$repeats_cellranger@meta.data) <- colnames(d$repeats_cellranger)
## colnames(d$repeats_cellranger)
```

Mixed: genes + repeats

```{r}

## class(GetAssayData(d$repeats_cellranger, slot = 'counts'))

common_cells <- intersect(colnames(d$genes_cellranger), colnames(d$repeats_cellranger))
## length(common_cells)
## d$both_cellranger <- CreateSeuratObject(
##     rbind(as.matrix(GetAssayData(d$genes_cellranger, slot = 'counts'))[,common_cells],
##           as.matrix(GetAssayData(d$repeats_cellranger, slot = 'counts'))[,common_cells]),
##     assay = "RNA",
##     min.cells = 0,
##     min.features = 0,
##     names.field = 1,
##     names.delim = "_",
##     meta.data = NULL)

d$both_cellranger <- CreateSeuratObject(
    rbind(GetAssayData(d$genes_cellranger, slot = 'counts')[,common_cells],
          GetAssayData(d$repeats_cellranger, slot = 'counts')[,common_cells]),
    assay = "RNA",
    min.cells = 0,
    min.features = 0,
    names.field = 1,
    names.delim = "_",
    meta.data = NULL)

## rownames(d$both_cellranger@meta.data) <- colnames(d$both_cellranger)
## d$genes_cellranger@meta.data$id <- colnames(d$genes_cellranger)


d$both_cellranger@meta.data$orig.ident <- d$genes_cellranger@meta.data[common_cells, 'orig.ident']


## head(d$both_cellranger@meta.data)
## head(d$genes_cellranger@meta.data)

stopifnot(all(d$both_cellranger@meta.data[common_cells, 'orig.ident'] ==
              d$genes_cellranger@meta.data[common_cells, 'orig.ident'] ))

rm(common_cells)

```


# Basic Seurat processing

## QC before filtering {.tabset}

```{r}
d$genes_cellranger[["percent.mt"]] <- PercentageFeatureSet(d$genes_cellranger, pattern = "^MT-")

```


```{r}
## QC is based upon genes 
item <- 'genes_cellranger'

libsize.drop <- isOutlier(d$genes_cellranger$nCount_RNA, nmads = 3, type = "both", log = TRUE)
feature.drop <- isOutlier(d$genes_cellranger$nFeature_RNA, nmads = 3, type = "both", log = TRUE)
mito.drop <- isOutlier(d$genes_cellranger$percent.mt, nmads = 3, type = "higher")

selected_cells <- colnames(d$genes_cellranger)[!(libsize.drop | feature.drop | mito.drop)]

```


Outlier tagging results

```{r}
print(data.frame(ByLibSize = sum(libsize.drop),
                 ByFeature = sum(feature.drop), 
                 ByMito = sum(mito.drop),
                 Remaining = length(selected_cells)))
```

```{r, results = 'asis', fig.width = 8, fig.height = 4}
for (item in names(d)) {
    cat("### ", item, "\n\n")
    ## stop(grep("^MT-", rownames(d[[item]]), value = TRUE))
    if (item == 'genes_cellranger') {
        d[[item]][["percent.mt"]] <- PercentageFeatureSet(d[[item]], pattern = "^MT-")

        print(FeatureScatter(d[[item]], feature1 = "nCount_RNA", feature2 = "percent.mt") +
              geom_vline(xintercept = attributes(libsize.drop)$thresholds) +
              geom_hline(yintercept = attributes(mito.drop)$thresholds))
        
        print(VlnPlot(d[[item]], features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
                      ncol = 3))

    } else {
        print(VlnPlot(d[[item]], features = c("nFeature_RNA", "nCount_RNA"), ncol = 2))
    
    }

    d[[item]]@meta.data$qc_pass <- rownames(d[[item]]@meta.data) %in% selected_cells
    
    print(FeatureScatter(d[[item]], feature1 = "nCount_RNA", feature2 = "nFeature_RNA",
                         group.by = 'qc_pass', shape.by = 'qc_pass'))
    
    cat('\n\n')
}

```

## Basic stats pre-QC

```{r}
summarized <- list()
## summarized[['header']] <- c('dataset','filter', 'ncells',
##                             'ncount mean', 'ncount sd', 'nfeature mean', 'nfeature sd')
    
for (item in names(d)) {
    summarized[[item]] <- c(
        item, 'unfiltered', nrow(d[[item]]@meta.data),
        mean(d[[item]]@meta.data$nCount_RNA), sd(d[[item]]@meta.data$nCount_RNA),
        mean(d[[item]]@meta.data$nFeature_RNA), sd(d[[item]]@meta.data$nFeature_RNA))

    if ('percent.mt' %in% colnames(d[[item]]@meta.data)) {
        summarized[[item]] <- c(summarized[[item]],
                                mean(d[[item]]@meta.data$percent.mt, na.rm = TRUE))
    }
    else {
        summarized[[item]] <- c(summarized[[item]], NA)
        
    }
}

summarized <- as.data.frame(do.call(rbind.data.frame, summarized))
colnames(summarized) <- c('dataset','filter', 'ncells',
                          'ncount_mean', 'ncount_sd', 'nfeature_mean', 'nfeature_sd',
                          'mito_pct_mean')
```

```{r}
DT::datatable(summarized)
```

## QC after filtering {.tabset}

Filtering out cells according to the repeats expression, and forwarding the selection to other layers

```{r}
## the subsetting applies to all slots, even though the metrics come from genes only
for (item in names(d)) {
    d[[item]] <- subset(d[[item]], cells = selected_cells)
}

(d)
```

```{r, results = 'asis', fig.width = 8, fig.height = 4}
for (item in names(d)) {
    cat("### ", item, "\n\n")
    ## stop(grep("^MT-", rownames(d[[item]]), value = TRUE))
    if (item == 'genes_cellranger') {
        ## d[[item]][["percent.mt"]] <- PercentageFeatureSet(d[[item]], pattern = "^MT-")
        print(FeatureScatter(d[[item]], feature1 = "nCount_RNA", feature2 = "percent.mt"))
        print(VlnPlot(d[[item]], features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
                      ncol = 3))

    } else {
        print(VlnPlot(d[[item]], features = c("nFeature_RNA", "nCount_RNA"), ncol = 2) +
              ggtitle(item))
    
    }
    print(FeatureScatter(d[[item]], feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
          ggtitle(sprintf('%s', item)))
    cat('\n\n')
}

```

## Basic stats post-QC

```{r}
summarized <- list()
## summarized[['header']] <- c('dataset','filter', 'ncells',
##                             'ncount mean', 'ncount sd', 'nfeature mean', 'nfeature sd')
    
for (item in names(d)) {
    summarized[[item]] <- c(
        item, 'filtered', nrow(d[[item]]@meta.data),
        mean(d[[item]]@meta.data$nCount_RNA), sd(d[[item]]@meta.data$nCount_RNA),
        mean(d[[item]]@meta.data$nFeature_RNA), sd(d[[item]]@meta.data$nFeature_RNA))

    if ('percent.mt' %in% colnames(d[[item]]@meta.data)) {
        summarized[[item]] <- c(summarized[[item]],
                                mean(d[[item]]@meta.data$percent.mt, na.rm = TRUE))
    }
    else {
        summarized[[item]] <- c(summarized[[item]], NA)
        
    }
}

summarized <- as.data.frame(do.call(rbind.data.frame, summarized))
colnames(summarized) <- c('dataset','filter', 'ncells',
                          'ncount_mean', 'ncount_sd', 'nfeature_mean', 'nfeature_sd',
                          'mito_pct_mean')

```

```{r}
DT::datatable(summarized)
```

## Variable features {.tabset}

We take the top 10% of the features for genes or genes + repeats, but 25% for repeats.

```{r, message = FALSE, fig.width =8, fig.height = 4, results= 'asis'}
for (item in names(d)) {
    cat("### ", item, "\n\n")

    d[[item]] <- NormalizeData(object = d[[item]],
                               normalization.method = "LogNormalize",
                               scale.factor = 10000)

    ## if repeats, we take half of them
    ## else if genes, then the top 10 pct
    if (grepl('repeats', item)) {
        d[[item]] <- FindVariableFeatures(d[[item]], selection.method = "vst",
                                          nfeatures = round(nrow(d[[item]])/4))
    }
    else {
        d[[item]] <- FindVariableFeatures(d[[item]], selection.method = "vst",
                                          nfeatures = round(nrow(d[[item]])/10))
    }
    
    ## Identify the 10 most highly variable genes
    top10 <- head(VariableFeatures(d[[item]]), 10)

    ## plot variable features with and without labels
    plot1 <- VariableFeaturePlot(d[[item]])
    plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
    print(plot2 + ggtitle(item))
    cat('\n\n')
}



```

## Dimred PCA {.tabset}

```{r, message = FALSE, results ='asis'}
NUM_PCS <- 1:20
for (item in names(d)) {
    cat("### ", item, "\n\n")

    to_regress <- NULL
    if (as.logical(params$regress_genes_nCount)) {
        to_regress <- c(to_regress, 'nCount_RNA')
    }
    if (as.logical(params$regress_genes_nFeature)) {
        to_regress <- c(to_regress, 'nFeature_RNA')
        
    }
    
    if (item == 'genes_cellranger') {
        d[[item]] <- ScaleData(d[[item]], vars.to.regress = to_regress)
    } else {
        ## in this case, the variables to regress come from the genes!
        ## e.g. is the genes library size the one to normalize the repets'
        tmp <- d$genes_cellranger@meta.data
        tmp$cell <- rownames(tmp)
        d[[item]]@meta.data$cell <- rownames(d[[item]]@meta.data)
        d[[item]]@meta.data <- merge(d[[item]]@meta.data, tmp, by = 'cell',
                                     suffixes = c('', '_genes'))
        rownames(d[[item]]@meta.data) <- d[[item]]@meta.data$cell


        d[[item]] <- ScaleData(d[[item]],
                               vars.to.regress = if(is.null(to_regress)) NULL else paste0(to_regress, '_genes'))
        
        rm(tmp) 
    }
    
    d[[item]] <- RunPCA(d[[item]], features = VariableFeatures(object = d[[item]]))
    
    DimPlot(d[[item]], reduction = "pca")
    
    print(ElbowPlot(d[[item]]) + ggtitle(item))


    cat('\n\n')
}


```

```{r, message = FALSE, results = 'asis', eval = TRUE}
for (item in names(d)) {

    d[[item]] <- FindNeighbors(d[[item]], dims = NUM_PCS)
    for (res in seq(0.2, 2, by = 0.2)) {
        d[[item]] <- FindClusters(d[[item]], resolution = res, verbose = FALSE)

    }

    d[[item]] <- RunUMAP(d[[item]], dims = NUM_PCS)
        
    d[[item]] <- RunTSNE(d[[item]], dims = NUM_PCS, check_duplicates = FALSE )
}


```



## checkpoint

```{r}
saveRDS(d, file = paste0(params$seurat_output, 'debug1'))
```





## Dimred UMAP and tSNE {.tabset}

```{r fig.width = 8, fig.height = 5, results = 'asis', cache = FALSE, warning = FALSE}
for (item in names(d)) {
    cat("### ", item, "\n\n")
    
    p1 <- DimPlot(d[[item]], reduction = "umap", group.by = 'orig.ident')
    p2 <- DimPlot(d[[item]], reduction = "tsne", group.by = 'orig.ident')

    print(p1 + ggtitle(item))
    print(p2 + ggtitle(item))
    
    rm(p1, p2)
    cat('\n\n')

}
```


# Confusion matrix clusters

At a given resolution, compare both clusterings.

Color scale depicts Rand Index.


```{r}
aris <- list()
aris_stored <- list()
```

## Genes vs repeats (cellranger)

```{r, results = 'asis', fig.width = 4, fig.height = 4, cache = FALSE}
aris <- list()
i <- 1
for (genes_res in seq(0.2, 2, 0.2)) {
    genes_res_tag <- sprintf('RNA_snn_res.%s', genes_res)
    
    for (repeats_res in seq(0.2, 2, 0.2)) {
        repeats_res_tag <- sprintf('RNA_snn_res.%s', repeats_res)

        cells <- intersect(rownames(d$genes_cellranger@meta.data), rownames(d$repeats_cellranger@meta.data))
        
   

        aris[[i]] <- c(sprintf('genes_%s (n=%s)', genes_res,
                               length(unique(d$genes_cellranger@meta.data[cells, genes_res_tag]))),
                       sprintf('repeats_cr_%s (n=%s)', repeats_res,
                               length(unique(d$repeats_cellranger@meta.data[cells, repeats_res_tag]))),
                       adjustedRandIndex(d$genes_cellranger@meta.data[cells,genes_res_tag],
                                         d$repeats_cellranger@meta.data[cells,repeats_res_tag]))
        i <- i +1
    }    
}

aris <- do.call(rbind.data.frame, aris)
colnames(aris) <- c('genes_res', 'repeats_cr_res', 'ari')
aris$ari <- as.numeric(as.character(aris$ari))
```


```{r, fig.width = 6, fig.height = 6}
aris_wide <- reshape2::dcast(aris, genes_res~repeats_cr_res, value.var="ari")

rownames(aris_wide)<- aris_wide$genes_res
aris_wide <- aris_wide[,-1]

pheatmap(aris_wide, cluster_rows = FALSE, cluster_cols = FALSE,
         color = colorRampPalette(c("white","blue"))(20))
```

```{r store_aris}
## aris_stored <- list()
aris_stored[['genes__repeats_cr']] <- aris
```

# Differential repeat expression by 'true' identity {.tabset}

```{r}
## marker list
ml <- list()
```

```{r datatable_does_not_render, eval = TRUE}
for (item in names(d)) {
    cat('## ', item, '\n\n')
    
    Idents(d[[item]]) <- 'orig.ident'

    ## find markers for every scina identity (discarding the unplaced); report only positives    
    ml[[item]] <- FindAllMarkers(d[[item]],
                              only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

    ## DT::datatable(ml[[item]])
    print(head(ml[[item]]))
    
    cat('\n\n')
}

```

# F1 scores

Clustering vs sorted truth


Largely derived from Lukas Weber's https://github.com/lmweber/cytometry-clustering-comparison/blob/master/helpers/helper_match_evaluate_multiple.R (Hungarian algorithm)

@todo document parameters

```{r}
map_levels <- function(x){
    ## x<- as.character(iris$Species)
    if(is.factor(x)) { 
        nlevs <- nlevels(x)
        levs <- levels(x)
    } else {   # character
        levs <- unique(x)
        nlevs <- length(levs)
    }
    fd <- data.frame(level = levs, code = 1:nlevs)
    return(fd)
}

## data(iris)
## map_levels(iris$Species)
## map_levels(as.character(iris$Species))
```

```{r}

## inputs are factor vectors
hungarian_weber_with_labels <- function(clus_algorithm, clus_truth) {    
    ## number of detected clusters
    n_clus <- length(table(clus_algorithm))

    clus_names <- map_levels(clus_algorithm)
    clus <- as.numeric(clus_algorithm)

    truth_names <- map_levels(clus_truth)
    clus_truth <- as.numeric(clus_truth)

    ## remove unassigned cells (NA's in clus_truth)
    unassigned <- is.na(clus_truth)
    clus <- clus[!unassigned]
    clus_truth <- clus_truth[!unassigned]
    if (length(clus) != length(clus_truth)) warning("vector lengths are not equal")
    
    tbl_algorithm <- table(clus)
    tbl_truth <- table(clus_truth)

    ## detected clusters in rows, true populations in columns
    pr_mat <- re_mat <- F1_mat <- matrix(NA, nrow = length(tbl_algorithm), ncol = length(tbl_truth))
    
    for (i in 1:length(tbl_algorithm)) {
        for (j in 1:length(tbl_truth)) {
            i_int <- as.integer(names(tbl_algorithm))[i]  ## cluster number from algorithm
            j_int <- as.integer(names(tbl_truth))[j]  ## cluster number from true labels
            j_label <- as.character(truth_names[truth_names$code== j_int, 'level'])
                
            true_positives <- sum(clus == i_int & clus_truth == j_int, na.rm = TRUE)
            detected <- sum(clus == i_int, na.rm = TRUE)
            truth <- sum(clus_truth == j_int, na.rm = TRUE)
            
            ## calculate precision, recall, and F1 score
            precision_ij <- true_positives / detected
            recall_ij <- true_positives / truth
            F1_ij <- 2 * (precision_ij * recall_ij) / (precision_ij + recall_ij)
            
            if (F1_ij == "NaN") F1_ij <- 0
            
            pr_mat[i, j] <- precision_ij
            re_mat[i, j] <- recall_ij
            F1_mat[i, j] <- F1_ij
        }
    }
    
    ## put back cluster labels (note some row names may be missing due to removal of unassigned cells)
    rownames(pr_mat) <- rownames(re_mat) <- rownames(F1_mat) <- names(tbl_algorithm)
    colnames(pr_mat) <- colnames(re_mat) <- colnames(F1_mat) <- names(tbl_truth)
  
    ## match labels using Hungarian algorithm applied to matrix of F1 scores (Hungarian
    ## algorithm calculates an optimal one-to-one assignment)
    
    ## use transpose matrix (Hungarian algorithm assumes n_rows <= n_cols)
    F1_mat_trans <- t(F1_mat)
    
    if (nrow(F1_mat_trans) <= ncol(F1_mat_trans)) {
        ## if fewer (or equal no.) true populations than detected clusters, can match all true populations
        labels_matched <- clue::solve_LSAP(F1_mat_trans, maximum = TRUE)
        ## use row and column names since some labels may have been removed due to unassigned cells
        labels_matched <- as.numeric(colnames(F1_mat_trans)[as.numeric(labels_matched)])
        names(labels_matched) <- rownames(F1_mat_trans)
        
    } else {
        ## if fewer detected clusters than true populations, use transpose matrix and assign
        ## NAs for true populations without any matching clusters
        labels_matched_flipped <- clue::solve_LSAP(F1_mat, maximum = TRUE)
        ## use row and column names since some labels may have been removed due to unassigned cells
        labels_matched_flipped <- as.numeric(rownames(F1_mat_trans)[as.numeric(labels_matched_flipped)])
        names(labels_matched_flipped) <- rownames(F1_mat)
        
        labels_matched <- rep(NA, ncol(F1_mat))
        names(labels_matched) <- rownames(F1_mat_trans)
        labels_matched[as.character(labels_matched_flipped)] <- as.numeric(names(labels_matched_flipped))
    }
    
    ## precision, recall, F1 score, and number of cells for each matched cluster
    pr <- re <- F1 <- n_cells_matched <- rep(NA, ncol(F1_mat))
    names(pr) <- names(re) <- names(F1) <- names(n_cells_matched) <- names(labels_matched)
    
    for (i in 1:ncol(F1_mat)) {
        ## set to 0 if no matching cluster (too few detected clusters); use character names 
        ## for row and column indices in case subsampling completely removes some clusters
        pr[i] <- ifelse(is.na(labels_matched[i]), 0, pr_mat[as.character(labels_matched[i]), names(labels_matched)[i]])
        re[i] <- ifelse(is.na(labels_matched[i]), 0, re_mat[as.character(labels_matched[i]), names(labels_matched)[i]])
        F1[i] <- ifelse(is.na(labels_matched[i]), 0, F1_mat[as.character(labels_matched[i]), names(labels_matched)[i]])
        
        n_cells_matched[i] <- sum(clus == labels_matched[i], na.rm = TRUE)
    }
    
    ## means across populations
    mean_pr <- mean(pr)
    mean_re <- mean(re)
    mean_F1 <- mean(F1)

    ## add back original names
    for (item in c('pr', 're', 'F1', 'labels_matched', 'n_cells_matched')) {
        curr <- get(item)
        names(curr) <- truth_names[names(curr), 'level']
        assign(item, curr) 
    }
   
    return(list(n_clus = n_clus, 
                pr = pr, 
                re = re, 
                F1 = F1, 
                labels_matched = labels_matched, 
                n_cells_matched = n_cells_matched, 
                mean_pr = mean_pr, 
                mean_re = mean_re, 
                mean_F1 = mean_F1))
}

```

```{r, eval = TRUE}
a <- 1
saveRDS(object = d, file = paste0(params$seurat_output, '.debug2')) # checkpoint
## saveRDS(object = a, file = params$aris_output)
## saveRDS(object = a, file = params$cobra_output)
## saveRDS(object = a, file = params$markers_output)

```

```{r, eval = FALSE}
knitr::knit_exit()
```

```{r}
hung <- list()

## flavour <- 'genes_cellranger'

for (flavour in names(d)) {
    d[[flavour]]@meta.data$truth <- sapply(
        strsplit(as.character(d[[flavour]]@meta.data$orig.ident), '\n'),
        function(x) return(x[1]))

    ## head(d[[1]]@meta.data$truth)
    
    ## rownames(d[[flavour]]@meta.data) <- 
    resolutions <- grep('RNA_snn_res', colnames(d[[flavour]]@meta.data), value = TRUE)
    hung[[flavour]] <- list()
    for (res in resolutions) {
        rownames(d[[flavour]]@meta.data) <- d[[flavour]]@meta.data$cell
        known <- rownames(d[[flavour]]@meta.data)

        length(as.factor(d[[flavour]]@meta.data[known, res]))
        length(as.factor(d[[flavour]]@meta.data[known, 'truth']))
        hung[[flavour]][[res]] <- hungarian_weber_with_labels(
            d[[flavour]]@meta.data[known, res],
            as.factor(d[[flavour]]@meta.data[known, 'truth']))

    }    
}

```

```{r}

cobra <- list()
i <- 1
for (flavour in names(hung)) {
    for (res in names(hung[[flavour]])) {
    cobra[[i]] <- data.frame(cell = names(hung[[flavour]][[res]]$pr),
                             pr = hung[[flavour]][[res]]$pr,
                             re = hung[[flavour]][[res]]$re,
                             F1 = hung[[flavour]][[res]]$F1,
                             res = res,
                             flavour = flavour)
    i = i +1
    }
}

cobra <- do.call(rbind.data.frame, cobra)
cobra$id <- params$identifier

```


For each cell type, get the max F1 stratifying by flavour *regardless of the resolution*.

```{r}
tryCatch({
    ## cobra <- cobra[order(cobra$F1, cobra$cell, decreasing = TRUE),]
    ## maxf1 <- cobra[0,]
    maxF1 <- list()

    for (flavour in levels(cobra$flavour)) {
        for (cell in levels(cobra$cell)) {
            maxF1[[paste(flavour, cell)]] <- c(flavour, cell, max(cobra[cobra$flavour == flavour &
                                                                        cobra$cell == cell, 'F1']))
        }
        
    }


    maxF1 <- do.call(rbind.data.frame, maxF1)
    colnames(maxF1) <- c('flavour', 'cell', 'max_F1')
    maxF1$max_F1 <- as.numeric(as.character(maxF1$max_F1))
    
    maxF1$cell <- factor(as.character(maxF1$cell),
                         levels = unique(as.character(maxF1$cell[order(maxF1$max_F1)])))
}, error = function(x) print(x))
```

```{r}
DT::datatable(as.data.frame(cobra %>%
              group_by(flavour, cell) %>%
              summarize(max(F1)) %>%
              arrange(desc(`max(F1)`))))
````


```{r, fig.height = 8, fig.width = 10}
ggplot(data = cobra, aes(x = pr, y =  re, group = cell, col = cell)) +
    geom_point() +
    facet_grid(flavour~res, margins = FALSE) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) # +
    ## scale_colour_manual(values = setNames(pal$color, as.character(pal$identity)))

```

```{r, fig.height = 8, fig.width = 8}
ggplot(data = cobra, aes(x = res, y =  F1, group=flavour, col = cell)) +
    geom_point() +
    facet_wrap(~flavour) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ## scale_colour_manual(values = setNames(pal$color, as.character(pal$identity))) +
    labs(x = 'resolution')

```

```{r, fig.height = 7, fig.width = 10}

cobra$resolution <- gsub('RNA_snn_res_', '', cobra$res)
ggplot(data = cobra, aes(x = re, y =  F1, group=resolution, col =cell)) +
    geom_point() +
    facet_grid(flavour~res) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) # +
    ## scale_colour_manual(values = setNames(pal$color, as.character(pal$identity)))

```

```{r, fig.width = 5, fig.height = 5}

tryCatch({
    ## ## sorting in a more sensible way
    ## @todo adapt to this run
    ## maxF1$flavour <- factor(maxF1$flavour,
    ##                         levels = c('genes_alevin', 'genes_featurecounts',
    ##                                    'both_unique', 'both_multi', 'both_alevin',
    ##                                    'repeats_alevin',
    ##                                    'repeats_bowtie_unique', 'repeats_bowtie_multi',
    ##                                    'repeats_featurecounts_unique', 'repeats_featurecounts_multi',
    ##                                    'control_repeats_only_unique', 'control_repeats_only_multi'))
    

    ggplot(data = maxF1, aes(x= flavour, y = max_F1, col = cell, group = cell)) +
        geom_point() +
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        scale_colour_manual(values = setNames(pal$color, as.character(pal$identity)))
}, error = function(x) print(x)) 
```



# Export objects

List of seurat objects

```{r}
saveRDS(object = d, file = params$seurat_output)
saveRDS(object = aris_stored, file = params$aris_output)
saveRDS(object = ml, file = params$markers_output)
saveRDS(object = cobra, file = params$cobra_output)

```

# Timestamp

```{r sessionInfo, cache = FALSE}
date()
sessionInfo()
devtools::session_info()

```


